\documentclass[11pt]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm, bm}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}

% --- Page Setup ---
\geometry{letterpaper, margin=1in}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

% --- Document Metadata ---
\title{\textbf{The Seven Pillars of the Abstract Universe}}
\author{Juha Meskanen}
\date{December 2025}

\begin{document}

\maketitle

\section{Pillar 1 --- The Pre-Reality: Infinite Potential}
A theory of everything cannot assume anything. If any law, structure, logic, information, or distinction is predefined, the theory is incomplete by construction. Therefore, we assume reality begins not with laws, but with unlimited potential.

\textbf{Axiom 1 (Pre-Reality).} There exists an unconstrained domain $\mathbf{P}$, called \textit{Pre-Reality}, defined as the totality of all possible configurations:
\[
\mathbf{P} = \{\text{all possibilities}\}
\]
No structure, measure, ordering, logic, time, or information is defined on $\mathbf{P}$. Information does not exist at this level because information requires realized distinctions (entropy), and no distinctions are assumed.

Pillar 1 is not an empirical hypothesis but a boundary condition. Its role is to deny the existence of prior constraints on reality.

\section{Pillar 2 --- The Observer Exists}
A Theory of Everything without the observer is, by definition, incomplete. An \textbf{observer} is a finite, informational structure. 

\textbf{Corollary 2.1 (Existence of Structured Worlds).} Since no configuration is excluded from $\mathbf{P}$, there exists at least one finite subset:
\[
\exists\, S \subset \mathbf{P}
\]
within which internal relations define distinctions, information, mathematics, physical laws, and conscious observers. Existence requires no external explanation---nothing forbids it.

\section{Pillar 3: Configuration Space --- The Static Multiverse}
We formalize the existence of a finite informational structure with the simplest representation: the \textbf{binary set} and its ordered \textbf{bitstring} counterpart.
\begin{itemize}
\item A \textbf{configuration space} $\mathcal{C} \cong \{0, 1\}^n$ is the set of all possible bitstrings of length $n$.
  Each specific bitstring $\mathbf{s} \in \mathcal{C}$ represents a potential \textbf{state of the universe}---an Everettian world.
\item The total space $\mathcal{C}$ constitutes the \textbf{Superspace} of the Wheeler--DeWitt equation, where the universal wavefunction $\Psi$ is defined.
\item All these worlds coexist as a static \textbf{library of possibilities}; there is no inherent dynamics until a permutation (path) is selected.
\end{itemize}

\section{Pillar 4 --- Permutation Space: The Emergence of Time}
\begin{itemize}
    \item A \textbf{permutation} of the configurations defines a \textbf{path} or ``history'' through the static multiverse.
    \item \textbf{Time} emerges as the observer’s sequential experience along such a path. It is \textbf{not fundamental}, but a relational, emergent property of moving through ordered informational states.
    \item \textbf{The Arrow of Time} is a consequence of this ordering; it is the direction of increasing realized information (entropy) as the observer traverses the permutation.
\end{itemize}

\section{Pillar 5 --- Selection Filter: MDL and ASP}
Among all possible paths through a finite informational subset $S \subset \mathbf{P}$, the observer's finite memory and requirement for coherence naturally filter the paths that can be experienced.

\textbf{Minimal Description Length (MDL):} Smooth, predictable sequences are algorithmically compressible. Let $H = (s_1, s_2, \dots, s_n)$ denote a path. Sequences with low Kolmogorov complexity $K(H)$ are \textbf{more likely}, while incompressible sequences are suppressed.

\textbf{Algorithmic Selection Principle (ASP):} Paths of minimal description length are dominated:
\[
H_\text{MDL} = \arg\min_H K(H)
\]
\textbf{Emergent Finiteness:} Infinite-memory paths are algorithmically incompressible and therefore have zero probability. Laws of physics are not fundamental; they are the optimal compression of information.

\section{Pillar 6 --- The Wavefunction as an Optimal Encoding}
The wavefunction ($\Psi$) is the optimal compression algorithm. The formal properties of quantum mechanics (superposition, linear evolution, interference) arise as artifacts of any maximally efficient encoding.

\subsection*{The Born Rule (Decoding Measure)}
The Born Rule emerges as a natural consequence of consistently decoding the compressed representation:
\[
P(\gamma) \propto |\psi(\gamma)|^2 \sim 2^{-K(\gamma)}
\]
where $K(\gamma)$ is the Kolmogorov complexity of the path $\gamma$. The squared amplitude measures the effective probability of each continuation.

\section{Pillar 7 --- Emergent Geometry: Gravity as Geometric Interpretation}
Gravity is the geometric interpretation of probability gradients in configuration space.

\textbf{Geodesics as MDL Paths:} Geodesics are the most compressible and hence the most probable permutations, those satisfying the Minimal Description Length (MDL) criterion. Physical curvature is the manifestation of the selection pressure exerted by the Algorithmic Selection Principle.


\section{Conclusion: The Emergence of Necessity from Absolute Contingency}

The framework presented here suggests a radical reinterpretation of the nature of physical law.
Rather than assuming a specific set of initial conditions or a fixed Hamiltonian, we begin with the \textbf{Absolute No-Boundary Condition}:
the assumption that in Pre-Reality, there are no prior constraints, no predefined logic, and no inherent laws.

Under this condition of infinite potential, the emergence of a structured universe is not a violation of chaos, but an informational necessity.
The transition from \textbf{Pre-Reality} to \textbf{Physical Reality} follows a rigorous logical chain:

\begin{enumerate}
    \item \textbf{Absence of Constraint:} Because nothing is forbidden, finite structured subsets (observers) exist by definition.
    \item \textbf{Informational Economy:} Universe we observe is finite; permutations that are algorithmically compressible dominate.
    \item \textbf{Probability from Compression:} The Algorithmic Selection Principle (ASP) ensures that paths of Minimal Description Length (MDL) dominate the probability distribution of realized worlds.
    \item \textbf{Predictability as Law:} What we perceive as the ``Laws of Physics'' (gravity, quantum mechanics, conservation laws) are simply the mathematical manifestations of this compression. Predictable information compresses well.
\end{enumerate}


In this view, the universe is not governed by laws; rather, the \textbf{laws are the most probable outcome} of a system that starts with no rules.
Predictability is the inevitable byproduct of survival. Stability, inertia, and even the curvature of spacetime are the geometric interpretations of the fact that, among all possible ways for information to be ordered, the most compressible orderings are the ones that constitute a stable, observable reality.

The ``No-Boundary Condition'' is thus the ultimate boundary: by assuming nothing, we derive the necessity of everything.


\end{document}
