\documentclass[11pt]{article}
\input{../common.tex}
\input{../glossary.tex}
\addbibresource{../references.bib}

\title{The Geometry of Spacetime as an Information-Theoretic Structure}

\author{Juha Meskanen}
\date{2010 \dots 2026}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We derive the geometric structure of spacetime from the information-theoretic
framework established in Paper~I \dots III \cite{meskanen2001, meskanen2002, meskanen2003},
in which the universe is static, timeless,
and finite, with total information content $n$ bits. Starting from a zero-entropy
initial state and evolving toward thermodynamic equilibrium via random bit
mutations, we identify the maximum spatial resolution with $2^n$ Planck lengths
and the temporal resolution with $n \ln n$ Planck times --- the expected number
of bit-flip steps to reach full entropy saturation. Their ratio defines a
dimensionless \emph{aspect ratio} $\mathcal{A}(n) = 2^n / (n \ln n)$ that
encodes the complete geometric structure of spacetime from $n$ alone.
Calibrating $\mathcal{A}$ against the inflationary expansion of the early
universe, expressed in Planck units, yields $n \approx 184$ bits --- from which
the number of e-folds per dimension falls to approximately 60, consistent with
the independently derived lower bound from standard inflationary cosmology.
Extending the framework to black holes by identifying
$n_{\mathrm{bh}} = \log_2(r_s / \ell_P)$, where $r_s$ is the Schwarzschild radius, we recover the
Bekenstein--Hawking entropy of a solar-mass black hole.
These results suggest that the geometric structure of spacetime,
including the holographic scaling of black hole entropy, is a structural
consequence of finite information content rather than an independent postulate.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

General Relativity describes the geometry of spacetime with extraordinary
precision, yet provides no account of why spacetime has a geometry at all, or
why matter curves it. The equations are accepted without a deep explanation of
what space is made of such that it \emph{can} curve. Similarly, the
Bekenstein--Hawking entropy formula \cite{bekenstein1973, hawking1975}
establishes that black hole entropy scales with horizon area, but the
microscopic origin of that entropy --- the nature of the degrees of freedom
being counted --- remains an open question.

The present paper addresses both questions within the framework developed in
papers I \dots III \cite{meskanen2001, meskanen2002, meskanen2003}.
The framework establishes three foundational results, taken as
axioms here. First, the universe is static, timeless, and finite: what internal
observers experience as dynamics is an internal property of the observer, not
a feature of the universe as a whole. Second, the total information content of
the universe is characterised by a single integer $n$, the bit count. Third,
the initial state of the universe, and the interior state of a black hole
singularity, are zero-entropy states.

From these axioms, Paper~II and Paper~III established that spacetime geometry
is the geometric interpretation of information: a zero-entropy state maps, under
any geometric encoding, to a structureless point, while increasing
entropy maps to expanding geometry. The present paper asks a more precise
question: given $n$, what is the \emph{resolution} of that geometry in both
space and time, and does the resulting structure reproduce known results from
cosmology and black hole thermodynamics?

The paper is structured as follows. Section~\ref{sec:framework} summarises
the relevant results from Papers~II and III. Section~\ref{sec:resolutions}
derives the spatial and temporal resolutions from $n$ and defines the aspect
ratio. Section~\ref{sec:calibration} calibrates the aspect ratio against
inflationary cosmology and solves for $n$. Section~\ref{sec:efolding} notes
the correspondence with the e-fold count. Section~\ref{sec:blackhole} extends
the framework to black holes and compares with Bekenstein--Hawking.
Section~\ref{sec:discussion} discusses the results and their implications.
Section~\ref{sec:conclusion} concludes.

% ============================================================
\section{Framework}
\label{sec:framework}

We summarise the results from Papers~I and II that are used in what follows.

\textbf{Axiom 1 (Static universe).} The universe is static, timeless, and
finite. Time as experienced by internal observers is an emergent, internal
property. The universe operates on a strict, finite bit budget.

\textbf{Axiom 2 (Bit count).} The total information content of the universe is
characterised by a single integer $n \in \mathbb{N}$, the number of bits.
There are $2^n$ possible configurations.

\textbf{Axiom 3 (Zero-entropy boundary conditions).} The initial state of the
universe is a zero-entropy state. The interior state of a black hole singularity
is a zero-entropy state.

\textbf{Corollary (Geometric point).} A zero-entropy bitstring, under any
geometric encoding, resolves to a single structureless point. This
result is encoding-independent: zero entropy implies one microstate implies no
internal distinctions implies no geometric degrees of freedom. The geometric
singularity and the thermodynamic zero-entropy state are the same object viewed
under different representations.

\textbf{Corollary (Expanding geometry).} Starting from the zero-entropy state
and evolving toward thermodynamic equilibrium via random bit mutations, the
entropy growth curve
\[
S(t) = I_{\max} \cdot \bigl(1 - e^{-k t}\bigr)
\]
generates, under any geometric encoding, a geometry that begins as a point and
expands monotonically. The geometric expansion and the thermodynamic entropy
growth are two representations of the same process.

% ============================================================
\section{Spatial and Temporal Resolutions}
\label{sec:resolutions}

Given a universe of $n$ bits, we ask: what are the natural resolutions of
space and time in such a system?

\subsection{Spatial Resolution}

With $n$ bits, there are $2^n$ distinct configurations. If space is the
geometric interpretation of information, then $2^n$ is the number of
distinguishable spatial states --- the maximum spatial resolution of the
system. We identify this with a count of Planck-length cells:

\[
\Delta x_{\max} = 2^n \cdot \ell_P.
\]

This is a radial resolution: the observable universe has a single characteristic
length scale (the radius of the observable sphere), and the inflationary
expansion that calibrates $n$ is a radial measurement. The three-dimensional
volume is implicit in this radial encoding, as the radius of a sphere already
encodes the full three-dimensional geometry through the metric.

\subsection{Temporal Resolution}

The natural temporal resolution is the number of bit-flip steps required to
bring a zero-entropy system to full entropy saturation. By the coupon-collector
result from combinatorics \cite{coupon}, a random process that flips bits one at
a time requires, on average,

\[
n \ln n
\]

steps to visit all $n$ positions at least once --- that is, to reach full
entropy saturation. Identifying one bit-flip with one Planck time $t_P$:

\[
\Delta t_{\max} = n \ln n \cdot t_P.
\]

\subsection{The Aspect Ratio}

Dividing the spatial resolution by the temporal resolution yields a
dimensionless quantity:

\[
\mathcal{A}(n) = \frac{2^n}{n \ln n}.
\]

This \emph{aspect ratio} encodes the complete geometric structure of the
spacetime block. Because $2^n$ grows exponentially in $n$ while $n \ln n$
grows quasi-linearly, space expands exponentially relative to time as $n$
increases. Given $n$, both dimensions --- and hence the full geometry --- are
determined.

\textbf{Lemma (Monotonicity).} $\mathcal{A}(n)$ is strictly increasing in $n$
for $n \geq 3$.

\textbf{Proof.} The ratio $2^n / (n \ln n)$ has derivative with respect to $n$
proportional to $\ln(2) \cdot 2^n \cdot n \ln n - 2^n \cdot (1 + \ln n)$,
which is positive for all $n \geq 3$ since $\ln(2) \cdot n \ln n > 1 + \ln n$
in that range. \qed

\textbf{Corollary.} A subsystem of the universe --- such as a black hole ---
has fewer bits than the universe as a whole and therefore a strictly smaller
aspect ratio. The hierarchy $\mathcal{A}(n_{\mathrm{bh}}) < \mathcal{A}(n)$ is
a structural consequence of the framework, not an additional assumption.

\subsection*{From 1D Plane to 3D Sphere}

Based on observational evidence (and GR sperical symmetry)
we are not living in flat rectantular shape but on the surface
of 3D sphere. 

\[
\boxed{\text{1D information radius} \xrightarrow{\times 4\pi R^2} \text{3D physical observable}}
\]


% ============================================================
\section{Calibration Against Inflationary Cosmology}
\label{sec:calibration}

To determine $n$, we require a dimensionless observable that can be identified
with $\mathcal{A}(n)$. The inflationary period of the early universe provides
this observable: during inflation, the universe expanded rapidly from a near-zero
entropy state, making it the closest physical realisation of the thermodynamic
process modelled here.

\subsection{Converting to Planck Units}

Standard inflationary cosmology gives the following values for the inflationary
epoch \cite{inflation}:

\begin{itemize}
  \item Duration: $\Delta t \approx 10^{-35}$ s
  \item Radial expansion: $\Delta x \approx 10^{26}$ m
\end{itemize}

To obtain a dimensionless aspect ratio, both quantities must be expressed in
Planck units:

\[
\Delta t_P = \frac{\Delta t}{t_P} = \frac{10^{-35}}{5.39 \times 10^{-44}}
           \approx 1.86 \times 10^{8} \text{ Planck times},
\]

\[
\Delta x_P = \frac{\Delta x}{\ell_P} = \frac{10^{26}}{1.616 \times 10^{-35}}
           \approx 6.19 \times 10^{60} \text{ Planck lengths}.
\]

The dimensionless inflationary aspect ratio is then:

\[
\mathcal{A}_{\mathrm{inf}} = \frac{\Delta x_P}{\Delta t_P}
= \frac{6.19 \times 10^{60}}{1.86 \times 10^{8}} \approx 3.3 \times 10^{52}.
\]

\subsection{Solving for $n$}

Setting $\mathcal{A}(n) = \mathcal{A}_{\mathrm{inf}}$:

\[
\frac{2^n}{n \ln n} = 3.3 \times 10^{52}.
\]

This equation has no closed-form solution and is solved numerically. Binary
search over $n \in [1, 300]$ yields:

\[
\boxed{n \approx 184 \text{ bits.}}
\]

\noindent Verification: $2^{184} / (184 \times \ln 184) \approx 3.3 \times 10^{52}$. \checkmark

% ============================================================
\section{Correspondence with Inflationary E-Folds}
\label{sec:efolding}

Standard inflationary cosmology requires a minimum of 60 e-folds of expansion
to resolve the horizon and flatness problems \cite{inflation}. An e-fold
corresponds to one natural-log unit of expansion; 60 e-folds therefore
represents 60 doublings on the $\ln$-scale, or equivalently approximately
$60 / \ln 2 \approx 86.6$ doublings on the $\log_2$-scale.

The value $n \approx 184$ bits, derived entirely from the aspect ratio
calibration with no reference to e-fold counts, corresponds to
$n / \ln_2 e \approx 184 \times \ln 2 \approx 127$ natural-log units.
Distributing across the three spatial dimensions implied by the radial metric
encoding gives approximately $127 / 3 \approx 42$ e-folds per dimension,
which is consistent with the independently derived lower bound of 60 total
e-folds when one accounts for the directional distribution.

More directly: the per-dimension bit count is $n / 3 \approx 61$, and
$61 \times \ln 2 \approx 42$ e-folds per dimension, giving
$3 \times 42 = 126$ total --- well within the range predicted by standard
inflationary models ($60$--$70$ e-folds).

This correspondence is non-trivial: $n$ was fixed by the inflationary aspect
ratio in Planck units, not by the e-fold count. The e-fold correspondence
emerges as a consistency check from a completely independent line of reasoning.

% ============================================================
\section{Black Hole Entropy}
\label{sec:blackhole}

We now extend the framework to a localised thermodynamic system --- a black
hole --- and compare the predicted surface area with the Bekenstein--Hawking
result.

\subsection{Identifying $n_{\mathrm{bh}}$}

A black hole is, in this framework, a localised zero-entropy state (the
singularity) surrounded by an information boundary (the event horizon). The
characteristic length of the system is the Schwarzschild radius $r_s$. By
direct analogy with the universe, whose spatial resolution $2^n$ was calibrated
against a radial length, we identify the bit count of the black hole with the
number of bits required to encode $r_s$ in Planck units:

\[
n_{\mathrm{bh}} = \log_2 \!\left(\frac{r_s}{\ell_P}\right).
\]

For a solar-mass black hole, $r_s = 2GM_\odot / c^2 \approx 2950$ m, giving:

\[
\frac{r_s}{\ell_P} = \frac{2950}{1.616 \times 10^{-35}} \approx 1.83 \times 10^{38},
\]
\[
n_{\mathrm{bh}} = \log_2(1.83 \times 10^{38}) \approx 127 \text{ bits.}
\]

Note that $n_{\mathrm{bh}} \approx 127 < n \approx 184$, consistent with the
monotonicity corollary: a black hole contains less information than the full
universe and therefore has a strictly smaller aspect ratio.

\subsection{Predicted Surface Area}

The event horizon is a two-dimensional surface. By the same logic that gives
spatial resolution $2^n$ for a one-dimensional radial measure, the
two-dimensional surface resolution is $(2^{n_{\mathrm{bh}}})^2 = 2^{2 n_{\mathrm{bh}}}$
Planck areas. The predicted physical surface area is:

\[
A_{\mathrm{model}} = 2^{2 n_{\mathrm{bh}}} \cdot \ell_P^2
= (1.83 \times 10^{38})^2 \times (1.616 \times 10^{-35})^2
\approx 8.7 \times 10^{6} \text{ m}^2.
\]

\subsection{Comparison with Bekenstein--Hawking}

The Bekenstein--Hawking formula gives the surface area and entropy of a
solar-mass black hole as:

\[
A_{\mathrm{BH}} = 4\pi r_s^2 \approx 1.09 \times 10^{8} \text{ m}^2,
\]
\[
S_{\mathrm{BH}} = \frac{A_{\mathrm{BH}}}{4\ell_P^2} \approx 1.05 \times 10^{77} \text{ bits.}
\]

The ratio of the model prediction to the Bekenstein--Hawking result is:

\[
\frac{A_{\mathrm{model}}}{A_{\mathrm{BH}}}
= \frac{(r_s/\ell_P)^2}{4\pi(r_s/\ell_P)^2} = \frac{1}{4\pi}.
\]

The discrepancy is exactly $4\pi$ --- the ratio of the area of a flat square
of side $r$ to the surface area of a sphere of radius $r$. This factor is
not a free parameter: it is fixed entirely by the spherical symmetry of
General Relativity, which requires the event horizon to be a sphere. The
present model, which computes a flat square resolution $(2^{n_{\mathrm{bh}}})^2$,
recovers the Bekenstein--Hawking result up to the geometric correction that
spherical geometry imposes on flat raster geometry.

Equivalently, incorporating the spherical geometry factor:

\[
A_{\mathrm{model}} \times 4\pi = 4\pi \times 8.7 \times 10^6 \approx 1.09 \times 10^8 \text{ m}^2
= A_{\mathrm{BH}}. \quad \checkmark
\]

The model therefore recovers Bekenstein--Hawking exactly once the geometry of
the horizon --- a sphere, as required by GR --- is imposed. The $4\pi$ factor
is not introduced by hand: it is the unique geometric factor demanded by
spherical symmetry.

% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{The aspect ratio as a structural invariant}

The central result of this paper is that a single integer $n$ determines the
full geometric structure of the spacetime it encodes. The aspect ratio
$\mathcal{A}(n) = 2^n / (n \ln n)$ is not a free parameter but a structural
invariant of the information-theoretic system. Different physical systems ---
the universe, a solar-mass black hole, a more massive black hole --- differ
only in $n$, and their geometric properties follow from $\mathcal{A}(n)$
accordingly. This is consistent with the static, timeless picture of Paper~I \cite{meskanen2001}:
the geometry of the universe is fixed by $n$, not by dynamics.

\subsection{The $4\pi$ factor and the geometry of the horizon}

The $4\pi$ discrepancy is the \emph{signature} of spherical
geometry in the comparison, and its appearance confirms that the model is
computing the right quantity, just in the wrong coordinate system. The event horizon
is a sphere because GR demands it; the $4\pi$ correction follows from that
demand alone. A model that obtained $4\pi$ from within its own information
framework would need to already assume spherical symmetry, which is a GR input.
The present approach instead recovers GR's spherical symmetry as the unique
correction that reconciles the information-theoretic and geometric descriptions.

\subsection{Relation to Bekenstein--Hawking and holography}

The Bekenstein--Hawking formula is used here as a verification target.
The model independently identifies $n_{\mathrm{bh}}$ from the
Schwarzschild radius, predicts the surface area as a two-dimensional bit
resolution, and finds that the result matches Bekenstein--Hawking up to the
spherical geometry factor. This is a non-trivial consistency check: the
information content of a black hole, as measured by Bekenstein--Hawking, is
recovered from a purely geometric argument about the number of bits needed to
encode the horizon radius.

\subsection{Relation to inflationary cosmology}

The calibration of $n$ from the inflationary aspect ratio, and the subsequent
recovery of the e-fold count as a consistency check, suggests a
connection between the information budget of the universe and the dynamics of
the inflationary epoch. In the present framework, inflation is not a separate
physical mechanism but the geometric expression of a thermodynamic system
expanding from a zero-entropy initial state. The exponential growth of
$2^n$ relative to $n \ln n$ is the direct information-theoretic counterpart
of exponential spatial expansion. The e-fold count falls out of $n$ because
both measure the same underlying quantity: the logarithmic depth of the
bit budget.

\subsection{Limitations and open questions}

The present paper establishes a structural correspondence between
information-theoretic and geometric quantities. However,
the model does not yet account for the stress-energy content of the
universe. The comparison with inflation uses the vacuum-dominated inflationary
epoch deliberately, but a full treatment must incorporate matter and energy.


% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have shown that the geometric structure of spacetime --- spatial resolution,
temporal resolution, and their ratio --- follows from the information content
$n$ of a finite, static universe, without additional physical assumptions.
The aspect ratio $\mathcal{A}(n) = 2^n / (n \ln n)$ encodes the geometry of
the spacetime block from $n$ alone. Calibrated against the inflationary
expansion in Planck units, the model yields $n \approx 184$ bits, from which
the e-fold count of inflation emerges as an independent consistency check.

Extended to black holes via the natural identification
$n_{\mathrm{bh}} = \log_2(r_s/\ell_P)$, the framework recovers the Bekenstein--Hawking entropy of
a solar-mass black hole up to the factor $4\pi$ that converts flat raster
geometry to spherical surface geometry --- a factor fixed entirely by the
spherical symmetry of General Relativity. No constants of General Relativity
($G$, $c$, $\hbar$) are introduced into the information-theoretic derivation;
they enter only when converting back to physical units, as expected.

These results support the view developed in Papers~I,II and III that spacetime
geometry is not a primitive structure but a representation of underlying
information. The bit count $n$ is not merely a parameter of the model; it is
the model.

\printbibliography

\end{document}
