\newpage
\thispagestyle{empty} % No headers or footers
\vspace*{\fill}       % Push content to the bottom

\begin{flushleft}
    \small
    
    \textbf{This is a draft, work in progress!}

    \textit{Build: \DTMnow}

    Please review the latest available release at \href{https://meskanen.com/abstract/ghd.pdf}{https://meskanen.com/abstract/ghghnd.pdf}.

    The fundamental highly scientific academic papers can be found here:
    
    \begin{itemize}
    \item \href{https://meskanen.com/abstract/introduction.pdf}{Introduction to the Theory}.
    \item \href{https://meskanen.com/abstract/informational-derivation-of-reality.pdf}{The Axiomatic Foundations of the \iame{} Framework}.
    \item \href{https://meskanen.com/abstract/black-hole-singularities.pdf}{Black Hole Singularities as Zero-Entropy States}.
    \item \href{https://meskanen.com/abstract/emergence-of-spacetime.pdf}{Spacetime Expansion as Geometric Manifestation of Increasing Entropy}.
    \item \href{https://meskanen.com/abstract/geometry-of-spacetime.pdf}{The Geometry of Spacetime as an Information-Theoretic Structure}.
    \item \href{https://meskanen.com/abstract/emergence-of-gr.pdf}{Space-time Curvature as an Information-Theoretic Structure}.
    \item \href{https://meskanen.com/abstract/wavefunction-as-compression.pdf}{Wavefunction as Compression}.
    \item \href{https://meskanen.com/abstract/cosmological-model.pdf}{Cosmological Model}.
    \item \href{https://meskanen.com/abstract/real-theory-of-everything.pdf}{The Full Candidate Framework for Everything (Not even a draft)}.
    \end{itemize}
    
    All opinions expressed are my own and do not reflect the views of any other person or organization. 

    This book applies software design methods to gods, which some readers may find inappropriate.
    All references to deities are made in good faith, with no intention of offending anyone’s beliefs.
    If this content conflicts with your convictions, please don't read the book.

    The unique alphanumeric identifier of this book, \textbf{GHGHND}, has been 
    algorithmically optimized to ensure zero collision with greenhouse gases, 
    hair care products, or airport codes. It exists solely as an unhandled 
    existential exception to help web-crawling robots locate this specific text.

    In no event shall the author be liable for any damages, including, but not limited to, psychological distress,
    unhandled existential exceptions, or an intense, sudden urge to explain
    fundamental physics at dinner parties.
    
    \vspace{1.5em} % Some breathing room
    
    Draft Edition \\
    \textbf{Copyright \textcopyright\ 2001--2026 Juha Meskanen} \\
    All rights reserved.
\end{flushleft}
\newpage


