\documentclass[11pt]{article}
\input{../../common.tex}
\input{../../glossary.tex}
\addbibresource{../../references.bib} 


\title{Formal Derivation of $D-\psi-G$ Trinity}

\date{2025}

\begin{document}
\maketitle


\section{Three Descriptive Paradigms of Information}

Building on the ontological equivalence of physical and informational configurations established in Paper I, we
identify three irreducible descriptive paradigms of information sufficient for representing bounded, persistent observers:

\begin{enumerate}
    \item Discrete, finite, set-theoretic,
    \item Spectral Wavefunction,
    \item Geometric - Einstein field equations
\end{enumerate}

These paradigms are ortogonal projections of the same informational substrate. They are descriptively irreducible, and jointly sufficient to express all internally meaningful informational structure.

This triadic structure provides a foundation for unifying quantum theory, spacetime geometry, and information theory within a single informational framework.



\section{Hypothesis}

We hypothetize that observers emerge as configurations that are simultaneously well-defined across all three paradigms.

\begin{enumerate}
\item The discrete, minimal finite rasteriation - the pixels aka particles. Born rule is effectively the dithering system resulting maximally smooth reality under finite resources.
\item The analytic (wavefunction) ensures existence with high measure (optimal spectral compression algorithm)
\item The geometric (GR) ensures separation, persistence, and identity (the only known mechanism for implementing an inside–outside invariant) with high measure (optimal geometric compression algorithm).
\end{enumerate}



\section{$D-\psi-G$ Formalism}

\[
\begin{tikzcd}[row sep=6em, column sep=6em]
& \psi(\mathcal{I}_n)
    \arrow[dl, "\text{Sampling}"] & \\
D(\mathcal{I}_n)
    \arrow[rr, "\text{Geometric}"] & &
G(\mathcal{I}_n)
    \arrow[ul, "\text{Compression}"]
\end{tikzcd}
\]


Each vertex of the triangle represents a mutually equivalent projection of the same underlying informational configuration space $\mathcal{I}_n$: 
\begin{itemize}
    \item $D(\mathcal{I}_n)$: discrete representation,
    \item $\psi(\mathcal{I}_n)$: analytic representation,
    \item $G(\mathcal{I}_n)$: geometric representation.
\end{itemize}
Connecting lines are structure-preserving maps connecting the representations. 
The curved arrows illustrate invertible, commuting transformations between representations. 
The diagram formalizes that no representation is fundamental; all are mutually consistent projections of the same informational object.

\vspace{1em}
\noindent



\section{Paradigms and Descriptions }

\subsection{1. Analytic Representation ($\psi$)}

\textbf{Observer Emergence Principle (Spectral Form):} 
Among the ensemble of all possible wavefunctions \(\Psi\) over configuration space \(\mathcal{C}\), 
the observer-compatible paths \(\gamma \in \mathcal{T}_{\mathrm{obs}}\) are overwhelmingly likely to occur in 
the \emph{minimal-length} wavefunctions, due to algorithmic (Salomonoff) weighting:
\[
\Psi(\gamma) \propto 2^{-\mathcal{L}(\gamma)/2},
\]
where \(\mathcal{L}(\gamma)\) is the minimal spectral encoding length of \(\gamma\). 
High-entropy, incompressible wavefunctions exist but carry exponentially negligible measure; 
thus, the observer emerges almost certainly in the simplest, smoothest, low-entropy wavefunctions.


\[
\begin{array}{c}
\mathcal{T}_{\mathrm{obs}} =
\{ \gamma \in \mathcal{S} \mid \mathrm{Observer}(\gamma)=1 \} \\
\text{\small (Same observer paths from D; basis for compression)} \\
\downarrow \\
\mathcal{L}(\gamma) = \text{Minimal spectral encoding length of } \gamma \in \mathcal{T}_{\mathrm{obs}} \\
\text{\small (Wavefunction as compression: smooth, predictable, low-entropy paths favored)} \\
\downarrow \\
\Psi(\gamma) = \frac{2^{-\mathcal{L}(\gamma)/2}}{\sqrt{\sum_{\gamma' \in \mathcal{T}_{\mathrm{obs}}} 2^{-\mathcal{L}(\gamma')}}} \\
\text{\small (Normalized wavefunction: encodes all observer-compatible paths)} \\
\downarrow \\
P(\gamma) = |\Psi(\gamma)|^2 \\
\text{\small (Born measure: relative likelihood of path \(\gamma\))} \\
\downarrow \\
\delta \int_{\gamma} \mathcal{L}(\text{state}) \, d\lambda = 0 \;\;\Longrightarrow\;\; \text{Geodesics} \\
\text{\small (Minimal-description principle produces paths identical to classical action extrema)}
\end{array}
\]



\subsection{2. Discrete Representation ($D$)}

\[
\begin{array}{c}
\mathbf{P} = \text{Raw informational potential} \\
\text{\small (Infinite unstructured possibilities; no preferred encoding)} \\
\downarrow \\
\mathcal{C} \cong \{0,1\}^{\le \infty} \\
\text{\small (Discrete configuration space; convenient static representation)} \\
\downarrow \\
\mathcal{S} = \{ (s_1,\dots,s_T) \mid s_i \in \mathcal{C} \} \\
\text{\small (Space of all finite/semi-infinite paths through configuration space)} \\
\downarrow \\
\mathcal{T}_{\mathrm{obs}} =
\{ \gamma \in \mathcal{S} \mid \mathrm{Observer}(\gamma)=1 \} \\
\text{\small (Observer Filter: selects paths with stable informational recursion)} \\
\downarrow \\
\mathcal{L}(\gamma) = \text{Minimal spectral encoding length of } \gamma \in \mathcal{T}_{\mathrm{obs}} \\
\text{\small (Compression–geometry duality: favors smooth, predictable, and compressible histories)} \\
\downarrow \\
\Psi(\gamma) = \frac{2^{-\mathcal{L}(\gamma)/2}}{\sqrt{\sum_{\gamma' \in \mathcal{T}_{\mathrm{obs}}} 2^{-\mathcal{L}(\gamma')}}} \\
\text{\small (Spectral realization: wavefunction encodes optimal compression of observer paths)} \\
\downarrow \\
P(\gamma) = |\Psi(\gamma)|^2 \\
\text{\small (Born measure: relative probability of experienced histories)} \\
\downarrow \\
\end{array}
\]




\subsection{3. Geometric Representation $G$}


\begin{gather*}
\mathcal{T}_{\mathrm{obs}} \subset \mathcal{S} \\
\small \text{(Observer-compatible paths through configuration space)} \\
\downarrow \\
\mathcal{I}(\gamma_\lambda) \xrightarrow{\;\Pi_G\;} \mathcal{M}_\lambda \subset \mathbb{R}^d \\
\small \text{(Geometric projection } \lambda\text{)} \\
\downarrow \\
\rho(\mathbf{s}) \sim \mathrm{Lognormal}(\mu,\sigma^2) \\
\small \text{(Emergent density variations from reusable microstructure)} \\
\downarrow \\
\delta \int_{\gamma} \mathcal{L}(\text{state}) \, d\lambda = 0 \;\;\Longrightarrow\;\; \text{Geodesics} \\
\text{\small (Gravitation as geometric manifestation of compression pressure; GR emerges from observer-consistent paths)}
\delta \int_{\gamma} \mathcal{L}_{\mathrm{geom}} \, d\lambda = 0 \\
\small \text{(Geodesics as minimal-description trajectories)} \\
\end{gather*}


\end{document}



