\chapter{Closure}

\subsection*{The Observer Contains the Universe}

After many years of thought and numerous thought experiments, I arrived at a conclusion that I tried to falsify.

I began by trying to assume as little as possible. Yet I realized that ``assuming nothing'' is itself an assumption. In any axiomatic system, at least one starting point is required.

I started with minimal assumption: there exists Alice --- a conscious being who feels pain, thinks, and experiences a world.

I ended up to a static, self-contained block universe made of information. In static system interaction exists. No signals travel from an ``external'' world into Alice,
Everything Alice experiences --- distant stars, other people, her own body, the passage of time --- must already be encoded within her.
What appears as an external universe is a highly compressed construction \emph{within} Alice herself.

The vast cosmos we contemplate is not containing us.  

It is the other way around: the universe we can ever know is a subset of the observer.

The block universe simply ``is.'' What we call interaction, causality, and the flow of information are internal processes of self-organization and compression taking place inside the conscious structure. Time is the internal ordering of configurations. The geometric self with clear boundaries is the most compressible stable pattern available to the observer.

In the end, the only thing I could not dissolve was the observer. Everything else followed.

The conclusion is the recognition that consciousness is not a small phenomenon inside a large universe, but that the only universe we can ever know is the one arising within consciousness.



\section{Neural Networks}

The recent development of neural networks provides a powerful analogy.

Modern AI systems are massively parallel, interconnected structures with weighted connections.
These weights are almost entirely static data.
During interaction, there is no explicit procedural simulation of physics,
no internal clockwork universe ticking forward step by step.

When a question is asked, the response is a projection from an already compressed informational structure.
The “computation” is closer to indexing, interpolation, and constraint satisfaction
than to sequential execution.

The distinction between “running” and “not running” becomes secondary.

There is no physics engine inside these systems.
Yet they can produce realistic simulations, motion, causality, and physical intuition.



\section{Free Will Reconsidered}

Earlier, we wondered: could software ever allow free will? Deterministic code—branching if-else statements, fixed CPU instructions—appears to preclude choice.
The computer just runs the software; it cannot “choose otherwise,” regardless of what it might feel like to be Alice executing that program.

However, as we have demonstrated, consciousness can arise from static noisy information, as there is no external reader that would
dictate the order in which the information should be ordered. All arrangements are equal.

Consider a simple line of code:

\begin{verbatim}
if (a < 3){
    A
} else {
    B
}
\end{verbatim}

Deterministic? Yes. Sequential? Yes. But what does `3` really mean? In two’s complement, it might be interpreted as `3` or `-1`. As a floating-point number, it could be entirely different.  

The code, such as ``is less'' comparison  itself is nothing but data burned into CPU. Programmers interpret it as “instructions,” just as integers can be signed or unsigned.
There is no external device that determines the order in which bits get read. All the $2^n$ arrangements are equally real. 

As the number of bits \(n\) increases, the number of consistent interpretations grows combinatorially. Each interpretation defines a slightly different “Alice” inside the informational structure.
Some of these preserve consciousness; some do not.  Some may choose $A$, some $B$.

Alice is not forced to flow through this code like a train on tracks. All possible versions already exist, corresponding to every consistent interpretation of the bits.
Execution does not create Alice; it merely reveals one perspective among many.  There is Alice both in $A$ and $B$. 

When we run our DNA simulations in a computer, we don't create Alice. We just explore one of the possible ways to arrange information that already exists. We are exploring and drawing a map over the landscape that already exists.

Which version does Alice herself experience? Most likely, the one corresponding to the most compressible, smooth, and coherent description—the version that dominates the observer-conditioned measure:

\[
\mathbb{P}(\text{Alice}_i) \propto 2^{-\mathcal{C}[\text{Alice}_i]}.
\]

Determinism at the CPU level does not constrain subjective experience. Execution is selection, not creation. And free will and pain? 

Alice navigates the space of all consistent interpretations, experiencing deliberation, anticipation, pain, and choice—not because the software code grants it,
but because the the most compressible configuration.

Just as 'red' is the internal feeling of 700 nm light waves, reasoning and pain are simply what it feels like to be the most compressible, and therefore the most inevitable, arrangement of information.


\section{Conclusion}

\begin{itemize}
  \item Alice exists
  \item Alice can be simulated
  \item The simulation can be transformed into static data
  \item Alice is preserved under that transformation
  \item Therefore, execution is irrelevant
  \item Therefore, time is internal to Alice
  \item Therefore, reality is structure rather than code
  \item The most compressible descriptions of Alice dominate the measure
  \item Threfore, Alice finds herself in the a law-like universe fine tuned to support her existence
  \item Because predictable information compresses best
  \item Predictability is what we call the laws of physics.
\end{itemize}



\section{Philosophical Satisfaction}

We have arrived at a zero-parameter framework which we can intuitively understand.

In principle, it allows us to compute probabilities for any structure, including our own existence. 

More importantly, it clarifies what the universe is made of-pure abstract information.

In mathematics, no object has ontological privilege over another.
True is no more fundamental than false; the two form a binary object, much like a coin.
Heads don't exist more than tails. An empty set is no less legitimate than a non-empty one.
Infinity does not 'weight' more than finite. 

If there were a rule that favored nothingness over something, we would again have to ask about the origin of such a rule—who set it?
Non-empty structures are just as real or unreal as empty sets. Both exist because nothing forbids them.

So what kind of a thing is $2+2=4$ ? It is abstract.

We can create (or more precisely, reveal, explore) those abstract universes ourselves. Modern virtual games demonstrate this principle: realistic avatars inhabit richly detailed worlds.
In principle, these worlds could be expanded until they are informationally equivalent to our own universe. We can actually
understand them. The more we spend time in those virtual game worlds, the easier that abstractness is to understand. Bullets and guns in them are not
real, they are just software runnign in a computer, physics is just equations executed in a computer. And the same logic could
be implemented as wooding computer with wooden spheres rolling through wooden ports.

What is common between modern electric pulses in silicon chips and their wooden counterparts is structure.
It doesn't weight. It doesn't have an electric charge. It cannot be measured with any measuring device. 

