\chapter{Definition of Foundational Theories}

\section{Minimal Requirements}

To build a fundamental theory of physics, one obviously needs to know the minimum requirements
for a theory to qualify as a fundamental one.

Based on common sense, the theory must answer three basic questions:
what exists, how likely different possibilities are, and how complex structures arise from simple rules.

A candidate theory must therefore specify three ingredients: ontology, measure, and emergence.

\subsection*{1. Ontology (What exists)}

First, we must define the basic building blocks of reality.
This means specifying a well-defined space of possible states of the system.
In other words: what are the most elementary ``things'' the theory talks about?
For example, in Quantum Mechanics, the basic objects are quantum fields (or state vectors in Hilbert space) defined over spacetime.
All particles and interactions are described as excitations of these fields.

\subsection*{2. Measure (What is typical or likely)}

A theory must also explain what is typical among all possible states.
Not all states are equally relevant: some are more probable or more ``natural'' than others.
This requires a rule that assigns weights or probabilities to different states in the state space.
For example, in Quantum Mechanics, the Born rule assigns probabilities to measurement outcomes
based on the quantum state. This tells us which outcomes are likely to be observed.

\subsection*{3. Emergence (How structure arises)}

Finally, a fundamental theory must explain how complex structures arise from simple underlying rules.
This is the bridge between microscopic laws and the macroscopic hierarchical structures we observe.
For example, chemistry emerges from Quantum Mechanics:
atoms form molecules, and molecules form complex structures, even though the underlying description
is entirely in terms of quantum fields and their interactions.

\subsection*{Hierarchy}

Without ontology, the theory has nothing to describe. Without a measure, the theory cannot make predictions.
Without emergence, the theory cannot connect to the macroscopic hierarchical structures we observe.

As with any well-designed modular software, the design should also be mutually orthogonal without cyclic dependencies.
Ontology defines the domain. Measure provides weighting on that domain, so it depends on the ontology.
Emergence depends on both ontology and measure. This yields a clean layered design free of cyclic dependencies.


\section{Cyclic Dependencies in Current Theories}

Let us stress test our design against existing theories, e.g., QM.

In standard QFT:
\begin{itemize}
    \item Ontology: Hilbert space states (or quantum fields).
    \item Measure: Born rule, which computes probabilities from the quantum state.
\end{itemize}
The fields (or state vectors) exist independently of the Born rule. No cyclic dependency arises. This is a strength of the framework.

However, we deliberately chose one clear case here. Many candidate theories implement these three ingredients to varying degrees of completeness.

\section{Classifying Current Theories}

\subsection*{Ontology (What Exists)}
\begin{itemize}
    \item \textbf{Physical primitives:} Particles, fields, spacetime (e.g., Standard Model, General Relativity)
    \item \textbf{Information-theoretic:} Binary distinctions, bits (e.g., Wheeler's ``It from Bit'')
    \item \textbf{Computational:} Programs, Turing machines or cellular automata
    \item \textbf{Mathematical:} Abstract mathematical structures (e.g., Tegmark's Mathematical Universe Hypothesis)
\end{itemize}

\subsection*{Measure (Typicality)}
\begin{itemize}
    \item \textbf{Explicit probabilistic measure:} Classical statistical mechanics (e.g., Boltzmann distributions)
    \item \textbf{Algorithmic measure:} Program-weighted probability (e.g., Solomonoff prior in Algorithmic Information Theory)
    \item \textbf{Implicit or undefined measure:} Many multiverse or mathematical universe proposals
    \item \textbf{Emergent measure:} Measure derived from structural or dynamical properties
\end{itemize}

\subsection*{Emergence}
\begin{itemize}
    \item \textbf{Dynamical laws:} Evolution governed by differential equations (e.g., GR, QFT)
    \item \textbf{Statistical emergence:} Order arising from fluctuations and typicality (e.g., Boltzmann)
    \item \textbf{Computational emergence:} Structure arising from execution of simple programs or rules (e.g., cellular automata)
    \item \textbf{Undefined or implicit emergence:} Present in many abstract frameworks without explicit mechanism
\end{itemize}

\subsection*{Placement of Selected Modern Approaches}
While the tripartite classification is useful, some contemporary proposals test its boundaries:
\begin{itemize}
    \item \textbf{String theory / M-theory:} Primarily structure-first with a rich ontology (higher-dimensional objects and geometries). It aims to derive particles, forces, and possibly spacetime itself from a single underlying structure, though it still faces significant measure and emergence challenges (landscape problem).
    \item \textbf{Loop Quantum Gravity and causal set theory:} Ontology-first (discrete spacetime or spin networks). They seek to recover General Relativity and quantum field theory as effective emergent descriptions.
    \item \textbf{Quantum information / constructor theory approaches:} Strong emphasis on possible vs.\ impossible transformations, sometimes blurring the line between ontology and measure.
\end{itemize}

Note that in certain advanced proposals (e.g., some interpretations of Many-Worlds or holographic approaches), the separation between ontology and measure can become strained, as the measure may appear partly emergent from the dynamics or entanglement structure.



\section{Conclusions}

Of the three pillars, emergence is arguably the least fundamental.
Given a sufficiently rich ontology together with a suitable measure or dynamical law, higher-level structure may arise automatically.
In that sense, emergence can often be viewed as a consequence rather than an independent primitive.

Consequently, foundational theories can often be classified according to which of the two remaining pillars they prioritize:

\begin{enumerate}
    \item \textbf{Measure-first theories.}
    These begin with a space of possible configurations and a measure over that space, explaining observed reality in terms of what is typical, probable, or algorithmically simple.
    Examples include Boltzmann-style statistical reasoning, Algorithmic Information Theory, and some multiverse proposals.

    While conceptually elegant, such approaches have not yet yielded fully predictive physical theories capable of deriving the rich dynamical structure observed in nature.

    \item \textbf{Structure-first theories.}
    These begin with explicit dynamical laws, equations of motion, or symmetry principles, and derive observable behavior from them.
    General Relativity and Quantum Field Theory are the paradigmatic examples.

    These frameworks are extraordinarily successful empirically, providing precise and testable predictions.
    However, they rely on substantial prior assumptions---such as spacetime structure, gauge symmetry, or the form of the action---which they do not themselves explain.
\end{enumerate}

This exposes a central tension in theoretical physics:
approaches that minimize assumptions often struggle to generate predictive structure, while approaches that achieve remarkable predictive power typically rest on foundational assumptions they cannot themselves justify.


