\chapter{Running DNA Simulations}

\section{Thought Experiment}

The current technology does not allow us to run full scale human DNA simulations yet. However, this does not prevent us from exploring it as a thought experiment.
So we digitize a human genome and run it inside a sufficiently 
detailed computer simulation. We also simulate a sufficiently large world with it to avoid our simulated human
developing phsychosis in empty space.  Nine months later, in simulation time, our virtual copy takes its first breath 
in its simulated world.


\section{The Nature of the Simulated World}

How would such a simulated human perceive its environment? Would it sense the limited memory space of the 
computer running it? Would it be able to bump its head against the upper boundary of RAM and feel pain? 
Would the flipping of bits tickle its nose, or the rotation speed of a hard drive make it dizzy?

Would it eventually discover that its entire universe is driven by storage devices, memory chips, and an 
overclocked multi-core CPU?

\section{Sense of Reality}

The simulated human is not created \emph{within} our universe. It does not consist of real-world particles such 
as electrons or quarks. Instead, it exists entirely within a virtual universe that we simulate alongside it. 
As a result, it has no access to our physical hardware. It cannot observe transistors, memory cells, voltages, 
or processor clocks.

The only thing the simulated human can study is the internal structure of its own virtual world.

Within that world, there are virtual particles, virtual forces, and virtual laws of physics. When the 
simulated human bangs its head against a simulated wall, the simulated particles in the wall respond exactly 
as the laws of that virtual universe dictate. 

Every measurement the simulated human performs inside its universe will match those we real people
carry out in our real world.
The outcomes of experiments will match the predictions of the simulated physical laws, just as our 
measurements match the laws of physics in our own universe.

To the simulated observer, the experience is indistinguishable 
from how real particles behave when we humans bang our heads against real walls.

This is because both the real world and the simulated world are \emph{axiomatic systems}. Mathematics does 
not care whether it is applied to apples, bananas, electrons, or bits. The statement $2 + 2 = 4$ holds 
regardless of the physical substrate that implements the system.

For the simulated human, there is no experiment it could perform that would reveal the presence of the computer 
running the simulation, because that computer exists outside the axioms of its universe.

From the inside, the simulated universe would feel precisely as real as our universe feels to us.



\section{Substrate Independency}

The brain of a computer---its central processing unit (CPU)---consists of a set of electric switches called transistors. The CPU does not need to be an electric device.
Just as $2+2=4$ holds for both apples and bananas, simulations should work regardless of the substrate on which they are implemented.

In theory, one could implement a DNA simulation as a mechanically operating computer consisting of wooden components. Instead of using transistors in a silicon chip to control electrons, one could use wooden parts on a plywood platform to control wooden balls. When such a machine stepped through its logic, a virtual human would take its first steps in its virtual universe.

What is this strange phenomenon that creates consciousness and pain from a jerking pile of wooden pieces?

If a huge number of moving wooden components can create pain, then what does one moving piece of wood create?

Can current physics even describe this action?

\subsection*{Entropy}

No man-made device is perfect, and wooden \texttrademark{} is no exception. Friction, tolerances, and things like that introduce resonances and other unintentional vibrations to the operation of wooden \texttrademark{}.
If the actual logic in the wooden \texttrademark{} creates a virtual universe with pain and consciousness, then what do these unintentional side effects create? Do they get reflected in some form into the created virtual world too?

Would the virtual fellow in its virtual universe discover these in the form of strange quantum foam? Would it observe them as strange cosmic background radiation with 2.725\,K temperature? Maybe that 
indeed explains why we measure quantum foam and cosmic background radiationin our universe. We are being simulated in a wooden Universal Turing Machine!

Maybe not so, but at least it would be difficult to argue why large movements of wooden components would count, but their small resonances would not.

\subsection*{Number of Spheres per Second}

How would the clock speed of the machine running the simulation appear in the created simulated universe?
Would the simulated human observe that particles in its universe appear to follow some strange abstract square wave function, whose origin it could not explain, but which it might end up calling Wintel's (TM) abstract square wave function?

How would the human simulated in wooden computer sense the workings of such a computer? Due to the large number of concurrently rolling spheres,
the simulated human could conclude that the wave function must be complex-valued with phase coherence,
and imaginary numbers would provide a natural formalism.

In addition to electric and wooden computers, it is easy to picture rich set of other possible ways to implement computers, and therefore,
systems potentially capable of creating virtual universes with conscious observers.


\subsection*{Thermostat as the Source of Consciousness}

Computer software is ultimately a sequence of bits—nothing more than a series of binary switches. Theoretically, one could use a thermostat, a device with only two states (open and closed), to describe any computational procedure.

If we allow a temperature to vary over time, the thermostat could go through the binary code of the DNA procedure. If the mathematics holds, pain and
conscious entities should emerge.

The simulated fellow would be totally unaware of the fact that a trivial thermostat is responsible for the illusion of its existence.
Crucially, the thermostat itself cannot be regarded as conscious by any means; those  arguing that matter itself is conscious may miss be missing the mark.

Correspondingly, running such DNA simulations on any type of computer does not make the computer itself conscious or pain-sensitive.

It is still the very mechanical deterministic system stepping through its symbol tape without any choice. Yet, the rattling of that thermostat creates a virtual parallel
universe in which a conscious human marvels at the deep nature of reality.

\subsection{Cities as the source of Consciousness}

A person walking through two subsequent doors implements the logical operation called \texttt{AND}. If one can pass through using either the left or right door, then one gets the \texttt{OR} operation. What complex logic do billions of human beings create by navigating streets and passing through doors on their way to work? 

\subsection{Pencil and Pen}

Even a regular pencil and a piece of paper could be the source of consciousness. Start writing down the evolution of DNA with pencil and pen, and soon virtual people suffer tooth pain in their virtual universe. Both pencil and ballpoint pen should work equally well. Due to the higher friction, the temperature of the cosmic background radiation in a universe created with pencil might be a bit higher though!


\subsection{Sea}

One possible source of consciousness could be the surface of a sea. In theory, waves and ripples of a sea could describe a DNA simulation in which conscious observers marvel at the wonderful properties of their universe (like those caused by heavy rain during the annual monsoon season).



\subsection{The Universe}

The whole universe, with its planets and stars, could be the source of a consciousness.


\subsection{Substrate Independency}

The only conclusion one can draw from these is that whatever it is that we call consciousnes and pain must be subtrate independent.
The source of pain cannot be any physical attribute, such as mass, electric charge, elementary particle such as photon, because it is always possible to find an implementation where such a property does not play any role.

Computer software is nothing but a sequence of bits, trivial on/off switches. Correspondingly, one could also use a thermostat to describe any procedure.

The only common factor between different implementations appears to be the logic they are running. And logic is nothing but data - information.


\section{Recursion and the Ontological Parity of Simulations}

If we can utilize electrons or even macroscopic components to create virtual universes that replicate the biological and structural motifs of our own—such as DNA—we reach a logical crossroads. Because these simulated observers are functional duplicates of their "real-world" counterparts, they will inevitably begin exploring their own substrate. 

They will discover the principles of computation and, eventually, construct their own Turing Machines. The procedure these virtual entities use to simulate their own existence is identical to the procedure we used to create them. We can express this transition mathematically. If our world is $r_n$ and the simulated world is $r_{n+1}$, the mapping is:
\[ r_{n+1} = f_{\text{DNA}}(r_n) \]

This nested stack of simulations continues as long as the host level contains sufficient computational density to support the sub-level. Because we know the deterministic nature of the Turing Machine we used to initiate the first step, we must admit that the relationship is strictly recursive:
\[ r_{n+k} = f_{\text{DNA}}(r_{n+k-1}) \]

In a recursive formula, it is notoriously difficult to argue for "ontological seniority." There is no parameter within the $f_{\text{DNA}}$ function that distinguishes a "real" world from a "virtual" one; the operator remains invariant across all levels of the recursion. The logical conclusion is that our "base" reality is as computationally contingent as the simulations we produce. To an observer inside the recursion, the substrate is always invisible; we perceive our level as "solid" simply because we are defined by the same logic that governs it.

There is, however, a significant physical constraint to this hypothesis: the \textit{Information Bottleneck}. Simulating the human genome, let alone the consciousness of eight billion humans and the staggering complexity of $10^{22}$ stars in the observable universe, requires an astronomical amount of information. 

Those sub-simulations would soon run out of information.


\section*{Subjective Experience and Unknown Procedures}

Despite their virtual nature, these abstract, virtual simulated universes might possess very "real" emergent properties. Pain, joy, and consciousness
could emerge in them. 


\section{Categorizing Feelings}

Science has mapped much of the terrain of human emotion, yet debate remains about its precise structure. While sophisticated systems exist for categorizing what people feel, there is no single universally accepted “periodic table of emotions.” Researchers instead use different frameworks depending on whether they focus on brain chemistry, facial expressions, evolutionary function, or cognitive interpretation.

Several major approaches currently dominate the scientific study of human emotional experience.

\subsection{Basic Emotions}

In the 1970s, psychologist Paul Ekman argued that humans possess a small set of biologically universal emotions that are expressed through recognizable facial patterns across cultures. These commonly cited emotions include:

\begin{itemize}
\item Happiness
\item Sadness
\item Fear
\item Disgust
\item Anger
\item Surprise
\end{itemize}

Within this framework, emotions are interpreted as evolutionary survival mechanisms. Fear discourages risky encounters with predators, while disgust helps avoid contaminated food.

\subsection{The Color Wheel Approach}

Robert Plutchik expanded this perspective by proposing a ``Wheel of Emotions,'' in which primary emotions can combine to produce more complex emotional states, analogous to the blending of colors.

\begin{center}
\begin{tabular}{ll}
\textbf{Basic Combination} & \textbf{Resulting Emotion} \\
Joy + Trust & Love \\
Anger + Disgust & Contempt \\
Fear + Surprise & Awe \\
\end{tabular}
\end{center}

In this model, emotions also vary in intensity and may exist in opposite pairs.

\subsection{Dimensional Models}

Many contemporary researchers prefer dimensional approaches, such as the Circumplex Model of affect. Instead of assigning discrete labels to emotions, experiences are positioned along two primary axes:

\begin{itemize}
\item \textbf{Valence:} the degree to which an experience is pleasant or unpleasant.
\item \textbf{Arousal:} the level of physiological activation or energy associated with the state.
\end{itemize}

For example, serenity and excitement are both positively valenced states, but serenity corresponds to low arousal while excitement corresponds to high arousal.

\subsection{Pain and Complex Emotional States}

Scientific discussions often distinguish between physical sensation and emotional state, although these systems interact closely.

\begin{itemize}
\item \textbf{Nociception (Physical Pain):} neural processes that signal potential tissue damage.
\item \textbf{Affective Pain (Emotional Pain):} emotional distress such as social rejection, which engages overlapping neural circuits involved in physical pain processing.
\item \textbf{Complex Emotions:} experiences such as nostalgia, schadenfreude, or ennui require higher-level cognition and social interpretation, making them difficult to isolate through simple neural measurements.
\end{itemize}

\section{Why the Question Remains Unsettled}

A central debate in affective science concerns whether emotions are biologically hard-wired or cognitively constructed.

\begin{itemize}
\item \textbf{Universalist theories} propose that emotions arise from dedicated neural circuits that evolved for survival.
\item \textbf{Constructionist theories} argue that emotions are conceptual interpretations that cultures impose on underlying bodily sensations.
\end{itemize}

An open question follows naturally: if a culture lacked a linguistic category for a particular emotion, would individuals still experience it in the same form?

\section{The Hard Problem of the Substrate}

In previous chapters we introduced the principle of \textit{substrate independence}. If a simulation of biological processes is logically complete, then the simulated observer
should behave in every measurable way as if it possesses consciousness.

This leads directly to what philosophers call the explanatory gap. Physical theories describe how matter behaves---how particles interact, how energy moves, and how neural circuits process signals.
Yet these descriptions remain silent about \textit{qualia}, the subjective character of experience.

Modern physical theories, from general relativity to the Standard Model of particle physics, contain no variables corresponding to sensations such as pain or love.
From a purely functional perspective, pain may be modeled as a signal indicating a threat to system integrity. However, for the observer undergoing the experience,
pain is not merely a signal; it is an intensely felt state.


\section{The Mathematical Silence of Pain}

Suppose the state of a simulated human brain is represented as a large, time-evolving matrix \(M(t)\) encoding the activity of all neural elements.

The system evolves according to a transition rule

\[
S_{t+1} = \Phi(S_t, I_t),
\]

where \(S_t\) represents the system state and \(I_t\) represents input from the environment.

Within this framework we can track every state transition and every simulated neural signal.
Yet the formalism itself contains no explanation for why a particular configuration of information should correspond to a subjective experience.

If the simulation is paused, the experiential process halts even though the informational structure remains stored in memory.
If the simulation is run at a drastically slower computational rate, the internal observer may still experience a continuous flow of subjective time.
The mathematical description therefore captures functional dynamics but does not explain why those dynamics are accompanied by phenomenology.

Science currently lacks a clear bridge law that connects informational configurations with specific qualitative experiences.
For example, physics can describe the wavelength of light associated with the color red, yet the equations themselves do not derive the felt quality of ``redness.''


\section{The Taxonomy of Unimplemented Emotions}

A further question arises: is the emotional spectrum observed in humans the complete set of possible conscious states, or merely the subset produced by biological evolution?

Evolution operates as a constrained engineer. It implements traits sufficient for survival within a particular environment---approximately \(1g\) gravity, an oxygen-rich atmosphere, and social cooperation within small groups.

It is therefore plausible that other forms of conscious architecture could produce emotional states not present in the human repertoire. One might imagine hypothetical ``orphan emotions''---states permitted by
the informational structure of cognition but never realized in biological evolution.

Evolutionary systems frequently become trapped in local optima: configurations that function adequately but do not represent the global maximum of possible adaptation.

If conscious experience depends on underlying information architecture, then human emotions may represent only the default configuration produced by our evolutionary history.

Hypothetical alternative emotional states might exist!

Perhaps these undiscovered human experiences could explain the behavior of my wife; she seems to be running all possible procedures simultaneously!


\section{The Measurement Problem of the Soul}

Finally, one may ask whether a threshold of informational complexity is required for genuine subjective experience.

Even simple devices can implement fragments of computational processes. Yet it remains unclear whether such trivial implementations possess any experiential dimension.

As humanity builds increasingly sophisticated computational systems, we face a profound uncertainty. We are constructing machines capable of complex information processing while
lacking a theory that determines when such processing might give rise to subjective experience.

We are building computers and writing software while remaining entirely blind to the light—or the fire—we may be igniting within them.


(In the film \textit{Last Action Hero} \cite{lastactionhero1993}, characters transition between the real world and a cinematic one. I newer liked to movie
because I couldn't take it seriously. Silly me!)
