\chapter{The Deep Mystery of Mathematics}

\section{The Universal Agreement}

Logic and mathematics are abstract by nature. What is common to the addition of two bananas and the addition of two apples is the expression $2+2$.
Yet $2+2$ is not something one can eat, touch, or weigh. It is not located anywhere in space. It has no mass, no electric charge.
It cannot be detected with any measurement device, not even LIGO or JWST!  What, then, is it?

The most striking feature of mathematics is its stubborn universality. 
Two sentient beings, separated by light-years of vacuum or centuries of history, will inevitably converge upon the same Prime Number Theorem.
No matter how far we look in space and time, all the physics appear to follow the same universal rules, without an exception.

This suggests that mathematics is not a mere cultural artifact like music or fashion, but a reflection of a fundamental structure.

The physicist Eugene Wigner famously called this the ``Unreasonable Effectiveness of Mathematics in the Natural Sciences.''
Math does not appear to be just a language we speak, but a landscape we explore.

\section{Doctrines of Mathematics}

In the history of philosophy, several primary ``doctrines'' have attempted to explain this phenomenon.

\subsection*{Platonism}

Platonists argue that mathematical entities (numbers, sets, functions) are real, abstract objects that exist independently of us.
They don't exist in space or time, but they have a permanent existence in a ``Platonic realm.''

The ``unreasonable effectiveness'' of mathematics in the physical sciences suggests that the universe is built on a mathematical blueprint.
If we ``made it up,'' why does it predict the behavior of subatomic particles so perfectly?

\subsection*{Formalism}

Formalists, like David Hilbert, argue that mathematics is a formal game played with marks on paper.
We agree on the results not because we've discovered some universal truth, but because we started with the same axioms (rules).
If we both start playing Chess with the same rules, we will both agree on what a checkmate looks like.
That doesn't mean Checkmate is a fundamental law of the universe; it’s just the logical conclusion of the rules we agreed upon.

\subsection*{Intuitionism}

This group argues that math is entirely a construction of the human mind. Mathematics appears universal only because the human brain evolved to process logic and patterns in a specific way.
We don't see exceptions to math in the universe because we use math as the filter to understand the universe.
If something didn't fit our mathematical logic, we might not even be able to perceive it.

In this view, the universe may be chaotic, and math is merely our brain's organization tool. Imagine our eyes could only see red, green, and blue.
We would conclude the universe is made of those colors. However, the colors are only
properties of us observers. Math could be a similar filter, causing us to ignore the parts of reality that do not fit into equations.


\subsection*{Structuralism / Type Theory}

Structuralists argue that mathematics is about patterns and relationships, not the intrinsic nature of objects.  
Type theory formalizes this perspective: every object has a \emph{type}, and operations are only valid if the types match.
In programming, this is analogous to type-safe code: you cannot add a string to an integer without explicit conversion.

Type theory can serve as a foundation for modern mathematics, and it bridges naturally to computing, proof assistants, and functional programming.



\section{The Predictive Power Problem}

One of the strongest arguments for a Platonic universal structure view is when math leads the way and reality follows.

Astronomers didn't find Neptune by looking through a telescope first. They noticed Uranus wasn't moving the way Newton’s math said it should.
Urbain Le Verrier did the math and concluded that there must be a planet at specific coordinates. He pointed a telescope there, and there it was.

Paul Dirac wrote down an equation for the electron in 1928. The math had two solutions (like how $\sqrt{4}$ can be $2$ or $-2$). One solution described the electron;
the other predicted a positive electron. Years later, the positron was discovered.

Einstein’s equations predicted that a sufficiently massive star could collapse into a point of no size. Even Einstein found the idea of ``infinite density'' physically absurd,
yet we now have photographs of event horizons from the Event Horizon Telescope (EHT).

\subsection*{Is it a Discovery or a Map?}

If one believes math is a fundamental structure, one is essentially saying the universe can be understood as a mathematical structure.
Max Tegmark, a physicist at MIT, takes this view to the extreme with the \emph{Mathematical Universe Hypothesis}. It says our physical world is not just described by mathematics—it \emph{is} mathematics.




\section{Axiomatic Systems}

Mathematics is an \emph{axiomatic system}. An axiomatic system consists of a set of \emph{axioms}—statements assumed to be true—and rules of logical
inference that generate further statements, called \emph{theorems}.

Classical mathematics, from Euclid's geometry to modern set theory, operates within such frameworks. One starts from a set of axioms, and mathematics is then that follows.


\subsection*{Unification of Mathematics}

Just like different physical theories have been unified, modern mathematics has undergone consolidation.
In the early 20th century, it appeared that all disparate branches of math—from the curves of geometry to the
probabilities of statistics—could be expressed in the language of \textbf{Set Theory}.
By defining a ``number'' or a ``point'' as a specific arrangement of sets, mathematicians created a universal assembly language.

As of today, Zermelo-Fraenkel Set Theory (ZFC) can be referred to as the standard ``assembly language'' of mathematics.
One can define a number as a set, a function as a set of ordered pairs, and a geometric shape as a set of points.

However, a growing rival is \textbf{Category Theory}. Where Set Theory focuses on the ``insides'' of objects (what elements are in a set), 
Category Theory focuses on the ``relationships'' (how structures transform into one another).

Many argue that Category Theory is a more natural ``unified language'' for mathematics because it handles compositional structures more elegantly.

\subsection*{Programming Analogy}

There’s a surprisingly clean analogy between programming paradigms and the set-theoretic vs. category-theoretic perspectives. 

All software consists of two ingredients: data and code. Depending on which perspective you emphasize, you get different paradigms:

\begin{verbatim}
# Functional style (focus on code / transformations)
shift_left(integer)
\end{verbatim}

\begin{verbatim}
# Object-oriented style (focus on data / objects)
integer.shift_left()
\end{verbatim}

Set theory is analogous to the object-oriented / data-centric perspective: everything is built from elements and sets of elements.  
Category theory is analogous to the functional / code-centric perspective: it focuses on how structures relate and transform, abstracting over composition.

\subsection*{Category Theory Equation Example}

In category theory, objects and morphisms (arrows) are central. A simple example:

\[
f: X \to Y, \quad g: Y \to Z \quad \Rightarrow \quad g \circ f: X \to Z
\]

- $X, Y, Z$ are objects (sets, spaces, types, etc.).  
- $f, g$ are morphisms (functions, transformations).  
- $g \circ f$ is the composed morphism, showing how the functional perspective emphasizes \emph{relationships and composition} rather than internal content.

\subsection*{Mathematical Pluralism}

Modern scholarship has shifted toward \textbf{Mathematical Pluralism} \cite{friend2014pluralism}.
Proponents like Hamkins argue for a \textbf{set-theoretic multiverse} \cite{hamkins2012multiverse}, suggesting that instead of one absolute mathematical reality,
there exist diverse distinct concepts of sets, each instantiated in its own valid universe \cite{priest2024mathematical}.

Pluralism allows multiple internally consistent mathematical worlds to coexist. Only one appears realized in the observable universe, but multiple frameworks (set theory, category theory, type theory, etc.) can be valid within their own rules.

In computing, these ideas are practical: when building a programming language or automated theorem prover, one must choose which logic the system follows.



\section{The Universal Machine}

David Hilbert dreamed of a ``mechanical'' way to prove every truth. The work of Alan Turing suggested that math could be entirely mechanized.
Under this view, a mathematical truth is simply the output of a specific program.

Turing and Gödel proved there are limits: Turing showed some things a machine cannot calculate, even with infinite time.
For example, \textit{Chaitin's Constant} $\Omega$ is well-defined but uncomputable. Gödel proved that in any sufficiently powerful axiomatic system,
there are true statements that the system cannot prove.



\section{The Foundational Crisis}

All mathematical doctrines attempt to resolve the same unsettling observation: the universal consensus of mathematical truth. Whether in the mind of a human or the circuits of a probe, the internal logic of mathematics remains invariant.

There is no ambiguity as to whether data-oriented and code-oriented programming paradigms serve as two complementary perspectives on the same substrate—the Turing Machine. 

However, this analogy seems to break down when comparing set theory and category theory, the two primary paradigms of mathematics. Several puzzles remain:

\textbf{The Continuum Hypothesis:} Even if Set Theory (ZFC) is considered the ``assembly language'' of mathematics, there are basic questions it cannot answer. For example: Is there an infinity between the size of the integers and the size of the real numbers? ZFC's answer is effectively: ``I don't know.''

\textbf{Natural vs. Artificial Axioms:} Formalism suggests we simply ``make up rules.'' But why do these rules describe the physical world so effectively? This is the core of Wigner's ``unreasonable effectiveness.''

\textbf{Logical Pluralism:} Some argue there is not one mathematics, but many. Altering classical rules (such as the Law of Excluded Middle) yields a different mathematical universe. Yet, our physical universe appears to privilege classical logic.



\section{Can Science Unify Anything?}

There appears to be repeating pattern: the deeper we look, the more slippery the target becomes.

One might ask, can science actually unify anything?

It seems there is one solid conclusion one can draw: we do not inhabit a universe of Logical Pluralism.

What we observe in the physical world is not the full plurality of possible mathematical structures,
but a highly constrained subset.

At macroscopic scales, this subset behaves as if governed by classical logic:
statements appear either true or false, and contradictions are never observed.

However, at the fundamental level described by quantum mechanics, this classical picture is no longer exact.
The underlying structure allows superposition and non-classical correlations, suggesting that
classical truth may itself be an emergent phenomenon.

If a statement is false within a given theoretical framework, we do not expect to observe it.
However, the relationship between mathematical truth and physical realization is not one-to-one:
many mathematically consistent structures are never instantiated in nature.


\section{Non-Mathematics}

Mathematical pluralism does not undermine basic arithmetic.
Statements like $2+2=4$ remain invariant across essentially all reasonable frameworks.
The plurality arises only at deeper structural levels, where different axiomatic choices lead to genuinely distinct, but internally consistent, mathematical worlds.

However, one might ask whether entirely different kinds of universes could exist—ones that do not admit a description in terms of ordinary arithmetic.
Could there be universes where having two apples in both hands wouldn't mean having four apples in total?

If we speak of ``two apples in each hand,'' we are already assuming that objects are discrete, identifiable, and conserved when combined.
Under these assumptions, $2+2=4$ is not a convention but a structural necessity.

A universe in which this fails would not simply violate arithmetic; rather, it would violate the assumptions that make counting possible in the first place.
Objects might not have stable identity, might merge or split, or might not be countable at all.

In such a setting, expressions like $1+1+1+1$ need not correspond to $(1+1)+(1+1)$, because the underlying operation of ``combining'' may not be associative or even well-defined in terms of number.

A related situation arises in quantum mechanics.
The state of a system is described by a wavefunction in a Hilbert space, where states can be superposed.
Here, the fundamental operation is not counting discrete objects, but combining amplitudes.
As a result, the behavior of the system is governed by interference, not simple additive counting.

This does not violate arithmetic; rather, it reflects that the underlying structure is not one of discrete, countable entities.
The classical notion of ``how many'' emerges only in the limit where well-defined, non-overlapping objects can be identified.

Thus, the question is not whether $2+2$ could fail to equal $4$, but whether the world always admits a description in terms of discrete, countable entities.
Arithmetic is one mathematical structure among many, and it applies precisely when the structure of reality supports it.


\section{Assumptions and the Platonic Implications}

We might attempt to construct a theory of everything by assuming nothing.
But such an attempt fails at the very first step: by defining the task, we already make an assumption.
Without assumptions, there is no structure, no statements to evaluate, and no theory to construct.

If mathematics were something humans merely invented, could there be at least one person out of billions who could develop a ``non-mathematics''—a system that assumes nothing? The answer is no.
Even if everyone agreed to try, they could not succeed, because the very act of thinking or describing requires structure.

This observation doesn’t prove Platonism outright, but it strongly constrains alternatives:

\begin{itemize}
    \item \textbf{Intuitionism}, which holds that mathematics exists only in the mind, would need to explain why every human mind is bound by the same structural constraints, even if it ``made up'' math.
    \item \textbf{Formalism}, which treats mathematics as a game played from arbitrary axioms, would need to explain why the axioms themselves appear deeply tied to logic and rational thought.
\end{itemize}

In short, the impossibility of ``assuming nothing'' suggests that mathematics is not purely invented; it reflects necessary structures that any rational mind—human or otherwise—must encounter. 


\paragraph{Conclusion:}

Across all major doctrines—Platonism, Formalism, Intuitionism, Structuralism, and even Pluralism—one finds a common insight: mathematics is not arbitrary.
Its universality, predictive power, and internal consistency emerge from underlying structures that are inescapable for any rational observer, including average software
professionals suffering from wild imagination syndrome. 
