\chapter{The Big Questions and a few smaller Ones}

\section*{Do Parallel Universes Really Exist?}

Your brain has a finite number of neurons, which can describe only a finite amount of information.
This information content can be quantified by Shannon entropy as a number of bits, $n$.
These bits can be arranged in $2^n$ ways.
Some of these bits describe pure data and others describe code, allowing for a vast number of different compressed configurations.
One of these configurations describes the universe in which we live.
If we reject all supernatural and metaphysical explanations, that information is all there is.
Furthermore, there is nothing that renders any one particular configuration "more real" than the others—except for the fact that we find ourselves within one: the most probable one.


\section*{Can Computer Create and Feel Pain?}

A computer or even a simple thermostat, given sufficient time, could in theory duplicate the informational state of a suffering human.
However, this does not render the hardware itself conscious.
The virtual human with its consciousness and pain exists only within its virtual universe.
There is no residual mystery in the physical operation of the machine; it is simply a computer or a thermostat processing logic.

Running these simulations does not create suffering.
Rather, these processes merely reveal structures that already exists.

\section*{Are Quantum Computers so fast because They Run in Parallel Universes?}

Some theorists, like David Deutsch \cite{deutsch1997fabric}, argue that quantum computers achieve their speed by performing calculations across many parallel universes simultaneously.
However, under the lens of this theory, quantum computers do not "create" new answers through labor; they reveal information that already inherent exist. 

Much like the simulation of a human doesn't create new suffering but reveals a pre-existing configuration, a quantum computer uses interference to navigate a landscape of existing possibilities, revealing one of them into a single, observable result in our universe.


\section*{Does every bit change create a new version of consciousness?}

If the universe is a manifestation of information, we should expect to find ourselves within the most compressible configurations.
Predictability and the smoothness of physical laws exist because predictable and smooth data compresses well.
Structured data dominates the probability measure over chaotic, incompressible noise.

Based on personal observational evidence, what we experience as logic, thinking or consciousness is effectively a high-level structure.
It is compressed outcome of astronomical number of bits firing in the brain.
By organizing this vast data into a macroscopic structure—the brain discards the noise of individual bit-fluctuations to focus on the survival-critical patterns. Thus, consciousness might be a triumph of data efficiency: the process of resolving an astronomical number of microscopic states into a single, actionable, macroscopic identity.


\section*{Is this a "One-Parameter" Theory rather than Zero?}

Strictly speaking, \(\lambda\) appears as a parameter in the probability measure: 

\[
\mathbb{P}(\gamma \mid O) \propto \exp[-\lambda\, \mathcal{C}_O(\gamma)]
\]

However, \(\lambda\) is not an arbitrary input like the mass of a Higgs boson.
It represents the Global Compression Pressure—the degree to which the informational substrate is "packed" into predictable patterns.

\paragraph{The Analogy of Surface Tension}
Think of \(\lambda\) like the surface tension of water.
One doesn't "need" to input the tension to have a drop of water; the tension is an emergent property of how the molecules interact to find their lowest-energy state. 

In this theory:
\begin{itemize}
    \item \(\lambda\) is the Selection Pressure: It determines how "strictly" the universe favors simplicity over chaos.
    \item If \(\lambda \to 0\), the universe is white noise (infinite complexity).
    \item If \(\lambda \to \infty\), the universe is a single, static bit (maximal simplicity).
\end{itemize}



\section*{Solution to Divergences in GR and QFT?}

The primary "failure" of modern physics is the Singularity (infinite density/curvature).
In this theory, singularities are exponentially inprobable.

As curvature \(R\) increases, the geometric complexity \(C_G\) spikes.
Simultaneously, the spectral resources required to encode a point of infinite density would require an infinite frequency \(k\), making \(C_Q \to \infty\).
\[
\lim_{R \to \infty} \mathbb{P}(\gamma \mid O) \propto e^{-\infty} = 0
\]
The spectral complexity acts as a natural low-pass filter.
We don't find ourselves in the universe containing infinities because infinities don't compress well.


\section*{Where did all the matter in the universe come from?}

This question implicitly assumes temporal ordering and an external origin.
Both time and causality are emergent, observer-dependent phenomena (as established in Paper~1) \cite{meskanen2008}.
In the underlying informational substrate, the universe is abstract, static and timeless.


\section*{Why something rather than nothing?}

This is analogous to asking why ``heads'' rather than ``tails'': both are equally abstract and neither is privileged.


\section*{Why do particles follow a complex-valued wavefunction?}

Because we are observing compressed structure.
The wavefunction is the \textbf{compression codec} of the underlying informational structure of reality.
Observed wave-like behavior emerges from this compression.

Formally:
\[
\boxed{\text{Maximal Predictability}} \;\rightarrow\; \boxed{\text{Maximal Compression}} \;\rightarrow\; \boxed{\text{Maximal Probability}} 
\]

Maximal probability is exactly what we call the laws of physics.

Configurations with minimal observer-indexed complexity \( \mathcal{C}_O[\gamma] \) dominate the probability measure:
\[
\mathbb{P}(\gamma \mid O) \propto \exp[-\lambda \, \mathcal{C}_O[\gamma]].
\]
Hence, the ``laws'' of physics are statistical consequences of overwhelmingly probable, compressible configurations.


\section*{Why does the universe expand? Is there a link to entropy?}

Spacetime expansion is the geometric manifestation of increasing configurational entropy. A bitstring of zero entropy corresponds to a singularity—a geometric mapping with zero volume. As the entropy of the configuration increases, the corresponding geometric representation must necessarily 'stretch' to accommodate the growing density of internal constraints.


\section*{Why did the universe start with zero entropy?}

Low-entropy configurations are maximally compressible and thus most probable.
High-entropy starting states would require additional information to encode the observer, increasing \( \mathcal{C}_O[\gamma] \) and suppressing probability:
\[
\mathbb{P}(\gamma_{\rm high-entropy} \mid O) \ll \mathbb{P}(\gamma_{\rm zero-entropy} \mid O).
\]

Consequently, the arrow of time and increasing entropy emerge naturally:
\[
\text{Zero entropy} \;\Rightarrow\; \text{Maximal compressibility} \;\Rightarrow\; \text{Observer-compatible universe}.
\]

\textit{Speculative remark:} Other observer-compatible configurations may exist with different geometric histories.
Whether all must exhibit apparent low-entropy origins remains an open question.


\section*{Is the Universe Infinite?}

Based on all observational evidence, we observers are finite structures.
However, if reality is fundamentally informational, a natural question arises: is the total information finite or infinite?

If it were finite, one would have to ask: finite by how many bits? And who, or what, set that limit? Any fixed bound would itself require explanation.
A finite informational universe merely pushes the mystery back one level. 

The only non-arbitrary conclusion is that the deep nature of everything cannot be finitely bounded.
At the most fundamental level, reality must be infinite—perhaps not even informational in the familiar sense, but beyond any finite description.


\section*{What is spacetime made of?}

Observers find themselves within the most compressible (probable) configurations, and the path through these configurations is what observers sense as time and space.
Each configuration is slightly different, which observers perceive as gravity (geometric spacetime).

We ``fall'' because there are more copies of us ``down there.'' Gravity is the statistical bias toward the most probable (compressible) configurations describing the observer.


\section*{If there are no interaction or signals, how does the observer ``see'' anything?}

Observer is the universe. In static universe no bits flip down here because other bits flipped out there.
The universe perceived by an observer is already encoded within their wavefunction:

\[
\Psi_\gamma = \Psi_{\rm self} \otimes \Psi_{\rm env},
\]
where \(\Psi_{\rm self}\) defines the persistent observer structure and \(\Psi_{\rm env}\) encodes the surrounding environment. 

\textbf{Temporal and spatial extrapolation:} These are purely indexical—observer-compatible histories are extended by minimal-complexity continuations:
\[
\Psi_{\rm self}(t+1) \sim \arg\min_{\gamma_{0:t+1}} \mathcal{C}_O[\gamma_{0:t+1}], \quad \Psi_{\rm env} \sim \text{compressible extensions of } \Psi_{\rm self}.
\]

Apparent laws, structures, and predictability emerge statistically from correlations within the observer-compatible configuration space.


\section*{How does Gödel incompleteness affect the theory?}

Many highly respected authors have commented on this topic.

Gödel’s theorems constrain formal systems capable of arithmetic, but this framework targets ontological closure rather than formal deductive completeness.
Embedded observers will encounter Gödel-style limits in reasoning, which manifest as constraints on compression and representation, not on the underlying informational substrate itself.


\section*{Is a singularity a point of infinite curvature?}

A singularity is not a point of infinite physical magnitude, but a point of minimal informational rank. Because a zero-entropy state admits no internal degrees of freedom, the geometric mapping possesses no basis for spatial extension, resulting in a point-like representation. Furthermore, the traditional 'physical' singularity—conceived as a chaotic state of infinite density—is statistically suppressed; its lack of spectral regularity would result in a representation cost so high that its weight in the induced measure effectively vanishes. We do not see 'infinite curvature' because infinities compress poorly.


\section*{Are we living in a simulation?}

And who then would simulate the simulator? We are living in a information \cite{meskanen2008}. 
All observer-compatible configurations exist within a single informational substrate that is abstract by nature.
Simulated and non-simulated realities are ontologically equivalent. Hence, the question of ``living in a simulation'' is ill-posed;
everything arises from the same abstract information.


\section*{Is the recursive simulation hypothesis supported?}

Observers may discover how to simulate themselves as informational structures. In theory, recursive patterns compress well.
Such sub-simulations would then correspond to configurations within the same underlying informational substrate, rather than requiring a separate ontological level.

However, fully faithful and unbounded recursive simulations are informationally expensive. Within this framework, where probability is tied to compressibility,
such configurations are strongly suppressed. Highly complex, deeply nested simulations are less compressible and therefore less likely for an observer to inhabit.



\section*{Does the theory support QBism?}

QBism interprets the wavefunction as representing an agent’s subjective degrees of belief. This theory does not adopt a belief-based interpretation.
Instead, it treats probabilities as arising from objective structure, grounded in compressibility and statistical typicality.


\section*{Why do we age and die?}

The observer’s future is limited by the spectral complexity of their wavefunction, \(\psi_O\). Let \(\Sigma[\gamma]\) denote the number of independent frequency-phase components required to encode future history \(\gamma\).
The probability of persistence scales as:

\[
\mathbb{P}(\gamma \mid O) \propto \exp(-\alpha \, \Sigma[\gamma]).
\]

The accumulation of memories and internal structure increases \(\Sigma[\gamma]\), reducing the measure of compatible continuations. Aging and death are the geometric and statistical manifestations of combinatorial exhaustion.


\section*{Is this just Solipsism?}

This theory takes the observer’s existence as axiomatic, but it does not imply solipsism. The number of observers is a probabilistic variable.
Multiple observers dominate the measure. It would be very hard to argue that even from two different configurations; Alice with blond hair, and Alice with brunette hair, only the other one would be allowed. One would have to show that a single observer is statistically favored.



\section*{What motivates the replacement of Kolmogorov with Spectral Complexity?}

Observational evidence. We observe that the micro-cosmos behaves as waves. Furthermore, Kolmogorov complexity is non-computable and discontinuous.
Small input changes can yield massive complexity spikes, which is difficult to reconcile with smooth physical observations. 


\section*{Does the theory explain why the expansion rate is near-critical?}

Conditional on increasing entropy and typical emergence, observers are most likely to find themselves near the peak of the microstructure count (maximum compressibility).
This automatically corresponds to a near-critical expansion regime.


\section*{Is this AIT with Kolmogorov replaced by Spectral Complexity?}

The theory shares the principle of Solomonoff induction like concept: simpler descriptions occur more frequently in the ensemble of possibilities.
Uncomputable and non-continuous Kolmogorov complexity is replaced with smooth and continuous Spactral complexity.

\section*{Why are Gravitational Waves Real-valued}

They are the dithered state that the observer is currently indexing. There is no "complex-valued" gravity because gravity is the Geometric Interpretation—the final,
flattened, real-valued result of the spectral complexity.


\section{Do we have Free Will?}

Could software ever allow free will? Deterministic code—branching if-else statements, fixed CPU instructions—appears to preclude choice. The computer just runs the software; it cannot ``choose otherwise,'' regardless of what it might feel like to be an Alice-functional executing that program.

However, as we have demonstrated, consciousness arises from static informational configurations, not from dynamical execution.

Consider a simple line of code:
\begin{verbatim}
if (a < 3) {
    A
} else {
    B
}
\end{verbatim}

Deterministic? Yes. Sequential? Yes.
But what does \texttt{3} actually mean? In two’s complement, it might be interpreted as \texttt{3} or \texttt{-1}.
As a floating-point number, it could be something entirely different. 

Moreover, the code itself is nothing but data.
Programmers and CPUs interpret it as ``instructions,'' just as bit-patterns can be signed or unsigned.
However, there is no external device outside the universe that determines the universal ``instruction set.''
All $2^n$ arrangements are equally real. 

As the number of bits $n$ increases, the number of consistent interpretations grows combinatorially.
Each interpretation defines a slightly different ``Alice'' inside the informational structure.
Some of these interpretations preserve consciousness; others do not.
Some may interpret the bit-patterns that appear to be comparisons as something else entirely—some choosing $A$, some choosing $B$.

Alice is not forced to flow through this code like a train on tracks.
All possible versions of Alice already exist, corresponding to every consistent interpretation of the bits.
``Execution'' does not create Alice; it merely reveals one perspective among many. 

Which version does Alice experience? Most likely, the one corresponding to the most compressible, smooth, and coherent description—the version that dominates the observer-conditioned measure:
\[
\mathbb{P}(\text{Alice}_i) \propto 2^{-\mathcal{C}[\text{Alice}_i]}.
\]

Determinism at the ``CPU level'' does not constrain subjective experience because there is no fundamental CPU.
There is only the measure. 

And free will? Alice navigates the space of all consistent interpretations, experiencing deliberation, anticipation, pain, and choice—not because the software code grants it, but because the most compressible configurations are those in which Alice finds herself having reasoned and made decisions. 



