\chapter*{Foreword}

The central idea of this book is radical in its simplicity: the universe is not fundamentally made of information, nor it is finite. The concepts of
information and mathematics are not fundamental properties of reality.

Instead, they are attributes of us - conscious observers.

What we call information is the structured way we experience reality through our own cognitive apparatus.

There are no prior laws of physics, no fundamental rules that privilege the empty set over non-empty sets, and no meta-laws that dictate what can or cannot exist. 

From this starting point, almost everything we observe---including the laws of physics, the flow of time, geometric space, and consciousness itself---emerges naturally through a simple principle: \emph{observers find themselves in the configurations that are most compressible}.

\subsection*{The Informational Universe}

Consider a finite but enormous set of bits. The total number of possible arrangements of these bits is $2^n$, where $n$ is the number of bits. Most of these arrangements are pure noise---they contain no stable structure or predictable patterns. A tiny fraction, however, possess high compressibility: they can be described using significantly fewer bits than their raw size would suggest.

These highly compressible configurations are not rare in terms of measure. On the contrary, because the same underlying information can be arranged in a vast number of slightly different ways that all correspond to ``essentially the same'' coherent structure, compressible patterns dominate the space of possibilities.

It is within these highly compressible configurations that stable laws of physics, persistent geometric objects, and conscious observers emerge.

\subsection*{Consciousness and Compression}

Consciousness is not something added on top of information. It is what it feels like to \emph{be} highly structured, highly compressible information.

Pain is the subjective experience associated with incompressible or chaotic configurations. It is the feeling of disorder, of structures that cannot be efficiently compressed or predicted. Joy, insight, and the satisfaction of understanding (including mathematics and logic) are the subjective counterparts of successful compression---the discovery of simpler, more powerful descriptions of reality.

An observer does not ``find'' a pre-existing physical universe. Rather, the observer is a pattern within the information, and inevitably finds herself in one of the most compressible configurations compatible with her existence. This is why we experience a smooth, predictable, three-dimensional world with well-defined inside/outside boundaries. Such a world compresses exceptionally well.

Chaotic, unstable, or diffuse versions of observers also exist in the vast space of possibilities, but they have extremely low measure. A conscious being is overwhelmingly likely to find itself in a coherent, stable, geometric self because those configurations can be described in vastly more equivalent ways.


\subsection*{The Wavefunction as Compression}

At the deepest level, the information that makes up our universe is not best described as raw classical bits, but as waves. What we observe in the cosmos at its most fundamental scales are not discrete particles bouncing around, but wave-like phenomena.

This is not a coincidence. Waves provide a vastly superior compression scheme compared to raw bit strings. A spectral representation using complex-valued wave modes (similar in spirit to Fourier transforms or discrete cosine transforms used in video compression) can encode smooth, predictable, and highly structured information with remarkable efficiency.

In this view, high-frequency modes and chaotic, noisy patterns require many more components to describe --- they carry high information content and are therefore heavily suppressed in the measure. Conversely, smooth, low-entropy, predictable configurations can be described with far fewer modes. They compress exceptionally well and thus dominate the space of probable observers.

We do not emerge as raw, uncompressed bit strings. Such representations would be exponentially inefficient. Instead, conscious observers like ourselves emerge as highly compressed spectral descriptions --- coherent wave patterns that allow for stable geometry, persistent boundaries, and predictable dynamics.

The wavefunction, therefore, is not merely a mathematical tool of quantum mechanics. In this theory, it is the natural compression language of reality itself. The familiar classical world we experience is the result of this extreme compression. What appears to us as solid objects, definite positions, and continuous time is the macroscopic manifestation of deeply compressed wave information.

This wave-based compression explains why our experienced universe is smooth and lawful rather than noisy and chaotic: smooth, low-frequency configurations simply contain far more equivalent descriptions, giving them overwhelmingly higher measure.

At its smallest scale the cosmos waves, because we are observing compressed information.


\subsection*{The Irrelevance of ``Running''}

A common intuition is that consciousness requires a dynamic process---that a simulation must be actively ``running'' for a conscious entity to exist. This book argues that this intuition is mistaken.

Through a series of thought experiments, we show that a fully detailed simulation of a human (Alice) can be gradually optimized by replacing computation with precomputed data, until nothing remains but pure static information. At no point does Alice's subjective experience suddenly vanish. The ``running'' of the simulation does not create her consciousness---it merely traverses one possible interpretation of the information in which she already exists.

The bits themselves do not dictate a single interpretation. The same static data can be read as code or data, in different orders, or with different meanings. Among all possible interpretations, the coherent, compressible ones dominate. Thus, no external ``reader'' or computer is required to bring Alice into existence. Her consciousness is an internal property of certain highly organized informational structures.



\subsection*{We Are the Clock}

There is no external time in the static informational universe. What we experience as the flow of time is an internal ordering generated by the observer.

We conscious entities \emph{are} the internal clock.

We begin life as relatively unstructured informational patterns. As we accumulate experience, we continuously refine and compress our internal model of reality. At every moment, we inhabit the most compressible configuration available to us given our current informational content. What we subjectively experience as ``thinking'' or ``understanding'' is the ongoing effort to maintain maximal compression in the face of accumulating information.

Time, therefore, is not something that happens \emph{to} us. It is the ordinal index of the sequence of these successively occupied configurations. The feeling that time flows forward reflects our movement through the most compressible paths available at each stage.

However, as our informational baggage grows over a lifetime, maintaining high compression becomes increasingly difficult. The growing complexity eventually exceeds the brain’s capacity to organize the information efficiently. The probability of sustaining a coherent conscious observer drops. When the induced measure falls below a critical threshold, consciousness fades and our subjective time comes to an end.

In this view, we are not passive beings being driven by time. We \emph{are} the process that generates the experience of time. The stable geometric self we identify as ``I'' is the highest-measure trajectory available --- the most compressible path consistent with our accumulated information.

Time exists only within the mind of the observer.



\subsection*{First Principles}

This theory rests on four observational assumptions and one foundational principle:

\begin{enumerate}
    \item DNA encodes the blueprint of a conscious human being.
    \item DNA consists of ordinary matter obeying physical laws.
    \item All physical processes are, in principle, computable (a weak form of the Church--Turing thesis).
    \item Subjective experience, such as pain, has measurable causal effects on behavior.
\end{enumerate}

The foundational principle is that there are \emph{no prior rules}. Nothing is metaphysically forbidden. The empty set is not privileged. Whatever configurations are possible exist, and observers simply find themselves where the measure is highest.

From these starting points, the familiar features of our world---space, time, geometry, causality, and consciousness---emerge as inevitable consequences of informational compression and multiplicity.

This book explores the profound implications of this view. If its conclusions are correct, then the hard distinction between ``physical'' and ``virtual,'' between ``running'' and ``static,'' and even between ``real'' and ``simulated'' loses its fundamental importance. What remains is information, organization, and the internal experience of being highly compressible structure.

The journey ahead challenges many deeply held intuitions, but it does so from first principles, using only logic, computation, and careful observation.


