\chapter{Free Will}

\section{Code with Free Will}
  
If DNA is the blueprint of life and the operation of DNA can be
described with axiomatic systems, then everything in humans must be
axiomatic by nature. Therefore, according to the Church-Turing thesis,
we humans can be simulated in a computer.
  
Computers in general are deterministic systems that do not necessarily exhibit
truly random behavior. Even the random value generator in computers is based
on deterministic logic, generating values that only appear to be random. 
  
It appears evident that such a deterministic system cannot possess
free will as we understand it. Turing machines running the DNA
simulation are bound to follow their logic without the ability to
make alternative choices. 
  
This implies that the individuals within the simulated universe
cannot possess free will either. They are constrained to behave
according to the programming of the Turing Machine. 


Consider the following piece of code:

\begin{verbatim}
if (a < 3) {
   ...
} else {
   ...
}
\end{verbatim}

How much free will does the code give to Alice? Apparently, zero.

Regardless of how much pain Alice feels, or how deterministic she is in her reasoning,
the computer running the code does not care about Alice's feelings. It is bound
to run the code without any ability to choose otherwise. Alice's feelings cannot
change the hard-coded threshold `3`, or the comparison logic built into the CPU, any
more than we can make $2+2$ equal anything other than $4$.

\section{Random Indeterministic Systems}

Nothing prevents us from implementing a truly random generator. According to Quantum Mechanics,
there is genuine randomness inherent in nature, which we can
potentially leverage to create a truly indeterministic random generator. Could such a computer then grant simulated individuals free will? 
  
Let us consider a person who is contemplating whether to turn left
or not. Suddenly, a completely unexpected and indeterministic
event takes place, such as a high-energy particle burst emitted by
the sun. A few particles manage to pass through certain critical
brain cells, impacting the person's decision-making process in a genuinely unpredictable manner. As a result, instead of turning
left, the person decides to move forward. 

Did such a random, unpredictable event introduce something like free will?  
  
The person had no control over the distant particle burst, nor the way it affected
their decision-making. The person would therefore have no more free will with
this random event than without it!

The only effect of the random event was that the person made
a decision that had nothing to do with the current situation they
were facing. The decision they made was based on irrelevant data.


Instead of establishing free will, the person just made 
a decision that was not based on their own reasoning.
  
In the above thought experiment, this random behavior might not
matter too much. One might think that such a loss of predictability due to a random event is harmless and simply adds more variety to life.  
 
But let's consider a scenario where a person's life is at stake. 
In such a life-threatening situation, the person's
ability to make decisions based on relevant data becomes crucial. If an unexpected random event disrupts their logical
decision-making process, free will becomes the least of their
concerns. We certainly wouldn't want to define free will as
something that only applies in trivial matters that do not
really matter. Our existence relies on our ability to make valid decisions 
guided by relevant data. 


\section{The Non-Computable Alternative}

If we accept that both determinism and randomness fail to provide a substrate for free will, we are left with a vacuum.
This is where Roger Penrose's proposal becomes relevant. Penrose argues that consciousness—and by extension, the "choice" involved in it—is \textit{non-computable}. 

He suggests that the brain does not merely run a DNA-based algorithm, but taps into a level of physics that is neither deterministic in the classical
sense nor random in the quantum sense. According to the Orchestrated Objective Reduction (Orch-OR) theory, the collapse of quantum states in
neuronal microtubules is a physical event that follows a logic beyond the reach of any Turing Machine.

\section{Cognitive Architecture as Code}

We often view programming languages as alien, technical constructs, but they are actually mirrors of our own cognitive architecture. When we write a conditional statement, we are simply formalizing the way we navigate reality. For example: \textit{If my car is low on gasoline, and I am too busy to fill it up, then call a cab.}

\begin{verbatim}
if (self.not_enough_gasoline && self.not_enough_time) {
    action = self.call_a_cab();
}
\end{verbatim}

Our "reasoning" is a 1-1 analogy to the operations of a computer. Programming languages did not emerge as an arbitrary invention; they were developed to reflect the way we think and make decisions. 

However, this leads us back to the deterministic wall. If our decision-making is a mirror of this code, then the "choice" to call a cab is an illusion.
The outcome was pre-ordained the moment the variables \texttt{not\_enough\_gasoline} and \texttt{not\_enough\_time} were set to \texttt{true}.
Alice did not "decide" anything; she simply reached the only logical exit point of the algorithm.


\section{Consciousness in Static Timeless Universe}

If our four axioms hold, the universe isn't a single deterministic movie playing forward in external time.
It's the complete static library of all $2^n$ possible informational configurations, existing timelessly — as we proved in Paper I by gradually turning code into pure data.

In this picture, free will isn't about magically overriding $2 + 2 = 4$ or breaking mathematics.
Alice has full freedom to interpret the same underlying data in astronomically many different ways as executable code.
Correspondingly, in the full informational ontology there exist vast numbers of slightly different Alices, each following her own impeccable internal logic.

Does that mean there are billions of distinct conscious Alices, each with her own private inner life?
Or is consciousness something coarser — a property that emerges across large equivalence classes of compatible patterns, where tiny microscopic differences don't create new separate minds? 

Either way, the feeling of freedom can emerge naturally: from inside the pattern, Alice only ever experiences one configurationt a time — never the entire static ensemble at once.
Your conscious self gets to "live" the particular logical path that feels like a genuine decision. That's not an illusion; that's simply what rich, navigable information feels like when you're embedded in it.


\section{Conclusion}

We use language to describe our 'choices,' but language itself is an axiomatic system.
Whether we speak in English or C++, we are merely describing the rails upon which our thoughts are forced to run.

If our decision is not affected by indeterministic events, then
our decision-making is based on pure deterministic logic. Otherwise, our decision is based on phenomena we have no control over. 
  
Neither of the two cases demonstrates free will as we usually understand it.
  
The only logical conclusion is that allowing random events to
influence our decision-making does not introduce anything resembling free will, but rather hampers our logical
decision-making process. Instead of working out the decision based
on our own logic, we effectively let the coin decide.
  
We do not possess the freedom to choose when to commit a sin and when to compensate by
praying. Every conscious decision we make is based on logical
reasoning, where $2+2$ always results in $4$.

As with computer software, we are bound to follow our logic without the ability to choose otherwise.

