\chapter{Humans as Axiomatic System}

\section{The Penrose Problem}

According to Roger Penrose \cite{penrose1989emperor}, the operation of the human brain might not be an axiomatic system. 
Gödel's incompleteness theorem states that within any sufficiently powerful axiomatic system, there exist statements that are \emph{true} but cannot be proven within that system. 

Since computers and Turing Machines are equivalent to formal axiomatic systems, this implies that there are well-defined mathematical truths that no algorithm can ever compute. 
In other words, some problems are fundamentally uncomputable by any machine, regardless of its speed or memory.

Penrose speculates that the human brain can, at least in principle, apprehend such truths. 
If true, this suggests that human reasoning is not confined to algorithmic computation: the mind may access insights that no Turing machine can produce.  

What Penrose effectively argues is that mathematics is more expansive than what a Turing machine is capable of describing. Even if the machine had an infinitely long tape and infinite time to run, it could not implement all of mathematics. If human consciousness resides on that 'far side' of math, several consequences follow from this hypothesis:

\begin{enumerate}
    \item \textbf{Limits of Simulation:} If the brain is non-algorithmic, then no digital computer, no matter how advanced, can fully simulate human consciousness or understanding. 
    Even infinite computational resources would be insufficient to replicate every aspect of human thought.
    
    \item \textbf{Beyond Code + Data:} Unlike a program, which is entirely specified by its code and the data it operates on, the human mind may operate in a domain that transcends these structures. 
    Emotions, intuition, insight, and conscious awareness might not be reducible to sequences of bits or instructions.
    
    \item \textbf{Implications for Artificial Intelligence:} AI systems, however sophisticated, are fundamentally Turing machines. 
    If Penrose is correct, they can never truly replicate the full range of human understanding or consciousness; they will always be limited to algorithmically computable tasks.
    
    \item \textbf{Gödelian Insight in Humans:} The ability to see the truth of statements that are unprovable in a given formal system suggests that the human mind may operate by principles that go beyond formal axiomatic rules.
    Consciousness might involve non-computable processes, potentially of a physical nature not yet understood — perhaps even quantum mechanical, as Penrose has speculated.
    
    \item \textbf{Philosophical Implications:} If humans are not fully formal systems, then certain aspects of subjective experience, creativity, and insight are inherently inaccessible to purely mechanistic explanations. 
    This strengthens the idea that some phenomena — including consciousness, free will, or the subjective feeling of pain — may lie outside the reach of computational description.
\end{enumerate}

Taken together, these points indicate a profound divide between what is algorithmically computable and what the human mind can achieve. 
Even if mathematics itself can be formalized into axiomatic systems, the processes of human understanding may transcend those systems entirely. 


\section{Basic Assumptions}

Let's start with the following assumptions:

\begin{enumerate}
    \item DNA encodes the blueprint of life.
    \item DNA consists of ordinary matter that obeys physical laws.
\end{enumerate}

It should be noted that these are observable assumptions, and as such, they cannot be definitively proven. However, the observational evidence supporting them is strong.

The first assumption posits that the human genome encodes all the information required to construct a conscious, pain-sensitive human being within this universe and its observed laws of physics.

The second assumption states that DNA is composed solely of ordinary physical matter. It is made of the same matter as everything else and is governed by the same physical laws,
with no non-physical or supernatural influences affecting its operation.

It then follows that humans, unlike what Roger Penrose speculates, must be implementations of axiomatic systems. Consequently, consciousness can be described using the principles of mathematics.


\section{Church-Turing Thesis}

Let us make a third assumption:

\begin{enumerate}
    \setcounter{enumi}{2}
    \item Church--Turing Thesis holds.
\end{enumerate}

It states that all physical processes can, in principle, be simulated by a device known as a Turing machine. Modern computers are essentially Turing machines.
The Church--Turing Thesis  has never been formally proven, but if it were false, we would have good reason to worry about keeping our money in bank accounts \aside{what money}?.

From these three assumptions it follows that humans can be simulated by a Turing machine---or, in its modern incarnation, a computer. 


\section{DNA Simulation Thought Experiments}

Suppose we digitize a human genome and run it on a computer simulating a universe governed by the
same laws of physics as our own. As the simulation executes, the DNA evolves into a conscious,
pain-sensitive observer. The simulated human experiences an expanding universe where time flows from
past to future, and tooth pain is real.


\subsection*{Optimizing Code}

All software programs consist of two kinds of information: code ($c$) and data ($d$). In a DNA simulation, the code would describe the laws of physics, such as quantum mechanics and gravity.
The data would include the digitized DNA and a sufficiently large section of the surrounding universe. Let us assume the simulation software
consists of rougly equal amount of code and data.

\[
1.0 = \frac{\operatorname{sizeof}(\text{cccccccccccccccccccccccc})}
         {\operatorname{sizeof}(\text{dddddddddddddddddddddddd})}
\]


A well-known technique for optimizing slow, CPU-intensive code is to use lookup tables, which replace computation with precomputed data. For example, one can replace all \texttt{sqrt()} computations:

\begin{verbatim}
result = sqrt(arg);
\end{verbatim}

with precomputed values:

\begin{verbatim}
result = sqrt_lookuptable[arg];
\end{verbatim}

Empirically, software programs yield identical results regardless of how the result was computed. $2+2=4$, and it does not matter if we replace all $2+2$ equations in our code with precomputed value of $4$. This optimization therefore cannot affect the simulation's output, nor the observer's experience of time or pain. Delaying or accelerating computation affects only the external runtime, not the internal state transitions of the simulated system.

However, this optimization has the effect of reducing the amount of code and increasing the amount of data in our DNA simulation:

\[
0.9 = \frac{\operatorname{sizeof}(\text{cccccccccccccccccccccccc})}
         {\operatorname{sizeof}(\text{ddddddddddddddddddddddddddd})}
\]


Now, imagine we gradually optimize the DNA simulation by replacing algorithmic components with lookup tables.
As a consequence, the number of CPU cycles required to run the simulation decreases. Suppose we take this optimization to the extreme:
all computation is replaced by a static dataset encoding the entire execution trace.

\[
0.0 = r = \frac{\operatorname{sizeof}(\text{})}
         {\operatorname{sizeof}(\text{dddddddddddddddddddddddd}}
\]


As a result, we don't have anything to run in a computer. It is just a massive hard drive containing all procedures precomputed.

Does the simulated human still experience time and pain? 

The answer, within the axiomatic model, is yes. Affirming otherwise would certainly imply the existence of a new physical constant in our books of physics: a minimum code-data ratio required for consciousness to emerge.

Temporal structure and pain, therefore, must emerge from the internal relationships among states, not from the external runtime. From the internal perspective of the simulated observer, time still flows from past to future and pain is real.

As a conclusion, a static dataset can fully specify a universe containing conscious observers with subjective time. Consequently, time and pain must be properties of simulated observers, not fundamental properties of the universe.


\subsection*{Multi-threaded Simulation}

When a single DNA simulation---let's call her Alice---runs on a computer, the execution trace is easy to study. Every CPU instruction drives the computer (and Alice) to a new state. From Alice's perspective, time flows forward, and the effect of each CPU cycle can be mapped to a simulated particle in her world.

\[
| \text{Alice}|\text{Alice}|\text{Alice}|\text{Alice}|\dots|
\]

However, consider a system running multiple DNA simulations concurrently---say, Alice and Bob---where thread scheduling is governed by quantum randomness. The resulting execution trace interleaves their simulated lives in segments of unpredictable length.

\[
|\text{AliceAliceAli}|\text{BobBobB}|\text{AliceA}|\text{BobBo}|\text{Alic}|\text{BobB}\dots|
\]

Since both single- and multi-threaded computers are computationally equivalent, each observer must experience a coherent, continuous timeline.

Now, let's gradually shorten the number of CPU cycles until each thread is limited to running a single CPU cycle before switching. Let's also add more DNA simulations, like Robert, John, and Jill. As the number of concurrent simulations increases, the execution trace becomes increasingly fragmented. Additionally, modern multi-threaded systems often include many extra threads for operating system tasks, such as listening for network requests. In the limit of infinitely many perfectly interleaved simulations, the execution trace approaches pure white noise.

\[
|\text{A}| \text{B}| \text{OS}| \text{R}| \text{J}| \text{OS}| \text{l}| \text{Ji}| \text{i}| \text{ce}| \text{b}| \text{ob}| \text{n}| \text{l}| \dots|
\]

Is Alice still conscious?

The answer, again within the axiomatic model, is yes. Empirically, multi-threaded computers function reliably regardless of how few CPU cycles are allocated per thread switch or how narrow the CPU’s internal
registers are. From each observer's internal perspective, time still flows from past to future.

This raises an obvious question: how does Alice know which sequences belong to her and which do not, in order to remain conscious?

Conclusion: if there is any way to interpret static data as a "conscious observer," then that is exactly what happens: a conscious observer emerges.
If the initial assumptions hold, consciousness, pain, and subjective time can emerge from static data that resembles pure static noise.


\section{Causality}

It we can create a simulation with a computers, then it might seem obvious that those virtual universes only comes into existence when a simulation is actively executed—that the
computer must be powered on to run the simulation code for the simulated world to exist.
So if the simulation computer is never started, then no simulated virtual world emerges. No consciousness, definitely no pain. 

However, pure static data (such as the full execution trace of a simulation) has no notion of time. One cannot measure the time it takes from the ``execution'' to create the virtual world.
One cannot argue that one created the other. What appears as static information, e.g., execution trace of a computer to us external observers, appears as an expanding universe and
pain to the simulated human observing the data from inside. This relationship is representational, not causal. 
The computer and the simulated universe are two sides of the same coin: distinct arrangements of the same information. 


\section{Parallel Universes}

And there is more than just two sides on the coin. Consider an execution trace of $N$ bits. These bits can be arranged in $2^N$ ways.
Apparently, most of them describe chaotic universes with no conscious observers. One, however, describes our identical simulated twins---Alice, Bob, and others.
And one describes something we call a computer, which is simulating a computationally heavy procedure---DNA.


\section{Philosophical Zombies}

Philosopher David Chalmers \cite{chalmers1995facing} proposed the concept of a philosophical zombie: a hypothetical being that is physically and behaviorally identical to a
conscious human but lacks any subjective experience. Such a zombie would respond to pain stimuli in the exact same way a conscious person does---it would cry out,
flinch, and try to avoid the source of pain---but it would not feel anything.

According to Chalmers, even with a perfect simulation, we would only be observing the physical processes. We still wouldn't know if there's a "ghost in the machine"---a feeling of what it's
like to be that simulated being. A computer could be programmed to perfectly mimic the behavior of a person feeling pain without actually having the experience itself. 


\section{Pain Hypothesis --- From Philosophy to Physics}

Let's make a fourth assumption:

\begin{enumerate}
    \setcounter{enumi}{3}
    \item Pain has measurable effects.
\end{enumerate}

This assumption brings the concept of pain from philosophy to physics. Just like gravity, pain is assumed to have observable consequences that are physically detectable and measurable. Formally:

\[
\text{Human} \neq \text{Human + Pain}
\]

If a system's behavior is entirely determined by its physical components and their interactions, an axiomatic copy should exhibit identical behavior.
If their behavior is identical, then their internal states, including consciousness and pain, must also be identical. $2+2$ holds for both bananas and apples. 
If the original's behavior is driven by the experience of pain, the simulation must have that experience too. Otherwise there would be a contradiction.

Correspondingly, the P-Zombie is an impossibility. If Axioms 1--4 hold, the P-Zombie premise collapses:
\begin{itemize}
    \item Axioms 2 and 3 state the system is an axiomatic, computable entity.
    \item Axiom 4 states that the experience of pain ($P$) has a measurable, physical effect.
\end{itemize}


\subsection*{Falsification of the Hypothesis}

While full-scale DNA simulations are currently beyond our computational reach, this limitation is secondary to the theoretical framework. Much like our inability to calculate the highest possible prime numbers,
the current technical ceiling does not invalidate the underlying logic; what matters is that such simulations are possible in principle.

As soon as computational technology allows for high-fidelity biological modeling, we will be able to test whether a simulated human truly experiences subjective states, such as pain.
This provides a clear path for empirical validation. We can simulate a statistically significant sample of human subjects. By comparing the simulated subjects' responses to those of their
real-world counterparts, we can measure the difference. If the simulated subjects exhibit the necessary biological reactions but lack the qualitative experience (qualia) of pain,
the arguments presented here collapse. This would indicate that at least one of the four initial axioms is invalid.


\section{Conclusions}

If the four assumptions hold, then consciousness, time, and even pain can emerge from the structure of information alone.
The universe itself is made of abstract information. Observers, particles, and the flow of time are just patterns woven into
this tapestry, emerging wherever conditions allow. 

Time appears to be a subjective property of us intelligent and conscious observers, illusion that we created in our minds rather than a fundamental property of the universe.
The universe is informational and abstract by nature, and our everything in us observers can emerge even from static or noisy informational substrate.

Where did all the matter in the universe come from? The answer appears to be nowhere. There is no matter more than there is numbers, multiplications, or square roots. The nature of everything is fundamentally abstract and virtual by nature.
