\chapter{Life}

\section{Definition of Life}

Based on a naive classification, all entities that appear capable of experiencing pain share one common attribute—they are alive. Mechanical systems such as copper wires, relays, and home automation systems, to name just a few, are not.

However, life proves difficult to define.

The first obvious problem is those borderline cases. Entities such as viruses, prions, and sterile organisms
satisfy some criteria for life but not others. Any strict definition tends to exclude things many
scientists consider ``life'' or include things they do not.

Recent developments in computers and AI raise another problem, namely substrate independence.
We can build digital replications of living cells in computers. Can those digital replicas be
considered ``life'' as well?

Observer bias might also be a concern. Our definitions are based on life on Earth, so it is unclear whether
they would capture fundamentally different forms of life elsewhere.


\section{Evolution}

There is a science fiction book, \emph{The Black Cloud} \cite{hoyle1957blackcloud},
in which an intelligent cloud arrives and proceeds to cause all sorts
of trouble. The reason the book is science fiction rather than science,
and why intelligent clouds cannot exist in real life, is that there are
no known laws of physics on which such a cloud could plausibly be based.

The more intelligent and complex a system is, the smaller the
probability that it could simply appear spontaneously. For a giant
intelligent cloud, the probability should be practically zero.

The only theory we currently have that explains the existence of
truly intelligent systems is evolution. For evolution to work there
must be many candidates and a mechanism of natural selection capable
of eliminating the less successful ones. In the case of intelligent
clouds, there would have to be lots of clouds, and natural selection
would need to eliminate the weak and disorganized ones while favoring
those capable of maintaining structure. Only then could intelligent
clouds gradually develop.

While Charles Darwin's theory of evolution has been subject to heavy
religion-saturated debate in the past, believing in it is no longer a
question of faith. Modern science has revealed the genetic code in DNA
and confirmed the theory for all its essential parts. 

But how would evolution work for a cloud whose behavior is governed
by the Navier--Stokes equations of fluid dynamics? How could such a
cloud keep its information in order? What would happen if a storm
passed by and blew the cloud apart? Even relatively small disturbances to a human brain can cause severe
damage. A violent disturbance would correspond to putting a brain into
a kitchen blender and switching it on. It is not difficult to see why
the brain would not think very clearly afterward.
An intelligent cloud would need to know its boundaries and maintain
a stable internal structure to stay 'alive'. A gas cloud has neither. It would not
remain organized long enough to evolve intelligence.

In principle one could imagine some unknown form of matter with
properties suitable for maintaining information in a gaseous state.
However, astronomical observations give us little support for this
idea. When we analyze the spectrum of light coming from anywhere in
the universe—even from the most distant galaxies—we see essentially
the same electromagnetic fingerprints. This tells us that they are
made of the same raw materials as our own home sweet Milky Way.



\section{Alien Life}

It has been proposed that life could be far more common than generally believed and might not necessarily be carbon-based.
However, the challenge is that Earth appears to be an ideal habitat for diverse life forms.
If life emerged easily, shouldn't we observe "exotic" organisms—perhaps non-DNA-based—coexisting with us? Instead, we find only one biochemical lineage, all rooted in DNA and evolution.
Furthermore, our exploration of the solar system and our extensive monitoring of radio signals (SETI) have yielded no traces of life, intelligent or otherwise.

Perhaps life exists on a different temporal scale. If their biological processes run significantly faster or slower than ours, we might perceive them as either stationary matter or a blur too rapid to recognize as sentient.
However, while time-scale differences are a great sci-fi concept, biology is bound by the laws of thermodynamics and chemistry.
Chemical reactions (the basis of life) happen at specific rates dictated by temperature and molecular stability.
A creature "moving too fast" would likely burn up from the heat of its own metabolism; one "moving too slow" might not be able to gather enough energy to maintain its structure against entropy.

Maybe exotic life could be here, but we simply aren't looking for it correctly.
Most of our tools (PCR, DNA sequencing) are designed specifically to find DNA.
If a "non-DNA" microbe existed in the dirt, our current tests would likely dismiss it as "non-living" chemical noise.


\section{Super-Symmetric Life}

The vast majority of matter in the universe consists of "dark matter," the nature of which remains one of science's greatest mysteries.
String theory predicts a set of supersymmetric particles that could account for this phenomenon.
If these particles are capable of forming complex structures—analogous to atoms and molecules—then it is statistically plausible (given that dark matter is five times more prevalent than visible matter) that "dark life" exists.
We may be sharing the universe with an entire dark ecology that remains completely invisible to our senses.

Again, physics fights back. Dark matter (and most predicted supersymmetric particles) can be shown to be collisionless.
It doesn't interact with electromagnetism, which means it doesn't "clump" the way normal matter does.

To have life, one needs complex molecules. Normal matter forms molecules because electrons attract and repel each other.
Since dark matter doesn't seem to interact via the electromagnetic force, it can't form "dark atoms" or "dark DNA."
It mostly just passes through itself and us like a ghost. Without a way to bond particles together, one can't build a "dark person."



\section{Spritual Life}
  
If the Bible and other religious sources are to be believed, then
intelligent life can exist in spiritual form, such as in the form
of soul. But how trustworty are the stories put forward by religious texts?

Do they make any sense at all? 

For example, according to the Bible, God created mankind in His own
image; in the image of God, He created them - male and female He
created them. 

As we humans are undoubtedly a social species living in dense groups, this implies that
God must also have a social nature, or he wouldn't have created us in his image.
But in that case, where did other gods go? 

Applying the theory of evolution to God raises teh question of how such a magnificent system could have
emerged in the first place? The probability of the
spontaneous existence of an omnipotent being is even smaller than
the spontaneous existence of a smart cloud based on Navier-Stokes
equations. However, if there was only one God, how did he develop social properties that are emergent
only in species living in groups? 
  
All intelligence that we are aware of on earth is a result of
long process of evolution. It would therefore be  easier to accept 
the existence of highly  developed omnipotent creatures if they
were based on  the theory of natural selection.
However, in order for evolution to work, there had to be lots of
Gods and there had to be an environment hostile enough to  drive 
their evolution. 
  
According to Bible there is only one God,  so maybe something went wrong and 
all the other Gods died or moved some place else. One stayed  
and got lonely. Being a social animal God was in a bad need 
of good company, and created man. 
  
  
This would actually explain the friendliness of God. 
If there was only one God then why would he appear to have the  properties specific to
social animals? He could do whatever he feels like, without
worrying about anybody  else but himself. If only one God got
created spontaneously, then it would  be quite difficult to
explain how he got not only so incredible  omnipotent but also so
incredible friendly. Maybe Gods forgot their  social origins and
turned individualistic destroying each others using black holes
as bombs.  
  
  
By sending his son down to earth God wants to tell us
humans that we should  not be selfis but social to make sure we
humans don't end up destroying  each others the same way Gods did. 
  
Another possible explanation is that God simply exists and was
newer created.  God could simply be. According to Stephen Hawking the 
universe could just be, and hence it didn't need any creator. 
If a well recognized scientist says the entire universe 
could simply be, then surely could God simply be too! Both constructions 
are somewhat unbelievable structures to exist. Furthermore, omnipotent
God could surely create an universe appearing closed to us mortals.

Stephen Hawking also commented quite unfavourable on God 
by asking whether God can create a stone too heavy for him to lift.
This should be quite a trivial problem to model, so let's implement
this as a software.

Lifting is a process where one applies a force  to a rock.   
  
If the force is greater than the force caused by the gravity field of
earth, then one would be strong enough to lift the rock. 
  
How strong a force can an allmighty creature then apply?  God is
supposed to be almighty, so the only sensible answer is infinitely
strong. Otherwise, God wouldn't be  almighty at all, because one
could name a rock God wouldn't be able to  lift.
  
And for the same reason God would also have to be able to create  infinitely heavy
rocks. So the question whether God can create a stone he can't lift
is  the same question as if an infinite force can lift an infinite
mass.

Mathematics cannot really answer this question. This leaves us with three possible resolutions.
Either there is no such thing as omnipotency, the question itself doesn't make any sense, or we
just don't fully undererstand infinities.
  
One thing we do understand is that Newton's theory of  gravity is not representative
theory for rocks getting infinitely big. So one should perhaps use
General Relativity rather than  Newton's theory of physics. 
  
If Stephen Hawking is concerned, universe might not need
a creator. The uinverse could simply be. Then
why not could God be too?
  
One can try to evaluate which one of the two explanations is more probable. The more complex and intelligent the structure the smaller the changes
that it can spontaneously exist. This is a trivial application of statistical mechanics.


\textbf{The Proportion for God — Information-Theoretic Formulation}

Let \(K(X)\) denote the algorithmic complexity (Kolmogorov complexity) of a system \(X\), i.e., the length of the shortest description that fully specifies it. Then the probability of spontaneous existence of a low-entropy system can be modeled as

\[
P(X) \sim e^{-K(X)}.
\]

Applying this to God and the Universe gives

\[
\frac{P(\text{God})}{P(\text{Universe})} \sim 
\frac{e^{-K(\text{God})}}{e^{-K(\text{Universe})}} 
= e^{-[K(\text{God}) - K(\text{Universe})]}.
\]

Since God is assumed to be more sophisticated and informationally richer than the Universe itself, we have

\[
K(\text{God}) > K(\text{Universe}) \quad \Rightarrow \quad P(\text{God}) \ll P(\text{Universe}).
\]

This formalizes the intuitive idea that **highly ordered, complex systems are exponentially unlikely to appear spontaneously**, and a system more “complex” than the Universe (e.g., God) is correspondingly even more improbable to arise without creation.

If God really is capable to create the universe, then  the structure of God apparently most likely be more evolved and
intelligent one of the two. Correspondingly, the probability for spontaneous existence of God should be smaller compared to the probability of
spontaneous existence of the universe.

Across different periods and societies, humans have tended to ascribe divine qualities to a range of gods.
Why do we believe in less probable explanation? 


\section*{Can Science Unify Anything?}

Having wandered through the tangled forest of definitions for life, one begins to notice a familiar pattern: the deeper we look, the more slippery the target becomes.
Life, like the ultimate laws of physics, resists a neat, single-box definition. Cells, ecosystems, viruses, and AI each stretch the concept in different directions.
Every attempt at a universal definition seems to encounter a counterexample, a loophole, or a subtle exception.  

Physicists know this feeling well. The quest for a Theory of Everything—the holy grail that unites quantum mechanics and general relativity—is still ongoing.
No single, definitive “toe” exists to tie together all of reality. Perhaps science, in its elegant ambition, is more like a playful child trying to herd cats:
the cats are clever, they scatter, and sometimes they seem to stare back with a mischievous grin, asking if we really understand anything at all.  

So can science ever truly unify anything? Are we any closer finding answer to the hard problem of consciousness? 

Not really. However, certain characteristics appear repeatedly across all known living systems.

First, due to second law of thermodynamics complex systems tend naturally toward disorder rather than order. By favoring structures
that better preserve and replicate information, evolutionary processes gradually accumulate complexity and functionality. Our DNA plays
an essential role here.

Second, even a tiny amount of intrusion is typically lethal to living objects.
For example, if a human toe is exposed to a sufficiently strong acid, the consequences are usually catastrophic.
The atoms themselves remain, but the organized system they formed ceases to exist.
Life depends not only on internal complexity but also on the ability to maintain a stable boundary that protects its structure from environmental processes that tend to dissolve order into disorder.

Living systems, as we observe them, exhibit distinct well defined boundaries and maintain a relatively rigid structural organization.

