\chapter{Deep Nature of Quantum Mechanics}


\section{Introduction}

Now that we are, at least in spirit, fully trained practitioners of quantum mechanics, let us ask: what is the deep nature of the theory and its mysterious wavefunction? 

Why does everything at the subatomic level appear to behave like a wave?

If we view the system from a distance, we observe two things:
First, the microscopic building blocks are simple—a small set of fields with just a few parameters producing enormous emergent richness.
Second, the complex amplitudes in the wavefunction do all the heavy lifting.
Interference, entanglement, tunneling and superpositions are simply consequences of existing within a complex-valued Hilbert space. 

If we don’t pay attention to the above-mentioned little details, what remains?



\subsection*{The Movie Analogy}

In a conventional MPEG-compressed movie, each frame is composed of pixels. If we were observers made of pixels living inside such a movie, what would we perceive as the "laws of nature"?

We would observe that our constituent pixels follow mysterious, abstract, yet deterministic wave-like patterns. These patterns are the result of the Discrete Cosine Transform (DCT) or Fourier transforms—the underlying mathematical tools that MPEG codecs use to compress raw pixels into a maximally dense format for efficient storage and transmission.

From this perspective, the Quantum Mechanical wavefunction is simply an advanced, complex-valued version of the sinusoidal building blocks used in JPEG or MPEG images.

Just as a Fourier transform spreads image information across coefficients, the wavefunction distributes amplitude across configurations. 


\section{The Dithering Analogy}

The Born rule acts as a form of probabilistic dithering.

Much like how a printed newspaper smooths out limited resolution—using discrete dots to mask jagged edges and simulate gradients—nature samples amplitudes probabilistically. This allows the universe to reproduce a continuous distribution while utilizing the fewest possible discrete resources.





\section{Spectral Complexity}

Kolmogorov complexity, as utilized in Algorithmic Information Theory (AIT), provides a theoretical foundation for minimal description length. However, it is neither computable nor continuous, which limits its utility as a functional physical principle. 

Observational evidence suggests that physical systems already possess a natural, inherent mode of representation: their decomposition into \textbf{spectral modes}. This motivates the replacement of abstract algorithmic complexity with a physically grounded alternative:

\textbf{Spectral Complexity ($C_s$):} The total information required to specify the amplitudes, frequencies, and phases of the modes composing a physical state.


\subsection*{The Cost Function}

If we implement a wavefunction as a computational class, the complexity is determined by the bits required to instantiate its constituent modes.
We can model the total spectral complexity $C_s$ of a state $\Psi$ as the sum of its algorithmic overhead and the parameters of its $N$ modes:

\[
C_s(\Psi) = \text{Cost}(\text{base}) + \sum_{i=1}^{N} \left[ \text{Cost}(\phi_i) + \text{Cost}(A_i) + \log_2\left(\frac{\omega_i}{\Delta\omega}\right) \right]
\]

Where:
\begin{itemize}
\item \textbf{$\text{Cost}(\text{base})$}: Represents the constant $O(1)$ cost of the `sin()` method. In a universe-scale description, this is vanishingly small and treated as a constant.
\item \textbf{$\text{Cost}(\phi_i)$}: The bit-cost of the phase. Since phase is bounded within $[0, 2\pi]$, this is a small, fixed-width cost per mode.
\item \textbf{$\text{Cost}(A_i)$}: The information required to encode the amplitude.
\item \textbf{$\log_2(\omega_i / \Delta\omega)$}: The dominant term. It represents the number of bits required to specify the frequency $\omega$ relative to the system's minimum resolution $\Delta\omega$.
\end{itemize}

\subsection*{Solomonoff Suppression}

One might argue that a continuous universe implies an infinite number of modes, leading to infinite complexity.
However, within the framework of AIT, these infinities are naturally suppressed. According to the \textbf{Solomonoff Prior},
the probability $P$ of a state is exponentially inversely proportional to its description length $L$:

\[
P(s) \approx 2^{-L(s)}
\]


High-frequency "noise" and infinite-mode configurations are so spectrally expensive that their probability of manifestation vanishes. We do not find ourselves in a chaotic, infinitely complex universe for the same reason we do not find ourselves inside an uncompressed, raw-pixel movie: the simplest, most compressed descriptions are the ones that dominate reality. 


\section{Conclusions}

Quantum mechanics may be viewed fundamentally as a theory of representation. The wavefunction encodes physical reality in a continuous, spectral form that naturally supports superposition, interference, and entanglement.

By replacing the uncomputable Kolmogorov notion of complexity with spectral complexity, we establish a bridge between physical structure and informational description. Under this framework, the remaining components of Algorithmic Information Theory (AIT) still apply. Specifically, Solomonoff induction explains why we find ourselves within a law-like, smooth universe governed by symmetries, rather than manifesting as chaotic Boltzmann brains.

In this view, the wave-like nature of the microscopic world is not a mystery, but a reflection of how information is most effectively organized.

The cosmos waves because we are observing compressed structure.

The wavefunction is the compression scheme of the universe.


