\chapter{Religions}

\section{Soul as the source Consciousness}

Is there something like ``soul'' expaining why physics cannot find consciousness?
Should we turn to God? Should we replace our physics books with the Bible?

Modern science avoids the word ``soul'' because it cannot be tested, but philosophers, neuroscientists, and even some physicists continue to explore whether consciousness requires something beyond ordinary physical processes. In effect, some modern theories echo older soul-like concepts, even if they do not use the word.

John Eccles argued that consciousness cannot be explained by material processes alone; he proposed a non-physical ``self'' (similar to a soul) interacting with the brain.

Penrose \& Hameroff (Orch-OR theory) suggests that consciousness may arise from quantum processes in neurons and speculated that this might connect with the idea of a soul,
though their theory remains widely disputed.

David Chalmers formulation of the ``hard problem of consciousness'' does not invoke a soul explicitly, but it reopens the possibility by arguing that
subjective experience appears irreducible to brain activity.

Near-death experience researchers (e.g., Pim van Lommel) argue that NDEs indicate consciousness can exist independently of the brain, again resonating with
traditional soul concepts.


\subsection*{Near-Death Experiences (NDEs)}

This is a phenomenon that itself is undeniable. Millions of people across cultures and history have reported consistent experiences.
Serious institutions have investigated them for decades. The scientific mystery lies in how a brain with measurable zero electrical activity  (flat EEG) can
produce the most vivid, structured, and life-changing memories a person has ever had. 

While every account is unique, researchers have identified several core elements common to these episodes:

\begin{itemize}
\item \textbf{Out-of-Body Experiences (OBEs):} The sensation of floating above one’s own body and observing medical procedures or the immediate surroundings.
\item \textbf{The ``Tunnel'':} A feeling of being drawn through a dark space toward a brilliant, warm light.
\item \textbf{Life Review:} A panoramic replay of one’s life events, often experienced from the perspective of others involved.
\item \textbf{Intense Peace:} An overwhelming sense of love, euphoria, and the total absence of pain.
\item \textbf{The ``Border'':} Reaching a point of no return where the individual is told (or decides) to return to their body.
\end{itemize}

But how reliable are these findings?

Lommel, a Dutch cardiologist, published a landmark study in \emph{The Lancet} (2001).
His work is significant because, rather than relying on retrospective anecdotes, he studied patients resuscitated in hospital settings in real-time.
He found that some patients reported clear, structured consciousness at a time when their brains showed no measurable electrical activity.

Van Lommel argues that the brain might function more as a \emph{receiver} for consciousness than its producer.
In this view, if the radio (the brain) is broken, the broadcast (consciousness) continues to exist; it simply can no longer be translated into the physical realm.

The first AWARE study document one case that shook the researchers. A 57-year-old social worker described his resuscitation in such detail that it seemed
to defy the timeline of a "dead" brain. He accurately described the people in the room, the sounds of the machines, and the specific actions of the doctors.
Most importantly, he recalled the auditory "beeps" of an automated external defibrillator (AED) that only occurred several minutes after his
heart had stopped and his brain should have been inactive.

If a consciousness can truly float above the physical body, then it should observe objects on high
shelves or in adjacent rooms that are hidden from the patient's physical line of sight. This has been tested in AWARE studies. 

The AWARE study is a masterclass in the difficulty of studying the fringe. It confirmed that 'dying' brains can sometimes flare with gamma waves—the kind of high-level electrical chatter usually reserved for deep thought. However, to date, no one has returned from the brink to report what was written on the cards placed near the ceiling.
The lack of 'hits' doesn't necessarily proof that the soul stays put; it might just show that when you’re undergoing the most profound experience of your existence, you aren't exactly looking for a shelf.

All those NDEs might be hallucinations. What is  remarkable is that, unlike typical hallucinations, NDEs almost always result in a
permanent personality shift and the removal of the fear of death.

What science has been shown is that consciousness is more resilient than once thought; the lights do not go out instantly.



\subsection*{New Testament and Christianity}

According to Jesus (Matthew 10:28):
``Do not fear those who kill the body but cannot kill the soul.''

The New Testament indirectly links many aspects of consciousness
to the concept of the soul. Our conscious decisions determine
what happens to our soul once we die. The soul is described as the
immaterial and eternal part of a human being that is distinct from
the physical body.

Indeed, soul, just like consciousness and pain, appear to live in a domain
that is not physical. It is something that cannot be touched, weighted,
or directly measured by any measuring device.

How reliable is the New Testament? How strong of a case do
science and archaeological findings make to support the stories
within it? Is there any evidence to support that a person named
Jesus ever lived?

It turns out there is no single piece of direct archaeological
evidence for Jesus whatsoever. We just have to believe the story
put forward in the Gospels.

The Gospels are, however, based on a large number of ancient
archaeological documents written in Greek. Thousands of manuscripts
have been found, and new ones are discovered every year.

None of these manuscripts are original, but merely copies of
copies. Due to the manual copying methods used back then (copy
machines were invented much later), none of the found manuscripts
is identical. They all contain errors and differences.
However, by comparing the numerous copies found, researchers have
managed to restore the original document.

From the four Gospels, three would seem to tell essentially the
same story. These are Matthew's, Mark's, and Luke's Gospels, and
they are known as the \texttt{Synoptic} Gospels. There are 661
verses in Mark's Gospel, from which 607 are also included in
Matthew's Gospel, and 360 are included in Luke's Gospel. Matthew
and Luke have 230 common verses which, however, are not included
in Mark. Because of this, it would seem that Matthew and Luke are
based on Mark, as well as on some yet unknown source that is
called the ``Q'' document. The name ``Q'' comes from the German
word ``Quelle,'' which means ``source.'' Q is thought to be a
collection of sayings and teachings of Jesus that were shared by
Matthew and Luke but not found in Mark.

The Gospel of Mark is believed to have been written between
60--70 AD. Science has managed to pinpoint the dating of the
manuscripts with astonishing accuracy by considering many
factors, for example handwriting style, paleographic analysis,
and historical references.

So the fact is that the New Testament draws its foundation from
thousands of ancient documents. Rejecting their authenticity is
akin to disputing the existence of dinosaurs despite the
continual discovery of new fossils each year. Moreover, early
dating provides the New Testament with a degree of credibility,
offering testimony that, while not necessarily from eyewitnesses,
still carries weight.

\subsection*{Doubt}

However, a couple of concerns arise in the mind of an average
programmer.

All the manuscripts seem to be based on just two original
sources: the Gospel document and the yet-to-be-found Q-document
(as per the ``Two-Source Hypothesis''). How can one be certain
that the original text was not written by someone suffering from
a wild imagination, at the very least? How can one determine that
the authors of the original texts did not exaggerate to some
degree? How can we discern whether these authors were simply
storytellers of their time?

Furthermore, Jews do not seem to believe in Jesus as the Messiah.
According to Jewish tradition several reasons exist. Jesus did
not fulfill the messianic prophecies nor embody the personal
qualifications of the Messiah. He did not build the Third Temple
nor gather all Jews back to the Land of Israel. He also failed to
spread knowledge of the God of Israel so that humanity would be
united as one. The God of Israel is definitely not king over all
the world.

Jesus' teachings and the doctrines associated with Christianity,
including the concept of the Trinity and the divinity of Jesus,
are also in serious contradiction with Jewish theological
beliefs. Judaism emphasizes monotheism and the unity of God,
rejecting the notion of Jesus as a divine figure.

What worries an average programmer is that the God worshipped by
Christians and Jews is the same God. Both religions trace their
roots back to the Hebrew Bible (Old Testament) and acknowledge
God as the one and only.

The fact that Jews do not believe in Jesus, the son of God, as
the Messiah therefore feels like a serious matter. It is the very
soul of an average programmer that is at stake. Jesus was a Jew,
and if Jews themselves do not believe in him, then why should an
average programmer do so?

Are we Christians certain that we are on the right path? Are we
confident that we are headed toward heaven instead of eternal
suffering in the fires of hell?

And what about other religions, many of which seem even more
incompatible with Christianity? Do they fall into the category
of false religions, with believers whose unfortunate fate one
can only regret?


\section{Religion and Logical Reasoning}

Does the idea of God hold up under logical scrutiny?

For example, according to the Bible, God is one and only. God created mankind in His own
image; in the image of God, He created them - male and female He
created them. 

As we humans are undoubtedly a social species living in dense groups, this implies that
God must also have a social nature, or he wouldn't have created us in his image.
But in that case, where did other gods go? 

If there was only one God, how did he develop social properties that are emergent
only in species living in groups?

How such a magnificent system exist anyway? The probability of the
spontaneous existence of an omnipotent being is even smaller than
the spontaneous existence of a smart cloud based on Navier-Stokes
equations. 
  
All intelligence that we are aware of on earth is a result of
long process of evolution. It would therefore be  easier to accept 
the existence of highly  developed omnipotent creatures if they
were based on  the theory of natural selection.
However, in order for evolution to work, there had to be lots of
Gods and there had to be an environment hostile enough to  drive 
their evolution. 
  
Maybe something went wrong and all the other Gods died or moved some place else. One stayed  
and got lonely. Being a social animal God was in a bad need 
of good company, and created man. 
 
This would actually explain the friendliness of God. 
If there was only one God then why would he appear to have the  properties specific to
social animals? He could do whatever he feels like, without
worrying about anybody  else but himself. If only one God got
created spontaneously, then it would  be quite difficult to
explain how he got not only so incredible  omnipotent but also so
incredible friendly. Maybe Gods forgot their  social origins and
turned individualistic destroying each others using black holes
as bombs.  
  
By sending his son down to earth God wants to tell us
humans that we should  not be selfish but social to make sure we
humans don't end up destroying  each other the same way Gods did. 


\subsection*{Omnipotent God}

Stephen Hawking has commented quite unfavourable on God 
by asking whether God can create a stone too heavy for him to lift.
This should be a problem an avarege programmer could try to solve.

Lifting is a process where one applies a force  to a rock under a gravity field.
  
If the force is greater than the force caused by the gravity field, then one would be strong enough to lift the rock. 
  
How strong a force can an allmighty creature then apply?  God is
supposed to be almighty, so the only sensible answer is infinitely
strong. Otherwise, God wouldn't be  almighty at all, because one
could name a rock God wouldn't be able to  lift.
  
And for the same reason God would also have to be able to create  infinitely heavy
rocks. So the question whether God can create a stone he can't lift
is  the same question as if an infinite force can lift an infinite
mass.

Mathematics cannot really answer this question. 
  

\subsection*{Universe that Just Be}

According to Stephen Hawking, the universe might not need a creator. He suggested that the universe could simply be—existing without ever having been created. If that is the case, why couldn't the same be true for God? Perhaps God simply exists and was never created.

If a well recognized scientist says the entire universe 
could simply be, then surely could God simply be too! Both constructions 
are somewhat unbelievable structures to exist.

One can try to evaluate which one of the two explanations is more probable.
The more complex and intelligent the structure the smaller the changes
that it can spontaneously exist. This is a trivial application of statistical mechanics.

Let \(K(X)\) denote the algorithmic complexity (Kolmogorov complexity) of a system \(X\), i.e.,
the length of the shortest description that fully specifies it. Then the probability of spontaneous
existence of a low-entropy system can be modeled as

\[
P(X) \sim e^{-K(X)}.
\]

Applying this to God and the Universe gives

\[
\frac{P(\text{God})}{P(\text{Universe})} \sim 
\frac{e^{-K(\text{God})}}{e^{-K(\text{Universe})}} 
= e^{-[K(\text{God}) - K(\text{Universe})]}.
\]

Since God is assumed to be more sophisticated and informationally richer than the Universe itself, we have

\[
K(\text{God}) > K(\text{Universe}) \quad \Rightarrow \quad P(\text{God}) \ll P(\text{Universe}).
\]

This formalizes the intuitive idea that highly ordered, complex systems are exponentially unlikely to appear spontaneously, and a system more complex than the Universe (e.g., God) is correspondingly even more improbable to arise.

If God is truly capable of creating the universe, then God’s own structure must be more complex and intelligent than the universe itself. Consequently, the probability of God existing spontaneously should be lower than the probability of the universe existing spontaneously.

So why do we believe the less probable explanation?

