\chapter{Science vs. Religion}

It is not difficult to find people who do not believe in science.
Physics would then be nothing more than another belief system—a kind of religion.
Instead of sacred texts, people would place their faith in scientific papers.


\section{Practicing Physics}

Physics is a field of science that is fundamentally concerned with the study of observable phenomena.
Science relies fundamentally on logical reasoning. Mathematics provides the framework for this reasoning, allowing precise formulation of hypotheses and rigorous derivation of predictions.
In essence, science is the systematic study of reality using the rules of rational thought—a discipline built on mathematics, the science of reasoning itself.

All fields of science, including physics, follow the so-called
scientific method. The method defines how science is practiced.

First, one makes observations about the phenomenon
to be studied. Then one develops hypotheses to explain the phenomenon. In
the case of physics, these are typically described in the language of
mathematics. The new theory is then tested against available
data. Each new observation that is consistent with the predictions
of the theory increases the credibility of the theory.

However, no amount of experimentation can ever prove a theory to be correct.
Regardless of how many experiments have confirmed it so far, nothing guarantees that the next experiment will do the same.
A physical theory is always subject to falsification: even a single contradictory observation can prove it wrong.
For this reason, extraordinary claims in physics require extraordinary statistical evidence.
In fields such as particle physics, discoveries are typically not accepted until they reach a significance of five standard deviations
corresponding to a probability of only about one in several million that the result is a statistical fluctuation.

A physical theory is always subject to falsification.

And indeed, the list of falsified theories is breathtaking. No matter how elegant or compelling a theory appears, nature may simply refuse to comply with its predictions. In many cases, a theory works extremely well within a limited domain—for example, Newtonian mechanics accurately describes everyday motion, even though it ultimately fails at very high speeds or in strong gravitational fields. This is why such theories remain highly usable: they provide excellent approximations within well-defined regimes.

Physical theories are therefore often accompanied by precision boundaries—ranges of conditions over which the theory has been experimentally verified. For instance, the energies, temperatures, or distances over which a law has been tested may be reported with uncertainties as small as several standard deviations (e.g., five sigma) to quantify the confidence in the predictions. Beyond these boundaries, the theory may no longer hold, and new physics could emerge.


\section{Practicing Religion}

While different religions have different practices, there are some
key elements that many of them share. These include prophecy,
prayer, rituals and ceremonies, and moral and ethical guidelines.

At the heart of all religions, however, are sacred texts and
faith. People read these texts, memorize them, and believe
them.

Mathematics plays no essential role in religions. Sacred texts are not
compared with observations and  experiments are not carried out to validate them. This is because
applying rational reasoning to religious texts can lead to
logical contradictions, and hence to doubt. Doubt is something
between believing and not believing. Such doubts are often
associated with Satan and his attempts to lead people away
from the truth.

For example, according to some interpretations of holy texts,
the Earth is only a few thousand years old. However, we can
observe dinosaur fossils. According to science, and based on
overwhelming observational evidence, even common sense, they must be much older.
What one observes therefore seems to be in direct contradiction
with what one believes.

These apparent contradictions can be resolved by assuming that
God is so great and so far beyond human understanding that no
human being will ever come close to comprehending His actions.
With our pitifully thin layer of grey brain matter, it may seem
foolish even to question the holy texts. God might simply have
placed those dinosaur fossils there to test one's faith.

One can also explain many apparent contradictions in sacred
texts by assuming that they are not meant to be taken literally.
Instead, one allows a certain degree of flexibility in their
interpretation.



\section{Conclusions}

By comparing the attributes of the two systems-science and religion-the only conclusion one can draw is that they are fundamentally incompatible.
Religious texts are not taken literally, whereas scientific papers are interpreted in the strictest sense.
Religions demand total, unconditional belief, and any doubt is often discouraged.
In science, the situation is the exact opposite: a theory is accepted as scientific only when it is supported by a substantial body of experimental evidence.

In fact, the most central concept in physics - consistency with observations - would be lethal to religions. 
If the claims of religions could be experimentally verified, then there would be no room for believing anymore.

If we saw God, we would start studying His properties and develop mathematical laws to model them.
Observations would turn religion into science.

