\chapter{A Zero-Parameter Theory of Everything}

\section{Everything is Information}

If we accept the foundational assumptions—universes are informational, observers are abstract patterns, and time is not fundamental—the universe can be understood as \textbf{static information}.
There is no ontological generation, no external clock, no metaphysical “creation.”

Consider \(n\) bits of information. These bits can be arranged in \(2^n\) distinct configurations. Each configuration encodes a possible universe.

\textbf{Time} emerges internally: it is the ordinal sequence of configurations experienced by an observer.
Observers traverse the configuration space according to the \textbf{order of experience}, not because the universe “runs” like a program.

% 1. Configuration Space & Time
\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=1.0, every node/.style={scale=0.9}]
% nodes
\node[circle, draw, fill=blue!20] (gamma1) at (0,0) {$\gamma_1$};
\node[circle, draw, fill=blue!20] (gamma2) at (2,1) {$\gamma_2$};
\node[circle, draw, fill=blue!20] (gamma3) at (4,0) {$\gamma_3$};
\node[circle, draw, fill=blue!20] (gamma4) at (6,1) {$\gamma_4$};
\node[circle, draw, fill=blue!20] (gamma5) at (8,0) {$\gamma_5$};
% arrows representing observer ordering
\draw[->, thick] (gamma1) -- (gamma2);
\draw[->, thick] (gamma2) -- (gamma3);
\draw[->, thick] (gamma3) -- (gamma4);
\draw[->, thick] (gamma4) -- (gamma5);
% label
\node[below=1cm of gamma3] {Observer traverses configuration space; time is internal ordering};
\end{tikzpicture}
\caption{Observer traversing configurations $\gamma_i$. Time is emergent, the ordinal sequence of experience.}
\end{figure}



\section{Defining the Observer}

An \textbf{observer} is a coherent informational pattern capable of sustaining continuity of experience. Formally:

\begin{itemize}
    \item Let \(\Gamma\) be the set of all configurations \(\gamma \in \Gamma\).
    \item Let \(\mathcal{O}(\gamma)\) be the number of observers in configuration \(\gamma\).
\end{itemize}

The probability that an observer finds themselves in a particular configuration is determined by the \textbf{joint measure of spectral, geometric, and observer-specific complexity}:

\[
\mathbb{P}(\gamma \mid O) = \frac{1}{Z_O} \exp\Big[-\lambda\, \mathcal{C}_O[\gamma]\Big], \quad \gamma \in \Gamma_O
\]

\section{Unifying Quantum Mechanics and Gravity}

At the deepest level, the universe is static and informational. There is no dynamical generation, no external time, and no need for a simulator or metaphysical agent. Everything that exists is a configuration in an enormous information space. For $n$ bits, there are $2^n$ possible configurations. Observers experience a “flow of time” simply by traversing a sequence of these configurations:

\begin{itemize}
\item \textbf{Configuration Space:} All possible states exist as static information.  
\item \textbf{Traversal:} Time is the ordinal sequence an observer finds themselves in.  
\end{itemize}

In such a universe, the central question is: \emph{Which configurations are observers most likely to inhabit?} The answer lies in compression: observers overwhelmingly find themselves in the \textbf{most compressible} structures. This gives rise to both quantum mechanics and general relativity as emergent phenomena.  

We quantify this using two complementary notions of complexity:

\begin{itemize}
\item \textbf{Spectral Complexity} $C_Q$ — compression of microscopic wave-like correlations.  
\item \textbf{Geometric Complexity} $C_G$ — compression of macroscopic relational structures in spacetime.
\end{itemize}



\section{Spectral Complexity: Why Particles Wave}

The wavefunction is nature’s spectral compression of microscopic correlations. We penalize high-frequency components to favor smooth, predictable states:

\begin{equation}
C_Q = \sum_i k_{\text{eff}}^2 |A_i|^2,
\end{equation}

where $k_{\text{eff}} = \min(|k|, N-|k|)$ in a discrete domain $\mathbb{Z}_N$. Sharp gradients and high frequencies are costly, so wave-like behavior emerges naturally as the optimal spectral representation.  

Particles appear wave-like because “waves compress correlations best.”



\section{Geometric Complexity: Why Spacetime Curves}

Observers also require coherent, well-defined boundaries, which are efficiently encoded in a geometric manifold. The geometric complexity functional is:

\begin{equation}
C_G = \alpha \int R \sqrt{-g} \, d^4x + \beta \sum |a + \Gamma(v,v)|^2,
\end{equation}

where:  

\begin{itemize}
\item $R$ is the Ricci scalar curvature — the “cost” of bending spacetime.  
\item $\Gamma(v,v)$ represents acceleration along a trajectory; non-geodesic motion increases the description length.  
\item $\alpha, \beta$ are weighting constants.  
\end{itemize}

Smooth, geodesic motion compresses histories most efficiently.



\section{Resolving Singularities with Spectral Compression}

In classical GR, curvature can become infinite at singularities. In an informational universe, such singularities are impossible: they would require infinite spectral description. The probability of a configuration is weighted by $e^{-C_Q}$, so states needing infinite spectral resources are naturally suppressed:

\[
\text{Singularities } \Rightarrow 0 \text{ probability}.
\]

The spectral “codec” acts as a low-pass filter, smoothing out infinite curvature.  

Nature’s “compression algorithm” prevents infinities from appearing.

with

\[
\mathcal{C}_O[\gamma] = C_Q[\gamma] + C_G[\gamma] + C_\text{obs}[\gamma; O].
\]

Here:

\begin{itemize}
\item \(C_Q[\gamma]\) = \textbf{spectral complexity} (microstructure compression, i.e., wave-like correlations).  
\item \(C_G[\gamma]\) = \textbf{geometric complexity} (macroscopic spacetime compression, i.e., curvature, continuity).  
\item \(C_\text{obs}[\gamma; O]\) = \textbf{observer continuity cost} (internal boundaries and consistency).  
\end{itemize}

\textbf{Key idea:} the most probable universes are those that \textbf{maximize the number of observers} while minimizing total complexity.
Observers don’t “choose” a universe—they simply exist in the highly probable configurations where patterns like them dominate.

% 2. Observer as informational pattern
\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=1.0, every node/.style={scale=0.9}]
% Configuration space box
\draw[thick] (0,0) rectangle (8,4);
\node at (4,4.3) {Configuration Space};
% Observer pattern
\draw[fill=orange!50, thick] (2,1) rectangle (6,3);
\node at (4,2) {Observer $O$};
% note
\node[below=0.3cm of {$(4,0)$}] {Observer is a coherent informational pattern embedded in many configurations};
\end{tikzpicture}
\caption{Observer embedded in configuration space. The most probable configurations preserve internal structure.}
\end{figure}

Observers experience reality according to the \emph{joint minimization} of spectral and geometric complexity. The probability of a particular history $\gamma$ for an observer $O$ is:

\begin{equation}
\mathbb{P}(\gamma \mid O) = \frac{1}{Z_O} \exp \Big[ -\lambda ( C_Q(\gamma) + C_G(\gamma) + C_{\text{int}}(\gamma,G) ) \Big],
\end{equation}

where $C_{\text{int}}$ accounts for the interaction between spectral and geometric compression. Physics emerges where spectral and geometric simplicity align, producing a coherent and smooth world.


\section{Emergent Physical Laws}

Treating total complexity as a generalized action \(S\), the observed dynamics correspond to stationary points \(\delta S = 0\):

\subsection*{Wave Equations}

Varying spectral complexity with respect to the wavefunction yields:

\begin{equation}
\nabla^2 \psi = 0,
\end{equation}

which, in the presence of a clock or sampling frequency, recovers the Schrödinger or Helmholtz form.  

\subsection*{Field Equations}

Varying geometric complexity with respect to the metric reproduces Einstein’s field equations:

\begin{equation}
R_{\mu\nu} - \frac{1}{2} g_{\mu\nu} R = 8\pi G T_{\mu\nu}.
\end{equation}

The laws of physics are \emph{stationary points of minimal information cost}.



\section{Predictions and Observational Consequences}

Because a single wavefunction can be embedded in multiple geometric configurations with nearly equal complexity, the observer sees apparent stochasticity or “noise” at the interface of scales. This produces:

\begin{itemize}
\item Fundamental variance or “quantization error” when compressing high-dimensional spectral information into spacetime.  
\item Potentially measurable effects in extreme events, e.g., black hole mergers or Hawking radiation.  
\end{itemize}



\section{Conclusion}

The universe is fundamentally informational. Observers inhabit the most compressible trajectories in configuration space. Quantum mechanics emerges from \textbf{spectral compression} of microscopic correlations. General relativity emerges from \textbf{geometric compression} of macroscopic relational structures.  



\section{Spectral Compression: Why the Microcosmos Waves}

At the microscopic scale, the universe is best represented in the \textbf{spectral basis}:

\begin{itemize}
    \item Smooth, correlated structures compress efficiently with few spectral components.
    \item Sharp discontinuities or high-frequency fluctuations require many components and are exponentially improbable.
\end{itemize}

This explains:

\begin{itemize}
    \item Quantum waves and interference
    \item Linear superposition
    \item Smooth, predictable microphysics
\end{itemize}

Formally, the \textbf{spectral complexity} \(C_Q\) is the minimal number of frequency components needed to encode an observer-compatible history. Histories with low \(C_Q\) dominate the observer-conditioned measure:

\[
C_Q = \sum_i k_\text{eff}^2 |A_i|^2, \quad k_\text{eff} = \min(|k_i|, N-|k_i|).
\]



% 3. Spectral compression (wave vs noise)
\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=1.0, every node/.style={scale=0.9}]
% axis
\draw[->] (0,0) -- (8,0) node[right] {Position $x$};
\draw[->] (0,0) -- (0,3) node[above] {Amplitude};
% smooth wave
\draw[thick, blue] plot[smooth, samples=100, domain=0:8] (\x,{1.0 + 0.8*sin(2*3.1415*\x/8)});
\node at (7,2.5) {Smooth, low $C_Q$, high probability};
% noisy wave
\draw[thick, red, dashed] plot[smooth, samples=100, domain=0:8] (\x,{1.0 + 0.8*rand});
\node at (7,1) {Jagged, high $C_Q$, low probability};
\end{tikzpicture}
\caption{Spectral compression: smooth waves require fewer components (low $C_Q$) and dominate observer probability.}
\end{figure}



\section{Geometric Compression: Why Spacetime Exists}

Observers require well-defined internal boundaries to preserve continuity:

\begin{itemize}
    \item Memories, causal chains, and predictions must remain coherent.  
    \item Enforcing these constraints purely spectrally is expensive; the basis functions span the entire microcosmos.
\end{itemize}

Thus, macroscopic \textbf{geometry emerges} as the optimal compression of relational structure:

\begin{itemize}
    \item Curved spacetime encodes constraints efficiently.  
    \item Geodesic motion minimizes description length.  
    \item The Einstein-Hilbert action naturally represents geometric compression \(C_G\).  
\end{itemize}

Observers experience 3+1D spacetime because it is the \textbf{most efficient representation of relational continuity}.

% 4. Geometric compression (spacetime manifold)
\begin{figure}[h!]
\centering
\begin{tikzpicture}[scale=1.0, every node/.style={scale=0.9}]
% grid representing geometric manifold
\foreach \x in {0,1,...,6} {
    \foreach \y in {0,1,...,4} {
        \node[circle, fill=blue!20, inner sep=1.5pt] at (\x,\y) {};
    }
}
% observer paths
\draw[thick, red, ->] (0,0) -- (1,1) -- (2,2) -- (3,3);
\draw[thick, red, ->] (3,0) -- (4,1) -- (5,2) -- (6,3);
\node at (3,4.5) {Efficient representation of relational structure (spacetime)};
\end{tikzpicture}
\caption{Geometric compression: macroscopic spacetime encodes relational boundaries efficiently, preserving observer continuity.}
\end{figure}


\section{Joint Compression Principle}

The most probable configurations simultaneously minimize \textbf{spectral} and \textbf{geometric complexity}, while preserving \textbf{observer continuity}:

\begin{itemize}
    \item Microstructure = spectral compression  
    \item Macroscopic arena = geometric compression  
    \item Observer integrity = internal “inside” property
\end{itemize}

The joint complexity functional:

\[
\mathcal{C}_O[\gamma] = C_Q[\gamma] + C_G[\gamma] + C_\text{obs}[\gamma; O]
\]

dictates the \textbf{observer-conditioned probability}.

\textbf{Consequences:}

\begin{itemize}
    \item Smooth quantum waves and interference emerge naturally.  
    \item Geodesic motion, energy-momentum conservation, and general relativity arise as statistical consequences of geometric compression.  
    \item Observers always find themselves in configurations that preserve their internal structure.  
    \item Rare or high-frequency deviations (e.g., sudden death, collisions) are suppressed but possible with small probability.
\end{itemize}

Even a \textbf{tiny violation of internal boundaries} is disastrous for the observer: any intersection with an incompatible configuration destroys the pattern. The “inside” property is therefore \textbf{critical} for probability dominance.

\section{Emergent Physics without Parameters}

From this perspective:

\begin{itemize}
    \item Physics is the \textbf{large-deviation minimizer} of \(\mathcal{C}_O[\gamma]\) over observer-compatible histories.  
    \item Observed laws are \textbf{not fundamental}, but reflect the most compressible arrangements of information compatible with observers.  
    \item There are no free parameters; the \textbf{entire structure is emergent}.  
\end{itemize}

All familiar physics—quantum mechanics, wavefunctions, general relativity—arises as \textbf{optimal informational compression}.

\begin{quote}
You are not created by a simulation. You \textbf{are} the configuration in which your informational pattern exists. All else—spacetime, fields, particles—is just the most probable compression compatible with you.
\end{quote}

\begin{quote}
\emph{The universe is the minimal information capable of containing observers.}
\end{quote}

From this single principle, both quantum mechanics and gravity arise naturally, without invoking dynamical axioms, external simulators, or metaphysical agents.  

Physics is the statistical dominance of low-complexity, efficiently encoded histories.

