\chapter{Timeless Physics, Observer Conditioning, and Compressibility Measure}

\section{Static Configuration Space}

We assume a finite informational substrate of $n$ bits.
The total configuration space is

\[
\Omega = \{0,1\}^n,
\qquad |\Omega| = 2^n.
\]

The ensemble is static.
No configuration dynamically generates another.
There is no fundamental time parameter.

A \emph{history} is not fundamental.
Instead, an \emph{observer} is defined as an ordered sequence

\[
\gamma = (\omega_1, \omega_2, \dots, \omega_k),
\quad \omega_i \in \Omega,
\]

subject to:
\begin{itemize}
    \item memory continuity,
    \item internal logical consistency,
    \item compressibility constraints.
\end{itemize}

Experienced time is the ordinal index $i$ along such a sequence.
Time is therefore relational and emergent.

\section{Relation to Wheeler--DeWitt and Everett}

The Wheeler--DeWitt equation
\[
\hat{H}\Psi = 0
\]
describes a timeless universal wavefunction over superspace.

Similarly, Everettian quantum mechanics treats the universal state as static,
with all branches coexisting.

Both frameworks eliminate fundamental time but leave open:
\begin{itemize}
    \item Why quasi-classical histories dominate,
    \item Why wildly oscillatory universes are not typical,
    \item How probability arises non-circularly,
    \item Why observers experience ordered persistence.
\end{itemize}

The present framework agrees that the universe is timeless,
but introduces an additional ingredient:

\begin{center}
\emph{Observer-conditioned compressibility measure.}
\end{center}

Rather than modifying Wheeler--DeWitt, we reinterpret which solutions dominate.

\section{Compressibility as Fundamental Measure}

Let $D(\gamma)$ denote the description length of a sequence $\gamma$.
It includes:
\begin{itemize}
    \item Spectral complexity $C_Q(\psi)$,
    \item Geometric complexity $C_G(G)$.
\end{itemize}

Define the observer-conditioned probability:
\[
\mathbb{P}(\gamma \mid \text{observer})
\propto
\exp(-\lambda D(\gamma)),
\]
where $\lambda > 0$ is a universal scaling constant.

This replaces:
\begin{itemize}
    \item Euclidean weight $e^{-S_E}$,
    \item Lorentzian oscillatory weight $e^{iS}$,
\end{itemize}
with an information-theoretic penalty.

Histories that are expensive to encode are exponentially suppressed.

In contrast to the Euclidean rotation introduced by Stephen Hawking
to regulate oscillatory gravitational path integrals,
suppression here arises intrinsically from encoding cost.

\section{Reinterpreting the Path Integral}

The standard gravitational path integral
\[
Z = \int \mathcal{D}[g] \, e^{iS[g]}
\]
is reinterpreted as a sum over descriptions rather than trajectories:

\[
Z_{\mathrm{obs}}
=
\sum_{\gamma \ni \text{observer}}
\exp(-\lambda D(\gamma)).
\]

Key differences:
\begin{itemize}
    \item Sum over descriptions, not dynamical paths,
    \item Only observer-compatible sequences contribute,
    \item Suppression arises from encoding complexity rather than phase oscillation.
\end{itemize}

Highly oscillatory geometries possess:
\begin{itemize}
    \item Large spectral bandwidth,
    \item Poor compressibility,
    \item Large $D(\gamma)$.
\end{itemize}

They are therefore automatically suppressed.

\section{Spectral Compression and Quantum Mechanics}

The wavefunction is not ontologically primary.
It is a minimal spectral summary of ensembles of discrete histories
compatible with observer persistence.

Given a discrete trace $D$, many encodings exist.
Those requiring fewer independent frequencies and phases
are exponentially favored.

This naturally yields:
\begin{itemize}
    \item Linear superposition,
    \item Interference,
    \item Effective Hilbert space structure.
\end{itemize}

Quantum mechanics appears as the compression-optimal encoding language.

\section{Geometry and Gravity}

Geometry is not uniquely determined by spectral data.
There exists enormous degeneracy of geometries compatible with a given $\psi$.

Thus realized configurations minimize joint encoding cost:

\[
\gamma_{\rm realized}
=
\arg\min_\gamma
\left(
|\psi_\gamma| + |G_\gamma|
\right).
\]

Smooth, low-curvature manifolds dominate because:
\begin{itemize}
    \item They provide compact coordinate descriptions,
    \item They stabilize observer boundaries,
    \item They minimize adjacency encoding.
\end{itemize}

Gravity appears not as a fundamental force,
but as effective motion toward geometrically compressible configurations.

This explains why quantum mechanics and general relativity
appear complementary yet irreducible to one another.

\section{Holography as Compression Principle}

The holographic principle asserts that bulk degrees of freedom
can be encoded on boundaries.

In this framework, bulk redundancy is suppressed.
Minimal encodings concentrate information on boundary-like structures.

Dimensional reduction is not a postulate,
but a consequence of eliminating encoding redundancy.

\section{Arrow of Time}

Define accumulated description length:

\[
D_i = D(\gamma_{1:i}).
\]

The arrow of time corresponds to:

\[
\frac{d}{di} \mathbb{E}[D_i] > 0.
\]

Time flows in the direction of increasing minimal description.
Entropy increase becomes an encoding inevitability
under observer-conditioned selection.

\section{Interpolation and Extrapolation}

Observed past corresponds to interpolation (data-constrained encoding).

Future corresponds to extrapolation (model-constrained encoding).

Spectral extrapolation is generically unstable to high-frequency growth,
but compression penalizes such growth, stabilizing effective predictions.



\section{Conceptual Novelty}

This framework does not modify fundamental dynamical equations.
Instead, it introduces a non-dynamical, observer-conditioned typicality measure
based on computable description length.

Quantum mechanics, gravity, holography, and temporal flow
are interpreted as emergent statistical properties
of dominant compressible observer-compatible sequences.

\section{Scope and Falsifiability}

The theory would be undermined if:
\begin{itemize}
    \item Highly complex geometries were statistically dominant,
    \item Simple low-complexity unifications of $\psi$ and $G$ were found,
    \item Observed physics generically required high spectral bandwidth.
\end{itemize}

The framework does not yet replace predictive machinery.
It proposes a deeper explanatory substrate from which
existing theories may emerge as effective limits.

\section{Conclusion}

Quantum mechanics, general relativity, and holography
are not taken as axioms, but as consequences of observer-conditioned compression dominance.

The universe is fundamentally static. Dynamics is the experience of observers
navigating maximally compressible (most compressible) sequences within a timeless informational ensemble.
