\documentclass[11pt]{article}
\input{../common.tex}
\input{../glossary.tex}
\addbibresource{../references.bib}

\title{The  Derivation  of Reality}
\date{2024}

\begin{document}

\maketitle

\begin{abstract}
  A formal derivation of physical reality.
\end{abstract}

\ifdefined\ishtml
\begin{center}
\small \href{theory-of-everything.pdf}{Download PDF Version}
\end{center}
\fi


\section{Configurations, Paths, and the Origin of Time}

We begin from a minimal and representation-independent premise: reality is constituted by abstract information alone.
There is no privileged encoding, alphabet, ordering, geometry, probability measure, or temporal structure at the fundamental level.
Any concrete representation is a matter of convenience rather than ontology.

\subsection{Timeless Configuration Space}

For definiteness, we fix a binary representation.
With $n$ bits of information, the configuration space is
\[
\mathcal{C} = \{0,1\}^n,
\]
with cardinality $|\mathcal{C}| = 2^n$.
Each element $s \in \mathcal{C}$ is a complete, static configuration.

These configurations:
\begin{itemize}
    \item are unordered,
    \item all exist equally,
    \item are ontologically symmetric.
\end{itemize}

Most configurations resemble unstructured noise; only a vanishing fraction encode observers or observer-like structures.
Nevertheless, no configuration is ontologically privileged.
This timeless configuration space closely parallels both Everett-style interpretations (no collapse, no branching) and the Wheeler--DeWitt superspace of quantum gravity.

\subsection{Permutations and Movies}

One may conceptually arrange the configurations into complete orderings.
A permutation is a bijection
\[
\sigma : \mathcal{C} \to \mathcal{C},
\]
equivalently, a total ordering of all $2^n$ configurations.
The number of such permutations is $(2^n)!$.

Each permutation may be viewed as a complete ``movie'' containing every possible configuration exactly once.
This construction is mathematically valid, but observationally excessive:
almost all permutations are maximally incoherent and devoid of stable structure.
Permutations therefore serve as a conceptual upper bound, not as the fundamental object of experience.

\subsection{Observer Paths}

What observers actually experience are not permutations but \emph{paths}.
An observer path is a finite or semi-infinite sequence
\[
\gamma = (s_1, s_2, \dots, s_T), \quad s_i \in \mathcal{C},
\]
with no requirement to visit all configurations, and with possible repetition.

Observer paths:
\begin{itemize}
    \item traverse only a tiny subset of $\mathcal{C}$,
    \item preserve informational continuity,
    \item support memory, self-modeling, and prediction.
\end{itemize}

Every observer path is a subsequence of many permutations, but permutations are not fundamental.
The correct containment relation is:
\[
\text{Observer paths}
\;\subset\;
\text{finite sequences over } \mathcal{C}
\;\subset\;
\text{permutations of } \mathcal{C}.
\]

\subsection{Observer Filtering and Measure}

Among all possible paths, only a vanishing subset supports persistent observers.
These form the observer-compatible set $\mathcal{T}_{\mathrm{obs}}$.

Observer paths with high stepwise novelty---such as Boltzmann-brain-like trajectories involving random jumps across configuration space---are not forbidden.
They are logically and ontologically possible.
However, they carry exponentially suppressed measure.

The distinction is not one of possibility, but of typicality.
Paths that do not compress well have negligible weight under any reasonable description-length-based measure.

\subsection{Time as Induced Ordering}

The configuration space $\mathcal{C}$ is timeless.
Time is not fundamental; it is induced as the ordering along an observer path.

Observers traverse short, coherent subpaths through $\mathcal{C}$.
From regularities along these paths, they infer a past and extrapolate a future.
Cosmology, dynamics, and causality are therefore models of the path structure, not of the full configuration space.

This resolves the Wheeler--DeWitt timelessness without introducing external time or dynamics.

\subsection{Wavefunction as Compression}

The quantum wavefunction is not a physical medium, nor a dynamical law.
It plays the role of an optimal compression of the ensemble of observer-compatible paths.

Much like a video codec compresses a family of movies by exploiting shared structure, the wavefunction encodes:
\begin{itemize}
    \item large ensembles of compatible paths,
    \item interference as shared substructure in the compressed description,
    \item probabilities as relative description weight.
\end{itemize}

There is no external player, no execution substrate, and no need for temporal unfolding.
The existence of the information itself is sufficient for lived experience.


\subsection{Why Smooth Compressed Worlds Dominate}

Although chaotic and non-compressed observer paths exist, smooth, compressed, predictable paths dominate measure.
Given any fixed amount of information, that information can be arranged to form smooth observer movies in compressed format in vastly more ways than it can be arranged to form highly irregular ones without compression.

A random draw from the ensemble of observer paths is therefore overwhelmingly likely to yield a lawful, continuous world with compression artifacts (QM).
Our experience of smooth physics is not imposed by law, but emerges from combinatorial dominance under compression-weighted selection.

\section{The Eight Ladders to the Abstract Universe}

\begin{gather*}
  \mathbf{P} = ? \\
    \text{\small (Infinite informational potential)} \\
    \downarrow \\
    \mathcal{C} \cong \{0,1\}^n \\
    \text{\small (A convenient discretization into a static configuration space; no preferred base)} \\
    \downarrow \\
    \mathcal{S} = \{ (s_1,\dots,s_T) \mid s_i \in \mathcal{C} \} \\
    \text{\small (The space of all finite and semi-infinite paths through configuration space)} \\
    \downarrow \\
    \mathcal{T}_{\mathrm{obs}} = \{ \gamma \in \mathcal{S} \mid \mathrm{Observer}(\gamma)=1 \} \\
    \text{\small (Observer Filter: paths supporting stable informational recursion and continuity)} \\
    \downarrow \\
    \mathcal{L}(\gamma) = \text{Minimal spectral encoding length of path } \gamma \in \mathcal{T}_{\mathrm{obs}} \\
    \text{\small (Favoring smooth, predictable, and compressible observer histories)} \\
    \downarrow \\
    \Psi(\gamma) \propto 2^{-\mathcal{L}(\gamma)/2} \\
    \text{\small (The Spectral Realization: wavefunction as optimal compression of observer paths)} \\
    \downarrow \\
    P(\gamma) = |\Psi(\gamma)|^2 \\
    \text{\small (Born Rule: relative measure of experienced histories)} \\
    \downarrow \\
    \rho(\mathbf{s}) \sim \mathrm{Lognormal}(\mu,\sigma^2) \\
    \text{\small (Statistical density of reusable microstructure motifs in configuration space)} \\
    \downarrow \\
    \delta \int_{\gamma} \mathcal{L}(\text{state})\, d\lambda = 0
    \;\;\Longrightarrow\;\;
    \text{Geodesics} \\
    \text{\small (Gravitation as the geometric manifestation of compression pressure)}
\end{gather*}

\newpage

\section{The Ladders}

\subsection*{Philosophical Grounding}

A Theory of Everything must be self-contained; any theory predicated upon empirical assumptions is merely a subset.
If a theory requires an external constraint to define its constants or boundaries, or if it fails to account for us—conscious observers—it remains incomplete.

Is the universe finite or infinite? If it were finite, the source and specific value of that constraint (e.g., "Why N bits and not N+1?") would require a meta-theory to explain.
To avoid an infinite regress of constraints, we must conclude that the fundamental nature of reality is unbounded and infinite.

Why is there something rather than nothing? The most rational answer is that "nothingness" is but one unique configuration among an infinite ensemble of possibilities.
Within an infinite potential, the mathematical probability of absolute non-existence is zero.

Therefore, the fundamental nature of reality is probabilistic.
The universe is the most probable outcome in which conscious observers are likely to find themselves wandering and wondering.


\subsection*{Ladder 1: The Pre-Reality}
Reality begins as abstract information itself, with no privileged representation, alphabet, ordering, geometry, or probability measure.
Any concrete encoding (binary, ternary, continuous, or otherwise) is a matter of convenience rather than ontology.
Entropy, time, and physical law arise only relative to observer-defined representations and coarse-grainings.
There is no distinguished substrate or execution mechanism; existence is purely informational.

\subsection*{Ladder 2: The Observer Exists}
Since $\mathbf{P}$ excludes nothing, it necessarily contains finite informational subsets capable of self-reference.
The existence of structured, observer-containing worlds requires no external cause; such structures are unavoidable elements of infinite informational potential.
Observers are not added to reality—they are configurations within it.

\subsection*{Ladder 3: The Static Multiverse}
A concrete world may be represented as a finite configuration space
\[
\mathcal{C} \cong \{0,1\}^n,
\]
where each configuration $\mathbf{s} \in \mathcal{C}$ is a static informational state.
There is no motion or dynamics at this level—only the coexistence of all configurations.
The binary representation is not fundamental; it is chosen purely for convenience.
All configurations are ontologically equal.

\subsection*{Ladder 4: The Emergence of Time}
Time is not a fundamental dimension but an induced ordering.
It arises from traversals through configuration space.
An ordering of states is not given a priori; it is defined only relative to observer paths.
Timeless configuration space thus precedes experienced temporality.

\subsection*{Ladder 5: The Observer Filter}
An observer history is a finite or semi-infinite path
\[
\gamma = (s_1, s_2, \dots), \qquad s_i \in \mathcal{C}.
\]
Only a vanishing subset of all possible paths supports persistent informational recursion, memory, and self-modeling.
These paths form the observer-compatible set
\[
\mathcal{T}_{\mathrm{obs}} \subset \mathcal{S},
\]
where $\mathcal{S}$ denotes the space of all paths through $\mathcal{C}$.
Only paths in $\mathcal{T}_{\mathrm{obs}}$ possess a subjective ``inside'' and are experienced as reality.

\subsection*{Ladder 6: The Spectral Selection Principle (SSP)}
Observer paths are not equally weighted.
Their measure is governed by compressibility in a spectral representation.
Paths admitting efficient spectral encodings dominate the ensemble of experienced histories.
Formally, preferred histories minimize spectral description length:
\[
H_{\mathrm{SSD}} = \arg\min_H \mathcal{L}(H),
\]
where $\mathcal{L}(H)$ denotes the minimal spectral encoding length of path $H$.
This statistical dominance yields an emergent, finite, and predictable physics without invoking fundamental dynamics.

\subsection*{Ladder 7: The Wavefunction as Optimal Encoding}
The quantum wavefunction $\Psi$ is not a physical field but an optimal spectral compression of the observer-path ensemble.
Superposition and interference reflect shared substructure within compressed descriptions.
The Born Rule emerges as a measure over paths:
\[
P(\gamma) \propto 2^{-\mathcal{L}(\gamma)} \sim |\Psi(\gamma)|^2.
\]
Thus probability is not fundamental; it is a consequence of informational weighting.


\subsection*{Ladder 8 --- Emergent Gravitation: Spectral Compression Geodesics}

Physical gravitation emerges as a macroscopic manifestation of the statistical weighting of observer-compatible paths in configuration space.

\paragraph{1. Observer paths and spectral measure:}  
Let $\mathcal{C}$ be the static configuration space, and $\mathcal{T}_{\mathrm{obs}} \subset \mathcal{S}$ the set of observer-compatible paths $\gamma = (s_1, s_2, \dots)$.
Each path is assigned a spectral encoding length $\mathcal{L}(\gamma)$ and a corresponding quantum amplitude:
\[
\Psi(\gamma) \propto 2^{-\mathcal{L}(\gamma)/2}, \qquad P(\gamma) = |\Psi(\gamma)|^2 \propto 2^{-\mathcal{L}(\gamma)}.
\]

\paragraph{2. Variational principle:}  
The dominant paths are those minimizing cumulative spectral cost:
\[
\delta \int_\gamma \mathcal{L}(\text{state})\, d\lambda = 0.
\]
These paths correspond to the classical trajectories experienced by observers.

\paragraph{3. Effective metric from microstructure density:}  
Let $\rho(\mathbf{s})$ denote the density of reusable microstructure motifs within configuration space.  
High-density regions correspond to low spectral cost; an observer path naturally ``falls'' toward them.  
In the continuum limit, the effective metric $g_{\mu\nu}$ is defined by the gradient of spectral cost:
\[
\mathcal{L}(\gamma) \approx \int_\gamma \sqrt{g_{\mu\nu} \, \dot{x}^\mu \dot{x}^\nu} \, d\lambda,
\]
so that minimizing $\mathcal{L}$ coincides with minimizing proper length in spacetime — i.e., geodesics.

\paragraph{4. Connection to quantum theory:}  
Summation over all $\gamma \in \mathcal{T}_{\mathrm{obs}}$ weighted by $2^{-\mathcal{L}(\gamma)}$ reproduces a path-integral-like formalism:
\[
\langle \mathcal{O} \rangle = \sum_{\gamma \in \mathcal{T}_{\mathrm{obs}}} \mathcal{O}(\gamma) \, 2^{-\mathcal{L}(\gamma)},
\]
where $\mathcal{O}$ is any observable depending on path configurations.  

- \textbf{Everettian view:} All observer-compatible paths exist simultaneously; probability arises from spectral measure, not collapse.  
- \textbf{Wheeler–DeWitt view:} Configuration space is timeless; time emerges as an ordering along $\gamma$.  
- \textbf{Classical limit:} Low-$\mathcal{L}$ paths dominate, producing smooth geodesic motion as in GR.

\paragraph{5. Summary:}  
Gravitation is thus not a fundamental force but a statistical-geometric phenomenon:  
\emph{classical geodesics are the trajectories of minimal spectral encoding length through configuration space, weighted by observer-compatible path measure.}  

The ``singularity'' at vanishing microstructure density ($\rho \to 0$) corresponds to an algorithmic limit where observer-compatible structure (encoding matter and enrgy)
vanishes, making the region effectively unreachable.


\printbibliography

\end{document}

