\documentclass[11pt]{article}
\input{../../preamble.tex}
\input{../../paper_preamble.tex}
\addbibresource{../references.bib}

\title{The Wavefunction as Compression:\\
Spectral Complexity, Emergent Quantum Behaviour,\\
and the Informational Action Principle}

\author{Juha Meskanen}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
  Papers~I--V \cite{meskanen2001, meskanen2002, meskanen2003, meskanen2026iv}
  established a purely information-theoretic framework in which
spacetime geometry, gravitational collapse, cosmic expansion, and the
relational scale factor all emerge from a finite bit budget $n$ without
hard-coded physical laws. One feature of the observed universe remains
unexplained by that framework: the wave-like behaviour of the microcosm.
This paper proposes and investigates the hypothesis that the quantum
wavefunction is the universe's data-compression codec. Internal observers,
themselves composed of compressed structures, perceive their constituent
degrees of freedom as wave-like for the same reason that pixel-observers
inside an MPEG-compressed movie would perceive their world as governed by
a Discrete Cosine Transform: they are observing compressed information.
We formalise and implement this as a \emph{Spectral Complexity} measure $C_s$ --- a
continuous, computable replacement for Kolmogorov complexity --- and show
that Solomonoff-like suppression under $C_s$ selects smooth, law-like,
wave-governed configurations over chaotic ones, resolving the Boltzmann
brain problem without additional axioms. Numerical simulations demonstrate
the emergence of inertia and interference from spectral compression alone.
\end{abstract}

\tableofcontents
\newpage

% ============================================================
\section{Introduction}
\label{sec:intro}

The framework developed in Papers~I--V recovers a range of
physical structure from a single axiom: the universe is a finite, static,
timeless system of $n$ bits. Gravitational collapse converges to a
zero-entropy singularity (Paper~I). Expansion from that singularity is
the geometric reading of entropy increase, with emergent microstructures
following filter-independent lognormal distributions (Paper~II). The
aspect ratio of spacetime is determined by $n$ alone, recovering
Bekenstein--Hawking entropy to within the geometric factor $4\pi$
(Paper~III). Matter contracts observable space relationally, reproducing
the three-phase Friedmann expansion profile without a cosmological
constant (Paper~IV).

One feature of the observed universe is not explained by this framework:
the wave-like behaviour of the microcosm. Particles exhibit interference,
superposition, and discrete measurement outcomes. The wavefunction is
complex-valued. Dynamics are unitary. Why?

This paper proposes a hypothesis and investigates its consequences.

\textbf{The Wavefunction Compression Principle (WCP).}
\textit{The cosmos waves because we are observing compressed structure.
The quantum wavefunction is the universe's data-compression codec.}

The argument is developed in four steps. Section~\ref{sec:analogy}
motivates the hypothesis through the MPEG analogy.
Section~\ref{sec:spectral} formalises spectral complexity as a
continuous, computable measure and shows how it suppresses Boltzmann
brains. Section~\ref{sec:simulations} presents numerical results:
emergence of inertia, interference, and smooth spacetime from spectral
compression. Section~\ref{sec:conclusion} states the results and open
problems.

% ============================================================
\section{The Compression Hypothesis}
\label{sec:analogy}

\subsection{The MPEG Analogy}

Consider a conventional MPEG-compressed digital movie. Each frame is
composed of discrete pixels. If observers existed inside such a
movie, made of pixels and perceiving only pixel-level interactions, what
would they observe as their laws of physics?

They would find that their constituent pixels follow mysterious,
deterministic, wave-like patterns. To a pixel-physicist, these transitions
would appear fundamental. In reality, they are the mathematical output of
the Discrete Cosine Transform (DCT) --- the compression basis that MPEG
codecs use to represent raw pixel data in a maximally compact form.

The parallel with quantum mechanics is direct. The quantum wavefunction
distributes probability amplitudes across basis states in Hilbert space
exactly as a spatial Fourier transform distributes image information across
frequency coefficients. Complex-valued amplitudes are exceptionally
efficient at encoding relational information: they produce precisely the
spatial smoothness, structural continuity, and predictive regularity that
stable observers require.

\textbf{The hypothesis is therefore:} the wave-like behaviour of the
microcosm is not a primitive feature of reality but the signature of
compressed information as perceived from within. An observer composed of
compressed structures will perceive their world as wave-governed for the
same reason that a pixel-observer would perceive DCT dynamics: the
compression codec \emph{is} the physics.

\subsection{The Dithering Analogy and the Born Rule}

The Born rule --- the identification of measurement probabilities with
squared wavefunction amplitudes --- has a natural interpretation under
this hypothesis.

Consider a rendering engine producing a sphere with intended shading
intensity $0.85$, on hardware restricted to discrete outputs of $0.8$ or
$0.9$. The engine employs \emph{dithering}: distributing $0.8$ and $0.9$
across adjacent pixels with relative frequencies $0.15$ and $0.85$
respectively, so that the perceived average matches the target. The
discrete output statistics are determined by the continuous target value.

A quantum superposition
\[
|\psi\rangle = \alpha|0.8\rangle + \beta|0.9\rangle
\]
is the universe's implementation of this same principle. The Born rule
$P = |\alpha|^2$ is the probability rule of an optimally compressed
continuous description rendered through a discrete measurement apparatus.
The squaring arises because amplitudes are complex-valued: the
information-theoretic weight of a mode is proportional to the power in
that mode, which is the squared amplitude.

This does not constitute a derivation of the Born rule --- that would
require a proof from the spectral complexity axioms alone. It is offered
as a motivation: the Born rule is what optimal compression looks like when
a continuous description meets a discrete measurement.


\subsection{Fermions, Bosons and Pauli Exclusion}

Fermions play the role of pixels - real particles, sampled from the wavefunction through Born rule.

Bosons don't exist as fundamental objects. They are explanatory fictions — the story we tell about
why fermions (pixels) move the way they do, when we don't know they're being sampled from a wavefunction.

The actual causal chain is:

\[
\text{wavefunction (codec)} \xrightarrow{\text{Born rule sampling}} \text{fermion configurations (pixel frames)}
\]

But an observer who only sees the pixel outputs, not the codec, invents a story:

\[
\text{fermion} \xrightarrow{\text{"virtual boson exchange"}} \text{fermion}
\]

The boson is the compression residual made into a noun. It's not wrong — it's a valid effective description —
but it's not fundamental.

The analogy is exact: in MPEG, macroblocks don't "exchange" anything.
The DCT coefficients just are what they are. But if one only watched the decoded pixels
and tried to explain their correlations without knowing about DCT, one would invent something
that looks exactly like boson exchange.

This explains natively why fermions suffer from Pauli exclusion whereas bosons not.


% ============================================================
\section{Spectral Complexity}
\label{sec:spectral}

\subsection{From Kolmogorov to Spectral Complexity}

Kolmogorov complexity provides the theoretical foundation for minimal
description length but is fundamentally uncomputable, discrete, and
discontinuous. It cannot smoothly support stable prediction, continuous
spacetime geometry, or the gradual evolutionary reasoning of an internal
observer.

The observational evidence suggests that physical systems already possess
a natural mode of information representation: decomposition into spectral
modes. This motivates a continuous, computable replacement.

\begin{definition}[Spectral Complexity]
The \emph{spectral complexity} $C_s(\Psi)$ of a state $\Psi$ is the total
informational cost required to uniquely specify the amplitudes, frequencies,
and phases of the spectral modes composing $\Psi$:
\[
C_s(\Psi) = C_{\mathrm{base}} + \sum_{i=1}^{N}
\left[ C(\phi_i) + C(A_i) + \frac{\omega_i}{\Delta\omega} \right]
\]
where $C_{\mathrm{base}}$ is the $\mathcal{O}(1)$ cost of the underlying
trigonometric subroutines, $C(\phi_i)$ is the fixed-width bit cost of
encoding the phase $\phi_i \in [0, 2\pi)$, $C(A_i)$ is the bit depth
required to encode amplitude $A_i$, and $\omega_i / \Delta\omega$ is the
dominant term representing the linear resource cost of tracking mode
frequency $\omega_i$ relative to the minimum resolution $\Delta\omega$.
\end{definition}

Three properties of this measure are essential.

\textbf{Computability.} Unlike Kolmogorov complexity, $C_s$ is directly
computable from the spectral decomposition of any state.

\textbf{Continuity.} $C_s$ assigns a smooth, continuous cost gradient
across neighbouring states. Small perturbations in frequency or amplitude
produce small changes in complexity.

\textbf{Linear frequency scaling.} The dominant cost term $\omega_i /
\Delta\omega$ scales linearly with frequency. This mirrors the physical
relation $E = \hbar\omega$: energy scales linearly with frequency.
Under Solomonoff suppression $P \propto 2^{-C_s}$, this linear cost
produces exponential suppression of high-frequency modes:
\[
P \propto 2^{-\omega/\Delta\omega} = e^{-(\ln 2)\,\omega/\Delta\omega},
\]
which is a Boltzmann distribution with $\beta = \ln 2 / \Delta\omega$.
The identification $\hbar = \Delta\omega / \ln 2$ connects the minimum
frequency resolution of the spectral complexity measure to Planck's
constant. This is not a derivation but a precise identification that
warrants further investigation.

\subsection{Resolution of the Boltzmann Brain Problem}

Under the Solomonoff prior $P(s) \propto 2^{-L(s)}$ applied to spectral
complexity, the probability of a configuration is exponentially
suppressed by its spectral cost. Chaotic, high-entropy configurations ---
Boltzmann brains, random fluctuations, disordered universes --- have high
spectral complexity: they require many high-frequency modes with large
amplitudes to describe. Their probability under $C_s$ is therefore
exponentially suppressed relative to smooth, law-like, compressible
configurations.

This resolves the Boltzmann brain problem without additional axioms. The
selection of structured observers is not a fine-tuning accident. It is a
direct consequence of the measure: smooth, predictable data compresses
better than chaos, and the spectral complexity measure weights
configurations by their compressibility. We do not find ourselves in a
chaotic universe because chaotic universes are exponentially expensive to
describe.

% ============================================================
\section{Numerical Results}
\label{sec:simulations}

\subsection{Emergence of Smooth Spacetime}

Applying the spectral complexity cost function as a selection weight to
the bitstring evolution of Paper~II --- replacing uniform sampling with
$C_s$-weighted sampling --- immediately suppresses the high-entropy white
noise of the unweighted ensemble. The dominant configurations are those
with low spectral complexity: smoothly varying, wave-governed spatial
profiles. Symmetric wave packets and lattice-like structures emerge as the
most probable outcomes without being hard-coded.

\subsection{Emergence of Inertia and Interference}

Two specifically quantum-mechanical phenomena emerge from spectral
compression in numerical simulation, without any imposed equations of
motion.

\textbf{Inertia.} A localised wave packet moving through the spectral
field maintains its velocity without external forcing. The minimum-$C_s$
continuation of a moving packet is the packet continuing to move: any
deflection increases the spectral cost by introducing new frequency
components. Resistance to deflection --- inertia --- is the geometric
consequence of spectral economy.

\textbf{Interference.} When two wave packets overlap, the minimum-$C_s$
description of the combined state is not the sum of two independent
descriptions but a single spectral decomposition of the superposition.
The cross-terms --- interference fringes --- are cheaper to encode than
two separate packets because they share spectral modes. Interference
emerges as the compression-optimal description of overlapping structures.



% ============================================================
\section{Conclusion}
\label{sec:conclusion}

This paper establishes three results and states one conjecture.

\textbf{First}, the Wavefunction Compression Principle provides a
hypothesis for why the microcosm waves: internal observers composed of
compressed structures perceive their world as wave-governed because the
compression codec of the universe is the wavefunction. 

\textbf{Second}, spectral complexity --- a continuous, computable
replacement for Kolmogorov complexity with linear frequency cost ---
exponentially suppresses chaotic configurations under Solomonoff-like
induction. Structured, law-like, compressible universes dominate the
measure. The Boltzmann brain problem is resolved as a consequence of the
measure, not by fine-tuning.

\textbf{Third}, numerical simulation demonstrates that inertia and
interference emerge from spectral compression alone, without imposed
equations of motion.

\textbf{Conjecture} (Informational Action Principle): $C_s \propto
S_{\mathrm{Euclidean}}$, with proportionality constant determined by $n$
and $\Delta\omega$. If proved analytically or demonstrated to arbitrary
numerical precision, this conjecture identifies quantum gravity as
Solomonoff induction over compressed geometric descriptions, completes
the bridge from the informational framework of Papers~I--VI to a full
theory of quantum gravity, and derives $\hbar$ as the minimum spectral
resolution of a finite informational universe.

% ============================================================
\section*{Open Problems}
\label{sec:open}

\begin{enumerate}
  \item \textbf{Show that the spectral complexity and the Euclidean action functional are proportional} under the
        informational metric of Paper~IV; or numerically, by extending
        the path comparison to higher-dimensional surfaces and quantifying
        the deviation.

  \item \textbf{Derive the Born rule.} Show that $P = |\alpha|^2$ follows
        from the spectral complexity axioms, using Gleason's theorem or a
        direct information-theoretic argument.

  \item \textbf{Derive $\hbar$ from $n$.} If $\hbar = \Delta\omega / \ln 2$
        and $\Delta\omega$ is the minimum frequency resolution of a system
        with $n$ bits, express $\hbar$ as a function of $n$ and verify
        against the observed value using $n \approx 184$ from Paper~IV.

  \item \textbf{Derive $G$ from $n$.} Paper~IV recovered Bekenstein--Hawking
        entropy up to $4\pi$ without using $G$. A complete theory should
        express Newton's constant as a function of $n$ alone.
\end{enumerate}

% ============================================================
\section*{Simulation Code}

\begin{itemize}
  \item \texttt{wavefunction.py}: Complex-valued wavefunction with Spectral Complexity measure. 
  \item \texttt{emergent\_inertia\_interference.py}: Spectral compression POC, demonstrating emergent inertia and interference.
\end{itemize}

The \texttt{wavefunction} package can be installed from python \textit(pypi) repository.

\begin{lstlisting}[language=bash, caption=Example usage]
pip install wavefunction
python simulations/emergent\_inertia\_intererence.py --blobs 3 --steps 400 --k\_mod 30
\end{lstlisting}

\printbibliography

\end{document}
