\documentclass[11pt]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm, bm}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}

% --- Page Setup ---
\geometry{letterpaper, margin=1in}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

% --- Document Metadata ---
\title{\textbf{Einstein Equations as Probability Laws: \\ A Semiclassical Information-Theoretic Approach}}
\author{Juha Meskanen}
\date{June 2021}

\begin{document}

\maketitle

\section{Motivation}
\begin{itemize}
    \item \textbf{Geometric Entropy:} The discovery that while total physical entropy increases, the geometric space of a collapsing black hole can be understood as a geometric interpretation of information with decreasing entropy. Conversely, the Big Bang represents increasing entropy.
    \item \textbf{Emergent Laws:} Wavefunctions and General Relativity are emergent features arising from information serving as optimal compression algorithms.
\end{itemize}

\section{Core Concept: From Einstein to Probability}
Let worldlines be random paths $\gamma$ on a manifold $M$. We define a probability distribution on paths centered on General Relativity (GR):
\[
\mathbb{P}[\gamma] \propto \exp\!\left( -\frac{1}{2}\int_\gamma \frac{\|a_\gamma(\tau)-a_{\rm GR}(\tau)\|^2}{\sigma^2(\mathcal{I}(x(\tau)))}\, d\tau \right)
\]
where $a_\gamma$ is the stochastic 4-acceleration, $a_{\rm GR}$ is the GR geodesic acceleration, and $\sigma(\mathcal{I})$ is a scale-dependent noise amplitude depending on a curvature invariant $\mathcal{I}$ (e.g., the Kretschmann scalar $K=R_{abcd}R^{abcd}$).

\section{Multiplicative (Lognormal) Kinematics}
For any coordinate that shrinks under attraction (such as the radial coordinate $r$ in Schwarzschild geometry), we evolve the system multiplicatively in proper time:
\[
\ln r(\tau+\Delta\tau) = \ln r(\tau) + \mu(r)\,\Delta\tau + \sigma(\mathcal{I}(r))\,\sqrt{\Delta\tau}\,\xi
\]
with $\xi \sim \mathcal{N}(0,1)$. Consequently, increments of $\ln r$ are Gaussian, making the radial steps lognormal. The drift $\mu(r)$ is calibrated so that the most probable trajectory recovers the GR geodesic.

\section{Noise Suppression: The Entropy--Singularity Lemma}
\begin{quote}
\textit{Vanishing entropy corresponds to a collapsing geometry, i.e., a geometric singularity.}
\end{quote}

If entropy counts accessible microstates, then at the singularity there are none left ($S \to 0$). Therefore, the probability of particle emission or fluctuation at the singularity must vanish.

\subsection{Entropy as a Function of Curvature}
Let $I$ be a curvature invariant. We define a coarse-grained entropy $S(I)$ such that:
\[
S(I) \ge 0, \quad S(I) \xrightarrow[I\to 0]{} S_0 > 0, \quad S(I) \xrightarrow[I\to \infty]{} 0
\]
We model this using exponential decay:
\[
S(I) = S_0 \exp\!\left[-(I/I_P)^m\right], \quad m > 0
\]
where $I_P \sim \ell_P^{-4}$ sets the Planck curvature scale.

\subsection{Noise Amplitude and Probability Voids}
We set the noise variance proportional to entropy: $\sigma^2(I) = \kappa\, S(I)$.
\begin{itemize}
    \item \textbf{Far from the singularity ($I \ll I_P$):} $\sigma^2(I) \approx \sigma_0^2$. GR with finite fluctuations is recovered.
    \item \textbf{Near the singularity ($I \gg I_P$):} $\sigma^2(I) \to 0$. Fluctuations are quenched.
\end{itemize}
The singularity becomes a \textbf{probability void}. There is no stochastic flux into the singular region because there are no available microstates to support existence.

\section{Stress-Energy as a Bitflip Field}
We tie matter to information via a Poisson--lognormal field of bitflips per 4-volume:
\begin{align*}
N(x)\mid\Lambda(x) &\sim \mathrm{Poisson}(\Lambda(x)) \\
\ln \Lambda(x) &\sim \mathcal{N}\!\big(\mu_\Lambda(x),\, s^2_\Lambda(x)\big)
\end{align*}
Mapping flips to local stress-energy $T_{ab}[N]$, we take Einstein-in-expectation:
\[
G_{ab}(x) = 8\pi G\,\mathbb{E}[T_{ab}[N]] + \text{(finite fluctuation terms)}
\]

\section{Einstein to Probability Dictionary}
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{p{4cm} l}
\textbf{Deterministic GR:} & $G_{ab} = 8\pi G\,T_{ab}$ \\
\hline
\textbf{Probabilistic GR:} & $\mathbb{P}[\gamma] \propto \exp\!\left(-\frac{1}{2} \int \frac{\|a_\gamma-a_{\rm GR}\|^2}{\sigma^2(\mathcal{I})}\, d\tau\right)$ \\
\textbf{Matter Field:} & $N \mid \Lambda \sim \text{Poisson}(\Lambda), \quad \ln\Lambda \sim \mathcal{N}(\mu_\Lambda, s_\Lambda^2)$ \\
\textbf{Expectation:} & $G_{ab} = 8\pi G\,\mathbb{E}[T_{ab}[N]] + \text{finite cumulants}$ \\
\textbf{Kinematics:} & $\Delta \ln r = \mu(r)\Delta + \sigma(\mathcal{I})\sqrt{\Delta}\,\xi$ \\
\end{tabular}
\end{center}

\end{document}
