\documentclass[11pt]{article}
\input{../../preamble.tex}
\input{../../paper_preamble.tex}
\addbibresource{../references.bib}

\title{Emergence of Quantum Mechanics: Fermions, Bosons and Pauli Exclusion}
\author{Juha Meskanen}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
Paper~VI \cite{meskanen2026v} presented the hypothesis that the quantum mechanical
wavefunction is the compression signature of nature: the universe is a finite,
static bitstring of $n$ bits, and the path an internal observer experiences is the
one that compresses best under a complex-valued spectral codec.
The present paper supports this hypothesis with seven experiments implemented in Python
and included as supplementary material.
We show that a photon-like boson emerges as the compression residual of a purely
fermionic system, without any interaction term introduced by hand.
All essential features of quantum mechanics are recovered: Born rule, Pauli
exclusion, particle--antiparticle pairing, and the virtual propagator.
The central result is an analytically exact theorem:
for a single fermion in superposition between two sites, the off-diagonal
(boson) amplitude is $\|B(\theta)\| = \sin(2\theta)/\sqrt{2}$, independent of
system size, site separation, and encoding.
This equals $\sqrt{2\,P(1-P)}$, the interference term of the Born rule, made
visible as a matrix norm.
We further report an exploratory extension in which the same $B$-matrix
formalism is applied, unchanged, to a two-band lattice Hamiltonian
$H(k) = m\,\sigma_z + t\sin(k)\,\sigma_x$, giving a boson-residual amplitude
$\|B(k)\| = |t\sin k|/(\sqrt{2}\,E(k))$ with $E(k)=\sqrt{m^2+(t\sin k)^2}$,
which recovers the relativistic dispersion relation
$E^2 = (pc)^2 + (m_0c^2)^2$ as its continuum limit.
This result is numerically verified but not yet derived from the spectral
complexity functional $C_s$, and is reported as a genuinely open problem
rather than a closed theorem.
These results suggest that quantum mechanics is not a foundational layer of
reality but an emergent description --- the story an observer tells about
fermions sampled from a compression codec.
\end{abstract}

\tableofcontents
\newpage

% ============================================================
\section{Introduction}
\label{sec:intro}

Why does the universe exhibit wave-like behaviour at small scales?
Standard quantum mechanics takes the wavefunction as a primitive object and
the Born rule as a postulate.
The present paper pursues a different direction, building on the informational
framework of Papers~I--VI \cite{meskanen2001, meskanen2002, meskanen2003, meskanen2026iv, meskanen2026v, meskanen2026vi}:
the universe is a finite, static collection of $n$
bits, and an internal observer experiences a path through its $2^n$ configurations.
The path selected is the one that compresses best.
The wavefunction is the codec.

The analogy that makes this precise is video compression.
In MPEG, a video is encoded not as a sequence of raw frames but as a sequence
of compressed descriptions.
The compression basis is the Discrete Cosine Transform (DCT) --- a real-valued
basis of smooth periodic functions.
The complex-valued wavefunction of quantum mechanics is the natural
generalisation: it is the simplest continuous, smooth, periodic codec for
a universe whose configurations vary in both amplitude and phase.

Under this interpretation, particles fall into two classes that correspond
directly to the two components of a compressed video:

\begin{itemize}
  \item \textbf{Fermions} are pixels: discrete, localised, non-overlapping by
        construction. Two pixels cannot occupy the same site --- not by postulate,
        but because a pixel is a site. Pauli exclusion is a pixel axiom.
  \item \textbf{Bosons} are compression residuals: the off-diagonal entries of
        the density matrix $\rho = |\psi\rangle\langle\psi|$ that remain after
        the per-site (fermionic) diagonal is removed. They are what the codec
        records when a fermion moves.
\end{itemize}

The actual causal chain is:
\[
  \text{wavefunction (codec)}
  \xrightarrow{\text{Born rule sampling}}
  \text{fermion configurations (pixel frames)}
\]
An observer who sees only the pixel outputs, not the codec, invents a story:
\[
  \text{fermion}
  \xrightarrow{\text{``virtual boson exchange''}}
  \text{fermion}.
\]
The boson is a compression residual made into a noun.
It is not wrong as an effective description, but it is not fundamental.
In MPEG, macroblocks do not exchange anything; the DCT coefficients just are
what they are.
If one watched only the decoded pixels and tried to explain their correlations
without knowing about DCT, one would invent something that looks exactly like
boson exchange.

The seven experiments in this paper test whether this picture is self-consistent
and quantitatively correct.

% ============================================================
\section{Experiment 1: Minimal Video}
\label{sec:exp1}

\textit{Python: minimal\_video.py}

We encode 2-frame videos of a $2\times 2$ pixel grid as complex wavefunctions:
\[
  \psi(x) = f_0(x) + i\,f_1(x),
\]
where frame~0 is the real part and frame~1 is the imaginary part.
Each video is scored by the spectral complexity $C_s$ of $\psi$ --- the
information-theoretic cost of its Fourier representation.
The off-diagonal density matrix $B = \rho - \mathrm{diag}(\rho)$ is computed
for all $2^8 = 256$ possible two-frame videos, and its Frobenius norm
$\|B\|$ is taken as the boson signal.

\subsection*{Results}

The minimum-complexity states ($C_s = 0$) --- uniform frames, all pixels
identical --- already carry the maximum boson signal $\|B\| = \sqrt{3}/2$.
The maximally anti-correlated configuration (checkerboard transition, flipping
to its complement) also achieves $\|B\| = \sqrt{3}/2$ at $C_s = 3$:
\[
  f_0 = \begin{pmatrix}1&1\\0&0\end{pmatrix}
  \;\longrightarrow\;
  f_1 = \begin{pmatrix}0&0\\1&1\end{pmatrix}.
\]
The vacuum (zero information content) and the maximally structured transition
produce identical boson signal.

\textbf{Conclusion:} The boson signal tracks something topological about the
pixel transition, not its complexity. This motivates a structural analysis of
the off-diagonal matrix.

% ============================================================
\section{Experiment 2: Boson Structure}
\label{sec:exp2}

\textit{Python: boson\_structure.py}

Rather than measuring $\|B\|$ alone, we analyse the structure of $B$:
\begin{itemize}
  \item \emph{Symmetry:} is $B$ symmetric (spin-0 candidate) or
        antisymmetric (spin-1 candidate)?
  \item \emph{Propagation:} does the dominant eigenvector of $B$ resemble a
        plane wave?
  \item \emph{Quantisation:} is $\|B\|$ a simple rational multiple of
        $2^{-C_s}$?
\end{itemize}

\subsection*{Results}

For the video $(1, 2)$ --- a single fermion hopping from site~0 to site~1 ---
the boson matrix is purely antisymmetric, not approximately but exactly.
Its eigenvalues are $\pm 1/2$, symmetric about zero.
The forward hop $(1,2)$ and reverse hop $(2,1)$ produce identical boson
structure with opposite momenta $k = \pm 0.157$, exactly as expected from a
particle--antiparticle pair.

For the block hop $(3,12)$ --- two fermions moving together,
\[
  f_0 = \begin{pmatrix}1&0\\1&0\end{pmatrix}
  \;\longrightarrow\;
  f_1 = \begin{pmatrix}0&1\\0&1\end{pmatrix},
\]
the boson has a longer wavelength and lower effective
momentum, consistent with a heavier or composite mediator.

\textbf{Conclusion:} The compression residual of a single fermion hop is purely
antisymmetric, has definite momentum, and pairs with its time-reverse into a
particle--antiparticle pair.
No interaction term was introduced; these properties follow from the codec
structure alone.

% ============================================================
\section{Experiment 3: Scaling}
\label{sec:exp3}

\textit{Python: scaling\_1.py, scaling\_2.py}

We scale to longer chains ($L = 4, 6, 8, 12, 16$) and more frames, testing
whether momentum quantisation $k = 2\pi m/L$ holds and whether the boson
properties depend on system size.

\subsection*{The key result --- analytically exact}

For any single fermion hop from site~$0$ to site~$d$, regardless of $L$ or $d$:
\[
  \psi = \frac{1}{\sqrt{2}}(e_0 + i\,e_d).
\]
The boson matrix has exactly two non-zero entries:
\[
  B_{0,d} = +\tfrac{i}{2}, \qquad B_{d,0} = -\tfrac{i}{2}.
\]
For the nearest-neighbour hop ($d=1$, $L=4$) this reads explicitly:
\[
  B \;=\;
  \begin{pmatrix}
    0 & +\tfrac{i}{2} & 0 & 0 \\[4pt]
    -\tfrac{i}{2} & 0 & 0 & 0 \\[4pt]
    0 & 0 & 0 & 0 \\[4pt]
    0 & 0 & 0 & 0
  \end{pmatrix}.
\]
The same $2\times 2$ antisymmetric block $\bigl(\begin{smallmatrix}0&+i/2\\-i/2&0\end{smallmatrix}\bigr)$
is embedded at positions $(0,d)$ and $(d,0)$ for any hop distance $d$ and
any chain length $L$; all other entries are zero.
The matrix is always purely antisymmetric, always has eigenvalues $\pm 1/2$,
and the phase of $B_{0,d}$ is always $-\pi/2$, giving:
\[
  k \cdot d = -\frac{\pi}{2}
  \quad\Longrightarrow\quad
  k = -\frac{\pi}{2d}.
\]
This is not a plane-wave momentum but a phase relationship between the two
fermion sites: the boson encodes exactly one number, the $\pi/2$ phase shift
between where the fermion was and where it went.
A quarter-turn in phase space is the defining property of a propagator.
This is what a virtual photon does in QED.

\textbf{Conclusion:} The compression residual of a single fermion hop is an
antisymmetric, purely imaginary, $\pm 1/2$-eigenvalued object carrying a
universal $\pi/2$ phase, independent of chain length and hop distance.

% ============================================================
\section{Experiment 4: Two-Fermion Hops}
\label{sec:exp4}

\textit{Python: two\_fermion\_hop.py}

We test whether the universal $\pi/2$ phase breaks for two fermions, and
whether the breaking has physical meaning.

\subsection*{Result 1: Partial survival of universality}

Every two-fermion boson matrix contains entries with phase $-\pi/2$,
inherited from the individual single-fermion hops.
It also acquires new phases $0$ and $+\pi/2$, corresponding to
inter-fermion correlations with no single-particle analogue.
The two-fermion boson is a superposition of single-hop bosons plus an
interaction residual.

\subsection*{Result 2: Path independence}

Fermions crossing (passing through each other) produce an identical boson
matrix to fermions moving in the same direction.
The codec records only initial and final pixel configurations, not
trajectories.
Path independence emerges for free.

\subsection*{Result 3: Pauli exclusion as zero compressibility}

Two fermions assigned to the same site produce a wavefunction $\psi \propto e_j$
(localised on a single site), for which $B = 0$ exactly.
The density matrix of a single-site state is
$\rho_{ik} = |\alpha|^2\,\delta_{ij}\delta_{kj}$,
which vanishes off-diagonal for $i \neq k$ since both $i = j$ and $k = j$
cannot hold simultaneously.
Double occupancy produces no boson: it has zero compression residual and is
informationally invisible.
Pauli exclusion is not a law imposed on the system; it is the statement that
double occupancy has nothing for the codec to encode.

% ============================================================
\section{Experiment 5: Pauli Exclusion and the Scope of the Encoding}
\label{sec:exp5}

\textit{Python: pauli\_as\_zero\_residual.py}

We verify that $B = 0$ for stationary fermions is exact and encoding-independent,
and clarify the scope of the single-fermion codec.

\subsection*{The encoding question}

When a single wavefunction is used to encode two stationary fermions at sites $i$
and $j$:
\[
  \psi = (e_i + e_j) + i(e_i + e_j) = (1+i)(e_i + e_j),
\]
the result is not localised on a single site.
This superposition has non-zero off-diagonal entries even with no motion, giving
$\|B\| \neq 0$.

\textbf{Conclusion:} The single-wavefunction codec is a \emph{one-fermion} codec: each wavefunction
encodes the state of one fermion across two time slices.
A system of $N$ fermions requires $N$ wavefunctions, one per particle.
Encoding two particles into one wavefunction conflates a two-particle
configuration with a single delocalised fermion.

\subsection*{Resolution}

Under the correct one-fermion-per-wavefunction encoding:
\begin{itemize}
  \item $B = 0$ for any stationary single fermion, for any phase or amplitude,
        across all chain lengths. This is analytically exact.
  \item $\|B\| = 1/\sqrt{2}$ for any single fermion hop, independent of hop
        distance and system size.
  \item $\|B\| = \sin(2\theta)/\sqrt{2}$ for a fermion in superposition
        $\psi(\theta) = \cos\theta\,e_i + i\sin\theta\,e_j$,
        where $\theta$ interpolates between stationary ($\theta = 0$) and
        mid-hop ($\theta = \pi/4$) and back to stationary ($\theta = \pi/2$).
\end{itemize}

The multi-fermion case requires a tensor product of single-fermion wavefunctions.
This is consistent with standard quantum mechanics and is not a limitation of the
framework.

\subsection*{Results}

\begin{center}
\begin{tabular}{lll}
\hline
\textbf{Configuration} & $\|B\|$ & \textbf{Reason} \\
\hline
Single fermion, stationary & $0$ (exact) & $\psi \propto e_j$, off-diagonal forced to zero \\
Single fermion, any hop & $1/\sqrt{2}$ (exact) & Universal, $-\pi/2$ phase \\
$\psi(\theta) = \cos\theta\,e_i + i\sin\theta\,e_j$ & $\sin(2\theta)/\sqrt{2}$ & Sharp onset, no threshold \\
\hline
\end{tabular}
\end{center}

% ============================================================
\section{Experiment 6: The $\theta$-Scaling Theorem}
\label{sec:exp6}

\textit{Python: theta\_scaling.py}

We verify the $\sin(2\theta)/\sqrt{2}$ formula analytically and confirm it
holds across all chain lengths and site separations.

\subsection*{Proof}

For $\psi(\theta) = \cos\theta\,e_i + i\sin\theta\,e_j$ (normalised, $i \neq j$),
the density matrix $\rho = |\psi\rangle\langle\psi|$ has the block structure
(showing only the $i,j$ subspace, all other entries zero):
\[
  \rho\big|_{ij} \;=\;
  \begin{pmatrix} \cos^2\theta & -\tfrac{i}{2}\sin 2\theta \\[4pt]
                  +\tfrac{i}{2}\sin 2\theta & \sin^2\theta \end{pmatrix}.
\]
Removing the diagonal gives the boson matrix restricted to the same subspace:
\[
  B\big|_{ij} \;=\;
  \begin{pmatrix} 0 & -\tfrac{i}{2}\sin 2\theta \\[4pt]
                  +\tfrac{i}{2}\sin 2\theta & 0 \end{pmatrix}.
\]
In full, the components are:
\begin{align*}
  \rho_{ii} &= \cos^2\theta, \quad
  \rho_{jj} = \sin^2\theta, \\
  \rho_{ij} &= \psi_i\psi_j^* = \cos\theta\cdot(-i\sin\theta) = -\tfrac{i}{2}\sin 2\theta, \\
  \rho_{ji} &= +\tfrac{i}{2}\sin 2\theta.
\end{align*}
All other entries of $\rho$ are zero since $\psi_k = 0$ for $k \neq i, j$.
Therefore:
\[
  \|B\|^2 = |B_{ij}|^2 + |B_{ji}|^2
  = 2\left(\frac{\sin 2\theta}{2}\right)^2
  = \frac{\sin^2 2\theta}{2},
\]
giving
\[
  \boxed{\|B(\theta)\| = \frac{\sin 2\theta}{\sqrt{2}}.}
\]
This proof uses only the definition $B = \rho - \mathrm{diag}(\rho)$ and the
normalisation of $\psi$.
It is independent of $L$, the separation $|i - j|$, and any overall phase
factor on $e_j$.
Numerical verification confirms agreement to within floating-point precision
($< 4 \times 10^{-16}$) for $L \in \{4, 8, 16, 32, 64, 128\}$ and all hop
distances.

\subsection*{Connection to Born rule}

Let $P = \sin^2\theta$ be the Born-rule probability that the fermion occupies
site~$j$.
Then:
\[
  \|B(\theta)\| = \frac{\sin 2\theta}{\sqrt{2}}
  = \frac{2\sin\theta\cos\theta}{\sqrt{2}}
  = \sqrt{2}\,\sqrt{P(1-P)}.
\]
$\|B\|$ is $\sqrt{2}$ times the geometric mean of the hop and no-hop
probabilities.
This is the interference term of the Born rule, made visible as a matrix norm.
The boson amplitude is not an approximation or an emergent average: it is the
exact quantum interference encoded in the density matrix.

\subsection*{Lifecycle of a virtual boson}

The formula $\|B(\theta)\| = \sin(2\theta)/\sqrt{2}$ traces the complete
lifecycle:
\begin{itemize}
  \item $\theta = 0$: fermion stationary, $\|B\| = 0$. No boson.
  \item $\theta = \pi/4$: fermion in equal superposition, $\|B\| = 1/\sqrt{2}$. Boson at maximum amplitude.
  \item $\theta = \pi/2$: fermion arrived, $\|B\| = 0$. Boson has vanished.
\end{itemize}
The boson is born when the fermion begins to move, peaks at mid-hop,
and vanishes when the fermion arrives.
This is the lifecycle of a virtual particle --- a propagator in the language of
quantum field theory, derived here without postulating field theory.

% ============================================================
\section{Experiment 7 (Exploratory): Mass, Momentum, and the Dispersion Relation}
\label{sec:exp7}

\textit{Python: dispersion\_poc.py}

Experiments 1--6 establish the $B$-matrix formalism entirely within a
\emph{massless, scale-free} regime: $\|B\| = 1/\sqrt{2}$ for any single hop,
independent of chain length $L$ or hop distance $d$ (\S\ref{sec:exp3}).
That scale invariance is itself informative: a rest mass sets a scale, so a
theorem that is exact and scale-free by construction has no room for mass to
enter as an internal parameter. Momentum, via the hop phase $k = -\pi/(2d)$,
is already present in the formalism; energy and mass are not, because nothing
in Experiments 1--6 plays the role of a generator of time evolution or of a
second internal degree of freedom at a single site. This experiment asks
whether the existing $B$-matrix machinery survives, unmodified, once such a
degree of freedom is added, and whether the relativistic dispersion relation
\[
  E^2 = (pc)^2 + (m_0c^2)^2
\]
can be recovered as a consequence rather than assumed.

\subsection*{Construction}

We introduce a second internal band at a single site (rather than a second
spatial site) and adopt the standard two-band lattice (Dirac-on-a-lattice
/ SSH-type) Hamiltonian
\[
  H(k) = m\,\sigma_z + t\sin(k)\,\sigma_x,
\]
where $m$ is an on-site band-splitting term and $t\sin(k)$ is a
momentum-dependent inter-band coupling. This Hamiltonian is diagonalised
\emph{numerically} (not assumed analytically) at each $k$ on a dense grid,
for several values of $m/t$. The lower-band eigenvector $v(k)$ is mapped onto
the existing Experiment~6 convention $\psi(\theta) = \cos\theta\,e_A +
i\sin\theta\,e_B$ by setting $\sin^2\theta$ equal to the numerically computed
Born-rule population $|v_B(k)|^2$ on the second band. The unmodified
Experiment~6 formula $B = \rho - \mathrm{diag}(\rho)$ is then applied to this
$\psi(\theta)$.

\subsection*{Results}

Four claims were checked numerically for $m \in \{0, 0.1, 0.5, 1, 2\}$ (with
$t=1$) over a dense grid $k \in (-\pi/2, \pi/2)$:

\begin{enumerate}
  \item \textbf{Eigenvalues.} The numerically diagonalised eigenvalues agree
        with $E(k) = \sqrt{m^2 + (t\sin k)^2}$ to machine precision
        ($< 10^{-15}$) at every $m$ and every $k$ tested.
  \item \textbf{Formalism survives unmodified.} The boson norm computed
        directly from the $2\times2$ matrix $B = \rho - \mathrm{diag}(\rho)$
        agrees with the closed-form Experiment~6 result
        $\sin(2\theta)/\sqrt{2}$ to machine precision
        ($< 3\times 10^{-16}$). No change to the boson-residual formalism
        was required to accommodate the reinterpretation of $\theta$ as a
        momentum/mass mixing angle rather than a spatial-hop angle.
  \item \textbf{Mass--momentum identity.} Working the eigenvector algebra of
        $H(k)$ through by hand gives
        \[
          \sin^2\theta\,(1-\sin^2\theta) = \frac{(t\sin k)^2}{4E(k)^2}
          \quad\Longrightarrow\quad
          \boxed{\|B(k)\| = \frac{|t\sin k|}{\sqrt{2}\,E(k)}.}
        \]
        This closed form agrees with the numerically computed boson norm to
        machine precision ($< 3\times 10^{-13}$) at every $m$ and $k$ tested.
        At $m = 0$ it reduces \emph{exactly} to the universal
        $\|B\| = 1/\sqrt{2}$ of Experiments 3--6, recovering the massless
        case as a special case rather than contradicting it. As $m$ grows
        relative to $t\sin k$, $\|B(k)\|$ shrinks monotonically toward zero:
        a more massive state produces a smaller compression residual at
        fixed momentum.
  \item \textbf{Continuum limit.} Identifying $t\sin k \to p c$ and $m \to
        m_0 c^2$ in the small-$k$ regime, $E(k)^2$ agrees with
        $(pc)^2 + (m_0c^2)^2$ to a relative error that shrinks as $k \to 0$
        (e.g.\ $\sim 7\times 10^{-4}$ over $|k|<0.05$ at $m=0$, tightening
        to $\sim 5\times 10^{-7}$ at $m=2$), consistent with the expected
        $O(k^2)$ lattice correction to the continuum dispersion relation.
\end{enumerate}

\subsection*{Status of this result}

This experiment shows that the $B$-matrix formalism of Papers~VI--VII is
\emph{compatible} with a mass term and reproduces the relativistic dispersion
relation as a continuum limit, once a second internal band and a Hamiltonian
are introduced. It does \emph{not} show that this construction is forced by
the framework's own first principles. Specifically:
\begin{itemize}
  \item The two-band Hamiltonian $H(k) = m\,\sigma_z + t\sin(k)\,\sigma_x$
        is imported from standard lattice/Dirac constructions (SSH-type
        models); it has not been derived from the spectral complexity
        functional $C_s$ or the cost functional $C_D + C_\psi + C_G$ used
        elsewhere in the series.
  \item In particular, it is not yet shown \emph{why} a minimum-description-length
        codec should prefer a two-band internal structure, nor why the
        band-splitting term should take the form $m\sigma_z$ specifically.
        The natural next step is to test whether the same overlap/exclusion
        cost functional that produces Pauli exclusion in Experiment~4
        (\S\ref{sec:exp4}), applied to internal (band) rather than spatial
        degrees of freedom, forces this structure -- or a different one.
  \item The identification $t\sin k \to pc$, $m \to m_0c^2$ is a relabelling
        of the lattice dispersion relation, not yet a statement about where
        the constant $c$ originates within the finite-bit-budget framework
        of Papers~I--V.
\end{itemize}
Consistent with the three-tier accounting adopted in Paper~XI, this result is
reported as a \emph{strongly suggested correspondence}, not an analytically
exact theorem in the sense of Experiments 1--6: it is numerically exact given
its assumptions, but its assumptions are not yet derived.

% ============================================================
\section{Core Results}
\label{sec:results}

Experiments 1--6 establish four results, each analytically exact:

\begin{enumerate}
  \item \textbf{Pauli exclusion.}
        $B = 0$ for any stationary fermion, for any encoding of the form
        $\psi \propto e_j$.
        Double occupancy is informationally invisible: zero compression
        residual, zero boson. \\
        \emph{Proof:} $\rho_{ik} = |\alpha|^2\delta_{ij}\delta_{kj} = 0$
        for $i \neq k$. \qed

  \item \textbf{Universal boson.}
        Every single fermion hop produces $\|B\| = 1/\sqrt{2}$, independent
        of chain length, hop distance, and encoding.
        The boson carries a universal $-\pi/2$ phase.

  \item \textbf{Born rule identity.}
        $\|B(\theta)\| = \sqrt{2\,P(1-P)}$, where $P = \sin^2\theta$ is the
        Born-rule hop probability.
        The boson amplitude is the interference term of Born rule.

  \item \textbf{Virtual propagator.}
        The boson amplitude follows $\sin(2\theta)/\sqrt{2}$ over the fermion's
        journey: born at $\theta = 0$, maximum at $\theta = \pi/4$, vanished at
        $\theta = \pi/2$.
        This is the propagator of a virtual particle, derived without field theory.
\end{enumerate}

All four results follow from the single definition $B = \rho - \mathrm{diag}(\rho)$
applied to the compressed wavefunction.
No interaction terms, coupling constants, or postulated particle species were
introduced.

Experiment~7 adds a fifth, exploratory result, reported at a different
evidentiary tier:

\begin{enumerate}
  \setcounter{enumi}{4}
  \item \textbf{Mass--momentum--boson identity (numerically verified,
        not yet derived from $C_s$).}
        Under the two-band extension of \S\ref{sec:exp7},
        $\|B(k)\| = |t\sin k|/(\sqrt{2}\,E(k))$, which reduces to the
        universal $1/\sqrt{2}$ of Result~2 at $m=0$ and reproduces
        $E^2 = (pc)^2+(m_0c^2)^2$ in the continuum limit.
\end{enumerate}

% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Quantum mechanics as an emergent description}

The results above suggest a coherent picture: quantum mechanics is not
a foundational layer of reality but the effective description an observer
constructs when observing fermions that are sampled from a compression codec.

The wavefunction is not a physical field propagating through space.
It is the codec --- the algorithm that selects which fermion configuration an
observer experiences, given the constraint that the observed path through the
$2^n$ configurations must be the one of minimum description length.
Measurement is decompression.
The Born rule is Solomonoff induction applied to a complex-valued codec:
the probability of outcome $j$ is proportional to the squared amplitude
$|\psi_j|^2$ because $|\psi_j|^2$ is the spectral weight assigned to that
configuration by the minimum-description-length codec.

Bosons, in this picture, are not fundamental.
They are the codec's record of fermion motion, reified into particles by
observers who do not know they are watching compressed data.
This does not make quantum field theory wrong.
It makes it an extraordinarily accurate effective description of compression
residuals.

\subsection{The $4\pi$ structure and spin}

The universal $-\pi/2$ phase of the single-hop boson corresponds to a
quarter-turn in the complex plane.
Four such hops return the phase to its starting point: $4 \times (-\pi/2) = -2\pi$.
This is the spinor structure of a spin-$1/2$ particle: a $2\pi$ rotation
returns a fermion to its original state only up to a sign, while a $4\pi$
rotation is the identity.
The $\pi/2$ phase quantisation is not imposed; it follows from the imaginary
encoding $\psi = f_0 + i f_1$ and the definition of the boson matrix.

\subsection{The encoding uniqueness question}

The framework relies on the complex exponential $e^{i\omega x}$ as the codec
basis.
Paper~VI \cite{meskanen2026vi} argues that this is the minimum-description-length
periodic smooth basis: the algorithm ($e^{i\omega x}$) is a fixed one-time cost
shared across all wavefunctions, while the per-instance cost is carried entirely
by the spectral modes (frequencies), which are unbounded.
This is what makes spectral complexity $C_s$ both computable and physically
meaningful.


\subsection{Relation to existing approaches}

Several existing programs seek to derive quantum mechanics from information
theory: reconstructions based on information-theoretic axioms
\cite{hardy2001, chiribella2011}, the Bayesian/QBist programme
\cite{fuchs2014}, and entropic dynamics \cite{caticha2019}.
The present approach differs in two respects.
First, it is constructive: it identifies a specific codec (complex spectral
decomposition), a specific complexity measure ($C_s$), and derives specific
quantitative predictions (the $\sin(2\theta)/\sqrt{2}$ theorem) rather than
axiomatising quantum mechanics from abstract information principles.
Second, it is embedded in the larger cosmological framework of Papers~I--V,
which derives spacetime geometry, the cosmological constant, and black hole
entropy from the same finite bit budget $n$.
Quantum mechanics is not postulated at the bottom of this hierarchy;
it emerges at the same level as geometry, from the same codec.

The two-band construction of \S\ref{sec:exp7} places the mass question in
the same territory as lattice-Dirac and SSH-type condensed-matter models,
where a relativistic dispersion relation emerges from a discrete lattice
Hamiltonian in the continuum limit. The present contribution is not the
lattice-to-continuum result itself, which is standard, but the observation
that the pre-existing $B$-matrix boson-residual formalism attaches to that
construction without modification, and assigns a compression-residual
interpretation to the resulting mass-dependent coupling.

\subsection{Limitations}

The present experiments use a one-dimensional chain of discrete sites.
The extension to three spatial dimensions, continuous space, and the full
Fock space of many-fermion systems requires tensor products of single-fermion
wavefunctions and a treatment of indistinguishability that goes beyond the
scope of this paper.
Additionally, the experiments demonstrate the emergence of a photon-like
boson from a scalar codec. Whether the model predicts other bosons, e.g. the
massive bosons ($W$, $Z$, Higgs) and of the gauge structure of the Standard Model
are not yet addressed.
The mass term introduced in \S\ref{sec:exp7} is imported from standard
lattice constructions rather than derived from $C_s$; deriving the two-band
structure and the specific form $m\sigma_z$ from the same minimum-description-length
principle used elsewhere in the series remains open (\S\ref{sec:future}).

% ============================================================
\section{Future Work}
\label{sec:future}

\subsection{Many-fermion systems and the Fock space}

The single-fermion codec must be extended to $N$-fermion systems via tensor
products.
The key question is whether the antisymmetry of the fermionic Fock space ---
the requirement that $\psi(x_1, x_2) = -\psi(x_2, x_1)$ --- emerges from
the boson matrix structure or must be imposed separately.
The antisymmetry of $B$ found in Experiment~2 suggests it may emerge
naturally.

\subsection{Deriving the mass term from $C_s$}

Experiment~7 shows that the boson-residual formalism is compatible with a
mass term once a two-band Hamiltonian is assumed, but the Hamiltonian itself
was imported rather than derived. The natural next step is to test whether
applying the same overlap/exclusion cost functional that produces Pauli
exclusion in Experiment~4 to an internal (band) degree of freedom, rather
than a spatial one, forces a band-splitting term of the form $m\sigma_z$ --
or produces a different structure entirely, which would itself be
informative. If successful, this would upgrade Result~5 from a numerically
verified correspondence to an analytically exact theorem on the same footing
as Results~1--4.

\subsection{Connection to the Friedmann equation}

Paper~VI identified the analytic derivation of the Friedmann equation from
the relational scale factor $R(t)$. Worldlines in GR are themselves compressible.
An elliptic orbit is a single Fourier mode --- the minimum-complexity trajectory
consistent with the conserved quantities of the system. PoC to show
these most probable path through the $2^n$ yield objects orbiting each
reproducing Keplerian dynamics without a force law.



% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have shown that a photon-like boson emerges from a purely fermionic system
compressed by a complex-valued spectral codec, without postulating interactions,
coupling constants, or particle species.
The central result is:
\[
  \|B(\theta)\| = \frac{\sin 2\theta}{\sqrt{2}} = \sqrt{2\,P(1-P)},
\]
where $P = \sin^2\theta$ is the Born-rule hop probability.
The boson amplitude is the interference term of the Born rule.
Pauli exclusion follows from the analytical result $B = 0$ for single-site
states.
The virtual propagator lifecycle --- born, peaked, vanished --- follows from
the $\theta$-dependence of $\|B\|$.

These results are analytically exact, encoding-independent, and hold for all
system sizes and hop distances.
They suggest that quantum mechanics is an emergent description: the story
an observer constructs to explain fermions that are sampled from a
minimum-description-length codec operating on a finite universe of $n$ bits.

We additionally reported an exploratory extension (Experiment~7) in which
the same boson-residual formalism, applied to a two-band lattice Hamiltonian,
reproduces the relativistic dispersion relation $E^2=(pc)^2+(m_0c^2)^2$ as a
continuum limit. This result is numerically exact given its assumptions, but
those assumptions -- a two-band internal structure and a specific
band-splitting term -- are not yet derived from the spectral complexity
functional $C_s$. We report it as a genuinely open problem, in keeping with
the three-tier accounting standard of distinguishing exact results, strongly
suggested correspondences, and open problems, rather than as a closed
derivation of relativistic mass from first principles.

% ============================================================
\section*{Supplementary Material}
\label{sec:supplementary}

Python experiments:

\begin{itemize}
  \item \texttt{wavefunction.py} --- Wavefunction class with spectral
        complexity $C_s$ and Solomonoff weight $2^{-C_s}$.
  \item \texttt{minimal\_video.py} --- Minimal $2\times 2$, 2-frame video:
        enumerate all 256 configurations, compute $C_s$ and boson signal.
  \item \texttt{boson\_structure.py}, \texttt{boson\_structure\_2.py} ---
        Symmetry, propagation, and quantisation analysis of the off-diagonal
        density matrix.
  \item \texttt{scaling\_1.py}, \texttt{scaling\_2.py} --- Longer chains
        ($L = 4$--$16$), more frames; momentum quantisation test.
  \item \texttt{two\_fermion\_hop.py} --- Two-fermion hops; phase universality,
        path independence, Pauli exclusion.
  \item \texttt{pauli\_as\_zero\_residual.py} --- Encoding scope and
        analytical proof of $B = 0$ for stationary fermions.
  \item \texttt{theta\_scaling.py} --- Numerical and analytical verification
        of $\|B(\theta)\| = \sin(2\theta)/\sqrt{2}$ across all $L$.
  \item \texttt{dispersion\_poc.py} --- Two-band lattice Hamiltonian
        $H(k) = m\sigma_z + t\sin(k)\sigma_x$; numerical diagonalisation;
        verification of $\|B(k)\| = |t\sin k|/(\sqrt{2}E(k))$ and of the
        continuum-limit recovery of $E^2=(pc)^2+(m_0c^2)^2$.
  \item \texttt{exclusion\_poc.py} --- Tests whether the same mechanism that produced Pauli exclusion for spatial
    sites in the experiment 4 also operates on the internal band degree of freedom introduced in experiment 7 -- and
    whether it can do any work toward explaining a mass splitting, without
    importing m by hand this time.
\end{itemize}

\printbibliography

\end{document}
