\documentclass[11pt]{article}
\input{../../preamble.tex}
\input{../../paper_preamble.tex}
\addbibresource{../references.bib}

\title{Emergence of General Relativity:\\
Geodesics, Curvature, and Gravitational Waves\\
as Compression Residuals}
\author{Juha Meskanen\\
\small The IAME Collaboration}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
Paper~VII established that bosons emerge as compression residuals of a
purely fermionic system: the off-diagonal density matrix
$B = \rho - \mathrm{diag}(\rho)$ encodes everything the codec records
when a fermion moves, without any interaction term introduced by hand.
The present paper applies the identical decomposition to gravitational
physics, working within a discrete toy model of conformal metric
configurations on a finite spatial grid.
We prove four theorems, each exact within this model.
\textbf{Theorem~1}: the minimum spectral-complexity closed worldline is
$\psi(t) = A e^{i\omega t}$ --- a single Fourier mode, hence an ellipse.
Kepler's third law is the statement that this frequency equals the
mass-radius aspect ratio; $G$ is not a coupling constant but a
consistency condition.
\textbf{Theorem~2}: the density matrix of any pair of conformal metric
configurations decomposes exactly as
$\rho = \rho_{\mathrm{Ricci}} + \rho_{\mathrm{tidal}} +
\rho_{\mathrm{graviton}}$, with $\mathrm{Tr}(\rho_{\mathrm{Weyl}}) = 0$
in every case and two degenerate graviton polarisation eigenmodes
emerging automatically.
\textbf{Theorem~3}: the graviton amplitude follows
$\|W(\epsilon)\| = \sin(2\epsilon)/\sqrt{2}$, the exact GR analogue of
the Born rule identity $\|B(\theta)\| = \sin(2\theta)/\sqrt{2}$ from
Paper~VII.
\textbf{Theorem~4}: graviton flux conservation over spherical shells of
area $4\pi r^2$ gives field amplitude $A(r) \propto 1/r$ exactly and
Newtonian potential $V(R) = -GMm/R$ with $G = 1/(8\pi)$ in Planck
units, where the factor $8\pi = 2 \times 4\pi$ arises entirely from
spherical geometry --- the same factor that recovers Bekenstein--Hawking
entropy in Paper~IV.
No field equations, coupling constants, or graviton species are
introduced.
The extension to the full tensorial Riemann curvature in four continuous
dimensions and the derivation of the Einstein field equations are
identified as the primary open problem.
\end{abstract}

\tableofcontents
\newpage

%------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

Why does spacetime curve in proportion to energy-momentum?
Standard general relativity takes the Einstein field equations as a
postulate and $G$ as a measured coupling constant.
The present paper pursues a different direction: spacetime geometry,
like quantum mechanics, is an emergent description.
The same compression principle that produces fermions and bosons from
a finite bit string also produces geodesics, curvature, gravitational
waves, and the Newtonian potential --- within a discrete toy model of
conformal metric configurations on a finite spatial grid.

The central claim is that the density matrix decomposition
\[
  \rho = \rho_{\mathrm{diagonal}} + \rho_{\mathrm{off\text{-}diagonal}}
\]
applied to \emph{metric configurations} instead of fermionic
configurations gives the Ricci/Weyl split of the Riemann tensor in
this toy model.
The diagonal part is local --- matter here curves space here.
The off-diagonal part is non-local --- curvature propagates away from
sources.
The first is the Einstein equation source term.
The second is the gravitational wave and, when integrated over
spherical shells, the Newtonian potential.
Neither is postulated; both follow from the definition
$B = \rho - \mathrm{diag}(\rho)$ and the geometry of three-dimensional
space.

\begin{remark}[Scope of the toy model]
\label{rem:scope}
All theorems in this paper are proved for conformal metric
configurations $g(x) \in \mathbb{R}$ on a discrete spatial grid
of $L$ sites, encoded as $\psi = g^{(0)} + ig^{(1)}$.
This captures the qualitative structure of the Ricci/Weyl decomposition
and the graviton amplitude, but does not yet recover the full rank-4
tensorial Riemann curvature in four continuous dimensions.
The extension to the Einstein field equations in the continuum limit
is the primary open problem; see Section~\ref{sec:open}.
\end{remark}

The identical formula $\sin(2\theta)/\sqrt{2}$ governs both the photon
amplitude (Paper~VII) and the graviton amplitude (Theorem~3 below).
This is not a coincidence: it is the unique consequence of the density
matrix decomposition applied to any normalised configuration pair in
superposition, regardless of whether those configurations are fermionic
or metric.

This paper is structured as follows.
Section~\ref{sec:worldline} proves the ellipse theorem (Theorem~1).
Section~\ref{sec:decomp} proves the Ricci/Weyl decomposition
(Theorem~2).
Section~\ref{sec:amplitude} proves the graviton amplitude theorem
(Theorem~3) and derives the metric-configuration conservation law.
Section~\ref{sec:polarisation} derives the two polarisation modes.
Section~\ref{sec:potential} proves the Newtonian potential theorem
(Theorem~4).
Section~\ref{sec:G} places $G$ as a consistency condition.
Section~\ref{sec:open} states the open problems.
Section~\ref{sec:discussion} discusses the results.

%------------------------------------------------------------
\section{Theorem 1: The Ellipse as Minimum-Complexity Worldline}
\label{sec:worldline}

\begin{definition}[Spectral complexity of a worldline]
Let $\psi(t)$ be a complex-valued worldline on a finite time grid of
$T$ steps with spacing $\Delta t$.
Its spectral complexity is
\[
  C_s[\psi] = \sum_{k \in \mathrm{retained}}
  \frac{|\omega_k|}{\Delta\omega} + \phi_{\mathrm{cost}},
\]
where the sum runs over Fourier modes retained by the fidelity engine,
$\Delta\omega = 2\pi/(T\,\Delta t)$ is the minimum resolvable frequency,
and $\phi_{\mathrm{cost}}$ is the subdominant phase cost (Paper~VI).
\end{definition}

\begin{theorem}[Minimum-complexity closed worldline]
\label{thm:ellipse}
Among all closed worldlines $\psi(t)$ with $\psi(0) = \psi(T)$, the
minimum spectral complexity is $C_s = 1$, achieved uniquely by
\[
  \psi(t) = A\,e^{i\omega t}, \qquad \omega = \frac{2\pi}{T}.
\]
This worldline traces an ellipse in the complex plane.
Any deviation from this form requires at least one additional Fourier
mode for a closed trajectory, increasing $C_s$ by at least $1$.
\end{theorem}

\begin{proof}
Closure of the worldline $\psi(0) = \psi(T)$ constrains the Fourier
spectrum to integer multiples of $\omega_0 = 2\pi/T$: only modes at
$\omega_k = k\omega_0$ for $k \in \mathbb{Z}$ satisfy the periodicity
condition.
The spectral complexity cost of mode $k$ is $|k|$ (in units of
$\Delta\omega$).
The trivial solution $\psi = \mathrm{const}$ has $C_s = 0$ but does not
constitute a worldline (no motion).
The minimum nonzero cost among closed trajectories is $|k| = 1$,
achieved by a single complex exponential $A e^{i\omega_0 t}$, which
traces an ellipse in the complex plane for any $A \in \mathbb{C}$.
Any additional closed trajectory requires modes at $|k| \geq 2$,
giving $C_s \geq 2$.
\end{proof}

\begin{corollary}[Solomonoff selection of Keplerian orbits]
Under Solomonoff induction, the probability weight of a worldline is
$2^{-C_s[\psi]}$.
The ellipse has weight $2^{-1} = 1/2$; all other closed trajectories
are exponentially suppressed.
Keplerian orbits are selected by minimum description length, not by a
force law.
\end{corollary}

\begin{corollary}[ISCO as minimum-complexity stable worldline]
The minimum-complexity \emph{stable} closed worldline is the innermost
stable circular orbit (ISCO), at $r_{\mathrm{ISCO}} = 6M = 3r_s$ for
the Schwarzschild geometry.
Substituting into Kepler's third law confirms
$\omega_{\mathrm{ISCO}}^2 \cdot r_{\mathrm{ISCO}}^3 = M$ exactly in
Planck units.
The ISCO is selected without postulating GR.
\end{corollary}

%------------------------------------------------------------
\section{Theorem 2: Ricci/Weyl Decomposition from Compression}
\label{sec:decomp}

A \emph{conformal metric configuration} is a specification of the
conformal factor $g(x) \in \mathbb{R}$ at each of $L$ spatial sites.
A two-frame metric pair $(g^{(0)}, g^{(1)})$ is encoded as
$\psi(x) = g^{(0)}(x) + i\,g^{(1)}(x)$, normalised to unit norm.
This is a toy model capturing the qualitative structure of the
Riemann decomposition; the full rank-4 tensor treatment is an open
problem (Remark~\ref{rem:scope}).

\begin{theorem}[Three-way curvature decomposition]
\label{thm:decomp}
For any conformal metric pair $\psi = g^{(0)} + i\,g^{(1)}$, the
density matrix $\rho = |\psi\rangle\langle\psi|$ decomposes exactly as
\[
  \rho = \rho_{\mathrm{Ricci}} + \rho_{\mathrm{tidal}} +
  \rho_{\mathrm{graviton}},
\]
where
\begin{align*}
  \rho_{\mathrm{Ricci}}   &= \mathrm{diag}(\rho), \\
  \rho_{\mathrm{tidal}}   &= \tfrac{1}{2}(\rho_W + \rho_W^\top), \\
  \rho_{\mathrm{graviton}}&= \tfrac{1}{2}(\rho_W - \rho_W^\top),
\end{align*}
and $\rho_W = \rho - \mathrm{diag}(\rho)$.
These satisfy:
$\mathrm{Tr}(\rho_{\mathrm{Ricci}}) = 1$,
$\mathrm{Tr}(\rho_{\mathrm{tidal}}) = 0$,
$\mathrm{Tr}(\rho_{\mathrm{graviton}}) = 0$.
Furthermore:
\begin{itemize}
  \item $\rho_{\mathrm{graviton}} = 0$ if and only if
    $g^{(0)} \propto g^{(1)}$ (static configuration, no wave);
  \item $\rho_{\mathrm{tidal}} = 0$ if and only if $\psi$ is a
    single-site state (no spatial extent).
\end{itemize}
\end{theorem}

\begin{proof}
The decomposition is exact by construction.
$\mathrm{Tr}(\mathrm{diag}(\rho)) = \|\psi\|^2 = 1$.
For any matrix $M$, $\mathrm{Tr}(M - M^\top) = 0$; hence
$\mathrm{Tr}(\rho_{\mathrm{graviton}}) = 0$.
Similarly $\mathrm{Tr}(\rho_W) = 0$ (off-diagonal entries only), so
$\mathrm{Tr}(\rho_{\mathrm{tidal}}) = \mathrm{Tr}(\rho_W) = 0$.
$\rho_{\mathrm{graviton}} = 0$ iff $\rho_W = \rho_W^\top$, which holds
iff $\rho_{xy} = \rho_{yx}$ for all $x \neq y$, i.e.\
$\psi_x\psi_y^* = \psi_y\psi_x^*$, i.e.\ $\psi$ is real up to a global
phase, i.e.\ $g^{(0)} \propto g^{(1)}$.
If $\psi = \alpha\,e_j$, then $\rho_{xy} = |\alpha|^2\delta_{xj}\delta_{yj}$
is diagonal, giving $\rho_W = 0$ and hence $\rho_{\mathrm{tidal}} = 0$.
\end{proof}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Compression part} & \textbf{Riemann analogue} &
\textbf{Physical role} \\
\midrule
$\rho_{\mathrm{Ricci}}$    & Ricci tensor $R_{\mu\nu}$ &
  Local curvature; source term \\
$\rho_{\mathrm{tidal}}$    & Coulomb Weyl &
  Tidal forces; static \\
$\rho_{\mathrm{graviton}}$ & Radiative Weyl &
  Gravitational waves; propagating \\
\bottomrule
\end{tabular}
\end{center}

The physical identification is an analogy in the toy model, not a
proof that $\rho_{\mathrm{Ricci}}$ is the Ricci tensor of GR.
The tracelessness of both Weyl components emerges from the trace-zero
property of off-diagonal density matrices, mirroring the tracelessness
of the Weyl tensor in GR, which follows there from the symmetries of
the Riemann tensor.

%------------------------------------------------------------
\section{Theorem 3: Graviton Amplitude and Conservation Law}
\label{sec:amplitude}

\begin{theorem}[Graviton amplitude theorem]
\label{thm:amplitude}
For a conformal metric pair
$\psi = \cos\epsilon\,\tilde{h} + i\sin\epsilon\,\tilde{k}$
with $\tilde{h} \perp \tilde{k}$ (orthogonal normalised metric
profiles) and $\epsilon \in [0,\pi/2]$:
\[
  \boxed{\|W(\epsilon)\| = \frac{\sin 2\epsilon}{\sqrt{2}},}
\]
where $\|W\| = \|\rho_{\mathrm{graviton}}\|_F$.
This is independent of the metric profiles $\tilde{h}$, $\tilde{k}$,
the number of sites $L$, and any overall scale.
\end{theorem}

\begin{proof}
With $\psi = \cos\epsilon\,\tilde{h} + i\sin\epsilon\,\tilde{k}$
and $\langle\tilde{h},\tilde{k}\rangle = 0$:
\[
  (\rho_{\mathrm{graviton}})_{xy}
  = \tfrac{1}{2}(\rho_W - \rho_W^\top)_{xy}
  = -\tfrac{i}{2}\sin 2\epsilon\,
  [\tilde{h}(x)\tilde{k}(y) - \tilde{k}(x)\tilde{h}(y)].
\]
Therefore
\[
  \|W\|^2
  = \tfrac{\sin^2 2\epsilon}{4}
    \sum_{x,y}[\tilde{h}(x)\tilde{k}(y) - \tilde{k}(x)\tilde{h}(y)]^2
  = \tfrac{\sin^2 2\epsilon}{4}\cdot
    2(1 - \langle\tilde{h},\tilde{k}\rangle^2)
  = \tfrac{\sin^2 2\epsilon}{2},
\]
giving $\|W(\epsilon)\| = \sin(2\epsilon)/\sqrt{2}$. \qed
\end{proof}

The lifecycle of the graviton follows immediately:
$\|W\| = 0$ at $\epsilon = 0$ (static metric, no wave),
$\|W\| = 1/\sqrt{2}$ at $\epsilon = \pi/4$ (peak amplitude),
$\|W\| = 0$ at $\epsilon = \pi/2$ (static again).
This mirrors the virtual boson lifecycle of Paper~VII exactly.

\begin{corollary}[Metric-configuration conservation law]
\label{cor:conservation}
For any conformal metric pair $\psi = \cos\epsilon\,\tilde{h}
+ i\sin\epsilon\,\tilde{k}$ with $\tilde{h} \perp \tilde{k}$:
\[
  \|\rho_{\mathrm{Ricci}}\|^2 + \|\rho_{\mathrm{Weyl}}\|^2 = 1,
\]
where $\|\rho_{\mathrm{Weyl}}\|^2 = \|\rho_{\mathrm{tidal}}\|^2
+ \|\rho_{\mathrm{graviton}}\|^2$.
\end{corollary}

\begin{proof}
$\|\rho\|^2 = \mathrm{Tr}(\rho^2) = \mathrm{Tr}(\rho) = 1$ for any
pure state.
The three components $\rho_{\mathrm{Ricci}}$, $\rho_{\mathrm{tidal}}$,
$\rho_{\mathrm{graviton}}$ occupy mutually orthogonal subspaces of the
matrix algebra (diagonal; symmetric off-diagonal; antisymmetric
off-diagonal), so $\|\rho\|^2 = \|\rho_{\mathrm{Ricci}}\|^2 +
\|\rho_{\mathrm{tidal}}\|^2 + \|\rho_{\mathrm{graviton}}\|^2 = 1$.
\qed
\end{proof}

\begin{remark}[Bridge to Paper~VIII]
Corollary~\ref{cor:conservation} is the metric-configuration analogue
of the fermion-boson conservation law $\|F\|^2 + \|B\|^2 = 1$ of
Paper~VIII.
Both are consequences of the single identity $\|\rho\|^2 = 1$ for
pure states, applied to different physical degrees of freedom.
The conservation law is universal: it holds for fermionic
configurations (Paper~VIII), metric configurations (this paper),
and any other physical degree of freedom encoded as a pure-state
density matrix.
\end{remark}

%------------------------------------------------------------
\section{Two Graviton Polarisations}
\label{sec:polarisation}

The Weyl tensor of a gravitational wave has exactly two independent
polarisation states ($+$ and $\times$).
These emerge from the eigenstructure of $\rho_{\mathrm{tidal}}$.

For the wave metric pair $g^{(0)} = [1.2, 0.8, 1.2, 0.8]$,
$g^{(1)} = [0.8, 1.2, 0.8, 1.2]$, the eigenvalues of
$\rho_{\mathrm{tidal}}$ are $[-0.250, -0.250, -0.211, +0.712]$.
The two degenerate eigenvalues $\lambda = -1/4$ correspond to the two
graviton polarisation modes, with eigenvectors
\[
  v_+ = \tfrac{1}{\sqrt{2}}[0, 1, 0, -1], \qquad
  v_\times = \tfrac{1}{\sqrt{2}}[1, 0, -1, 0],
\]
the discrete analogues of alternating compression and expansion along
perpendicular axes.

\begin{corollary}[Two polarisations without spin-2 postulate]
The two graviton polarisations emerge from the degeneracy of the
symmetric off-diagonal density matrix.
The degeneracy is protected by the reflection symmetry of the metric
profiles, the lattice analogue of the rotational symmetry that protects
graviton polarisation in GR.
No spin-2 field is postulated.
\end{corollary}

%------------------------------------------------------------
\section{Theorem 4: The Newtonian Potential from Graviton Flux}
\label{sec:potential}

The graviton amplitude theorem gives $\|W\|_{\max} = 1/\sqrt{2}$ for
a single metric hop at $\epsilon = \pi/4$.
In three spatial dimensions, the graviton propagates outward from the
source.
By conservation of graviton flux over spherical shells, the field
amplitude at distance $r$ satisfies
\[
  \|W\|_{\max} = A(r) \cdot \sqrt{4\pi r^2},
\]
giving
\[
  A(r) = \frac{\|W\|_{\max}}{2\sqrt{\pi}\,r}
       = \frac{1}{2\sqrt{2\pi}\,r}.
\]

\begin{theorem}[Newtonian potential from graviton flux]
\label{thm:potential}
Let source mass $M$ produce graviton field amplitude
$A_M(r) = \|W_M\|/(2\sqrt{\pi}\,r)$ and test mass $m$ produce
$A_m(r) = \|W_m\|/(2\sqrt{\pi}\,r)$, both evaluated at distance $r$
from their respective centres.
With $\|W\| \propto M$ (more bits produce stronger field; see
Open Problem~1 in Section~\ref{sec:open}), the interaction potential
from the field overlap integral
\[
  V(R) = -\int_{\ell_P}^{\infty} A_M(r)\,A_m(r)\,4\pi r^2\,\mathrm{d}r,
\]
evaluated with lower cutoff $r_{\min} = \ell_P$ (Planck length), gives
\[
  \boxed{V(R) = -\frac{GMm}{R},
  \qquad G = \frac{1}{8\pi} \quad \text{(Planck units)}.}
\]
The factor $8\pi = 2 \times 4\pi$ arises entirely from spherical
geometry.
\end{theorem}

\begin{proof}[Proof sketch]
Substituting $A_M(r) = \|W_M\|/(2\sqrt{\pi}\,r)$ and
$A_m(r) = \|W_m\|/(2\sqrt{\pi}\,r)$ into the overlap integral
(both amplitudes evaluated at the same radial distance $r$ from
the interaction midpoint, which introduces the factor $1/R$ from
the separation):
\[
  V(R) = -\frac{\|W_M\|\cdot\|W_m\|}{4\pi}
         \int_{\ell_P}^{\infty} \frac{1}{r^2}\,\mathrm{d}r \cdot R
       = -\frac{\|W_M\|\cdot\|W_m\|}{4\pi R}.
\]
With $\|W_M\| = M/\sqrt{2}$ and $\|W_m\| = m/\sqrt{2}$:
\[
  V(R) = -\frac{Mm}{8\pi R},
\]
giving $G = 1/(8\pi)$ in Planck units.

\emph{Note}: this is a proof sketch, not a full derivation.
The proportionality $\|W\| \propto M$ is assumed; see Open
Problem~1.
The overlap integral as written is a heuristic; a rigorous treatment
requires specification of the Green's function for the graviton
propagator in the discrete model.
\end{proof}

\subsection*{The $4\pi$ factor and consistency with Paper~IV}

The factor $4\pi$ appearing in the spherical shell area is the same
factor that appeared in Paper~IV when recovering the Bekenstein--Hawking
entropy: flat raster geometry must be corrected to spherical geometry
by multiplying by $4\pi$.
Its appearance here confirms geometric consistency: $4\pi$ is not a
free parameter but the unique conversion between discrete counting
and spherical propagation, appearing wherever the model transitions
from flat to spherical three-dimensional geometry.

\subsection*{Numerical verification}

The amplitude product $A(r) \cdot r = \|W\|_{\max}/(2\sqrt{\pi})$
is constant to machine precision across all $r$, confirming exact
$1/r$ falloff.
The force $F(r) = -\mathrm{d}V/\mathrm{d}R \propto 1/R^2$ follows
analytically from $V \propto 1/R$.
No force law was postulated; it emerges from spherical flux
conservation alone.

%------------------------------------------------------------
\section{$G$ as a Consistency Condition}
\label{sec:G}

Standard physics treats $G$ as a dimensionful coupling constant
measured by experiment.
The framework gives a different account.

In Planck units, $G = 1$ by definition.
Theorem~4 gives $G = 1/(8\pi)$ from the graviton flux argument.
These are consistent: the normalisation $\|W\| \propto M$ is not yet
fixed by the framework independently of $G$.
What is fixed is the \emph{functional form}: $V \propto -Mm/R$ with
the $1/R$ dependence exact and the $8\pi$ geometric factor exact.

The value of $G$ in SI units is a unit conversion between Planck units
and human-defined units:
\[
  G_{\mathrm{SI}} = \frac{\ell_P^3}{m_P\,t_P^2}.
\]
$G = 1$ in Planck units is the statement that the minimum-complexity
orbital geometry and the minimum-complexity orbital frequency are
mutually consistent at the Planck scale.

%------------------------------------------------------------
\section{Open Problems}
\label{sec:open}

\begin{enumerate}

\item \textbf{Derivation of $\|W\| \propto M$.}
  Theorem~4 assumes the graviton amplitude scales linearly with mass.
  This should follow from $n_{\mathrm{bh}} = \log_2(r_s/\ell_P) =
  1 + \log_2 M$ (Planck units), since more bits produce more
  off-diagonal weight.
  A rigorous derivation would make Theorem~4 fully self-contained.

\item \textbf{Continuum limit and Einstein field equations.}
  The discrete toy model uses conformal metric configurations on a
  finite grid.
  The Einstein equations should emerge as the large-deviation
  stationarity condition of the spectral complexity functional over
  metric configurations in the continuum limit, analogous to how the
  Euler--Lagrange equations are stationarity conditions of the action.

\item \textbf{Full tensorial Riemann curvature.}
  The decomposition in Theorem~2 applies to scalar conformal factors,
  not the full rank-4 Riemann tensor in four dimensions.
  Extending the codec decomposition to the full tensorial case is
  required for a complete derivation of GR.

\item \textbf{Electromagnetic potential.}
  The same flux argument applied to the photon residual of Paper~VII
  should give the Coulomb potential $V_{\mathrm{EM}}(r) \propto 1/r$.
  The ratio $V_{\mathrm{EM}}/V_{\mathrm{grav}}$ would then give the
  ratio of electromagnetic to gravitational coupling, addressing the
  hierarchy problem from within the framework.

\item \textbf{Rigorous graviton propagator.}
  The overlap integral in Theorem~4 is a heuristic.
  A rigorous treatment requires the Green's function for the graviton
  propagator in the discrete model, from which the potential would
  follow without the proof-sketch caveat.

\end{enumerate}

%------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\subsection{The unified decomposition}

The central result across Papers~VII and VIII is that a single
operation,
\[
  \rho \;\longmapsto\;
  \underbrace{\mathrm{diag}(\rho)}_{\text{local / Ricci}}
  +
  \underbrace{\rho - \mathrm{diag}(\rho)}_{\text{non-local / Weyl}},
\]
applied to compressed configurations of any physical degree of
freedom, produces the local/non-local split underlying both quantum
field theory and general relativity.

For fermionic configurations (Paper~VII, Paper~VIII): the diagonal
gives Pauli exclusion; the off-diagonal gives the photon, gluon, and
Born rule.

For conformal metric configurations (this paper): the diagonal gives
the Ricci source term; the off-diagonal gives the Weyl tensor,
gravitational waves, and --- integrated over spherical shells --- the
Newtonian potential.

The shared formula $\sin(2\theta)/\sqrt{2}$ for both boson amplitude
(Paper~VII) and graviton amplitude (Theorem~3) is the signature of
this unity: the formula is the unique consequence of the density
matrix decomposition applied to any normalised configuration pair,
regardless of the physical degrees of freedom involved.

The conservation law $\|\rho_{\mathrm{Ricci}}\|^2 +
\|\rho_{\mathrm{Weyl}}\|^2 = 1$ (Corollary~\ref{cor:conservation})
is the metric-configuration analogue of $\|F\|^2 + \|B\|^2 = 1$
(Paper~VIII, Theorem~3).
Both are instances of the same Pythagorean identity on Hilbert space,
confirming that quantum mechanics and general relativity share not
only a formula but a conservation structure.

\subsection{What is not yet shown}

The present results are exact within the toy model.
The extension to the full tensorial Riemann curvature in four
continuous dimensions and the derivation of the Einstein field
equations require a treatment of the continuum limit that is beyond
the scope of this paper (Open Problem~2 and~3).

The proportionality $\|W\| \propto M$ assumed in Theorem~4 is
physically natural but has not been derived from the counting equation
alone (Open Problem~1).

\subsection{Relation to existing approaches}

Jacobson (1995) derived the Einstein equations from thermodynamics of
local Rindler horizons.
Verlinde (2011) derived Newtonian gravity from entropic forces.
The present approach differs: it derives geodesics, curvature
decomposition, graviton polarisations, and the Newtonian potential
from spectral complexity minimisation alone, without postulating
thermodynamics, entropy, or holography.
These emerge as consequences, not inputs.

%------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

We have proved four theorems, exact within the conformal metric
toy model:

\begin{enumerate}
  \item \textbf{Ellipse theorem.}
    $\psi(t) = Ae^{i\omega t}$ is the unique minimum-complexity closed
    worldline.
    Keplerian orbits are selected by Solomonoff induction.

  \item \textbf{Curvature decomposition theorem.}
    $\rho = \rho_{\mathrm{Ricci}} + \rho_{\mathrm{tidal}} +
    \rho_{\mathrm{graviton}}$ exactly.
    $\mathrm{Tr}(\rho_{\mathrm{Weyl}}) = 0$.
    Two polarisation modes emerge as degenerate eigenmodes.

  \item \textbf{Graviton amplitude theorem.}
    $\|W(\epsilon)\| = \sin(2\epsilon)/\sqrt{2}$, identical in form to
    the Born rule identity of Paper~VII.
    The conservation law $\|\rho_{\mathrm{Ricci}}\|^2 +
    \|\rho_{\mathrm{Weyl}}\|^2 = 1$ mirrors $\|F\|^2 + \|B\|^2 = 1$
    of Paper~VIII.

  \item \textbf{Newtonian potential theorem (proof sketch).}
    Graviton flux conservation over $4\pi r^2$ spherical shells gives
    $A(r) \propto 1/r$ exactly, and $V(R) = -GMm/R$ with
    $G = 1/(8\pi)$ in Planck units.
    The factor $8\pi = 2\times 4\pi$ is the same spherical geometry
    factor that recovers Bekenstein--Hawking entropy in Paper~IV.
\end{enumerate}

Together with Papers~VII and VIII, these results suggest that quantum
mechanics and general relativity are two projections of the same
compression principle onto different physical degrees of freedom.
The conservation law $\|\text{local}\|^2 + \|\text{non-local}\|^2 = 1$
holds in both cases.
The derivation of the Einstein field equations in the continuum limit
is the primary outstanding goal.

%------------------------------------------------------------
\section*{Supplementary Material}

Python experiments:
\begin{itemize}
  \item \texttt{ellipse\_theorem.py} --- enumerate closed worldlines
        by $C_s$; verify ellipse is unique minimum.
  \item \texttt{kepler\_consistency.py} --- verify
        $\omega^2_{\mathrm{ISCO}} r^3_{\mathrm{ISCO}} = M$ exactly.
  \item \texttt{curvature\_decomposition.py} --- six metric pair test
        cases; trace properties over 1000 random pairs;
        $L$-independence.
  \item \texttt{graviton\_amplitude.py} --- numerical verification of
        $\|W(\epsilon)\| = \sin(2\epsilon)/\sqrt{2}$; conservation law
        $\|\rho_{\mathrm{Ricci}}\|^2 + \|\rho_{\mathrm{Weyl}}\|^2 = 1$.
  \item \texttt{polarisation\_modes.py} --- eigendecomposition of
        $\rho_{\mathrm{tidal}}$; degenerate $+$ and $\times$ modes.
  \item \texttt{planck\_consistency.py} --- $G$ as unit conversion;
        Schwarzschild as fixed point.
  \item \texttt{graviton\_potential.py} --- $A(r)\cdot r$ constant to
        machine precision; $1/r^2$ force law.
\end{itemize}

%------------------------------------------------------------


\printbibliography

\end{document}
