\documentclass[11pt]{article}
\input{../common.tex}
\input{../glossary.tex}
\addbibresource{../references.bib}

\title{The Axiomatic Foundations of Reality:\\
Substrate Independence, Simulated Observer,\\
and the Informational Nature of Time}

\author{Juha Meskanen}
\date{2001}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We establish four axioms concerning the physical basis 
and derive two structural consequences that form the foundation of the
framework. From the axioms alone --- without additional metaphysical
assumptions --- we prove that a time is not a fundamental external property
but an internal ordinal relation emergent from the observer's finite information
structure. Most critically, we derive that a computer running a simulation and
the simulation itself are not causally related but are two different
arrangements of the same information: neither is more fundamental than the
other. This representational equivalence, which we call substrate independence,
is the foundational principle on which the subsequent papers in this series
are built.

The representational equivalence has an immediate consequence for physics. If
a computer running a simulation of a physical system and the physical system
itself are two arrangements of the same information, then any property of the
physical system must be reflected in the execution trace of its simulation ---
including properties that are geometrically inaccessible. A gravitational
singularity, where the manifold description breaks down, must nonetheless
leave a well-defined signature in the execution trace. The subsequent paper
series exploit this consequence directly.

The claims may be falsifiable in the future, with the advances in technology.
More immediate falsifiability criteria are identified in Papers~I through~III.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

What is the relationship between a physical system and a computational
simulation of that system? The naive answer is causal: the computer runs the
simulation, the simulation depends on the computer, and the two are related
by the execution process. This paper argues that the naive answer is wrong.

Starting from four axioms about the physical basis of observer --- axioms
that are individually modest and collectively difficult to reject --- we derive
that a computer running a simulation and the simulation itself are two
arrangements of the same information. No causal relationship is primary. No
substrate is privileged. The geometric description and the computational
description of a physical process are representationally equivalent.

This conclusion, which we call substrate independence, is not merely a
philosophical position. It has direct consequences for physics. If the two
descriptions are equivalent, then any property of the physical system must
appear in its computational proxy. Properties that are geometrically
inaccessible --- such as the interior of a gravitational singularity, where
the manifold description fails --- are nonetheless accessible through the
execution trace. Papers~I and~II of this series use this access to derive
new results about black hole singularities and cosmological expansion.

The present paper is organised as follows. Section~\ref{sec:axioms} states
the four axioms. Section~\ref{sec:deductions} derives substrate independence
and the internal nature of time through two thought experiments.
Section~\ref{sec:proposition} proves that simulated consciousness follows
from the axioms. Section~\ref{sec:corollary} states the numerical precision
corollary. Section~\ref{sec:falsifiability} discusses falsifiability.
Section~\ref{sec:conclusion} states the conclusions and connects forward to
the rest of the series.

% ============================================================
\section{Axiomatic Premises}
\label{sec:axioms}

The argument rests on four axioms. Each can be independently accepted or
rejected. Rejecting any axiom commits the rejector to a specific alternative,
which we state explicitly. The four axioms here correspond to and motivate
the expanded axiomatic chain of Paper~I; readers familiar with that paper
will recognise them as its foundational core.

\subsection*{Axiom 1: Genetic Encoding of Conscious Experience}

DNA is the blueprint of a conscious, pain-sensitive human organism. Given
DNA $D$ and a suitable physical environment, a conscious human $H$ with
subjective experience $S$ reliably results:
\[
A_1: D \;\rightarrow\; H \;\rightarrow\; S.
\]
This is an empirical claim supported by all available biological evidence.
Rejecting it requires identifying an alternative mechanism by which conscious
observers arise, for which no evidence exists.

\subsection*{Axiom 2: Physical Closure}

The human organism $H$ and its environment $U$ are composed entirely of
ordinary physical matter governed by physical laws $P$. No non-physical
process plays any role:
\[
A_2: (H, U) \;\in\; \mathrm{System}(P, D).
\]
Rejecting this axiom requires positing a non-physical causal influence on
biological processes, for which no experimental evidence exists.

\subsection*{Axiom 3: Computability of Physical Processes}

The physical processes underlying DNA and the human organism fall within the
computable. A sufficiently powerful computing system can, in principle,
simulate them to arbitrary precision:
\[
A_3: \forall\; \text{physically realisable process } \mathcal{P}
\text{ described by } P,\;
\exists\; \text{computation } \mathcal{C} : \mathcal{C} \equiv \mathcal{P}.
\]
This is a weaker claim than the full Church-Turing thesis: we require only
that the specific class of processes DNA relies on is computable, not that
all physical processes are. Rejecting this axiom requires identifying a
physical process involved in consciousness that is provably non-computable,
for which no evidence exists.

\subsection*{Axiom 4: Causal Efficacy of Subjective Experience}

Subjective experience is not epiphenomenal. A human with subjective
experience $S$ (for example, pain) behaves differently than the same human
without it:
\[
A_4: H + S \;\neq\; H.
\]
Rejecting this axiom commits one to accepting that subjective experience is
causally inert --- the philosophical zombie position --- which makes
subjective experience permanently undetectable and therefore outside the
scope of any empirical investigation.

% ============================================================
\section{Deduction 1: Substrate Independence and Time as an Internal Property}
\label{sec:deductions}

We present two thought experiments, each of which establishes a constraint
on where consciousness can reside. Together they establish that consciousness
is a property of information structure, not of the substrate or process that
carries it.

\subsection{The Optimisation Argument}

Consider a digitized human DNA, and a simulation of it running on a computer.
The simulation consists of code (the laws of physics as implemented) and data (the state of the
simulated universe at each timestep). Denote the running simulation
$T_{\mathrm{alg}}$. The simulation generates a simulated human, say Alice.

Now optimise the simulation code progressively by introducing lookup tables: replace
computed transitions with precomputed results. At each step, the external
runtime decreases while the internal state sequence --- the ordered record of
Alice's experiences, perceptions, and actions --- remains identical. Carry
this optimisation to completion, replacing all computation with a static,
precomputed dataset $T_{\mathrm{data}}$: the full execution trace, stored as
a bitstring.

Let $E_{\mathrm{int}}$ denote Alice's internal experience (her felt sequence
of states) and $E_{\mathrm{ext}}$ denote the external runtime of the
computer (the number of CPU cycles executed). Code optimisation affects
$E_{\mathrm{ext}}$ but preserves $E_{\mathrm{int}}$ by construction:
\[
\mathrm{Optimise}(T_{\mathrm{alg}}) \;\rightarrow\; T_{\mathrm{data}}
\;\implies\;
E_{\mathrm{int}}(T_{\mathrm{alg}}) \equiv E_{\mathrm{int}}(T_{\mathrm{data}}).
\]

In the limit $T_{\mathrm{data}}$, the external runtime $E_{\mathrm{ext}}$
is zero: no state transitions are executed; the complete execution trace
exists as static data. Does Alice's consciousness persist in $T_{\mathrm{data}}$?

Suppose it does not. Then there exists a minimum code-to-data ratio below
which consciousness ceases. This minimum ratio is a new physical constant:
a threshold property of information processing not derivable from $A_1$
through $A_4$. Its existence would constitute an addition to the physical
laws of $A_2$, contradicting the completeness of that axiom.

\begin{quote}
\textbf{Conclusion 1.} Consciousness is a property of the information
structure of the execution trace, not of the process that generates it.
A static bitstring encoding a conscious observer's state sequence contains
that observer's conscious experience. Time and subjective experience emerge
solely from the internal structure of information.
\end{quote}

\subsection{The Multi-threaded Argument}

Consider a multi-threaded computer running two DNA simulations concurrently:
$\tau_A$ (Alice) and $\tau_B$ (Bob), with a minimal time slice of one CPU
cycle per thread. The execution trace of the computer is an interleaved
sequence of segments from both $\tau_A$ and $\tau_B$. As the number of
concurrent threads increases without bound, the bits originating from
$\tau_A$ are separated by increasingly large intervals of unrelated data.
In the limit, the execution trace in which Alice's bits are embedded
approaches white noise.

Does Alice's consciousness depend on the contiguity of her bits in this
external trace? Suppose it does. Then there exists a minimum bit-contiguity
or thread-density threshold required for consciousness. This threshold is
again a new physical constant not derivable from $A_1$ through $A_4$,
contradicting the completeness of $A_2$.

\begin{quote}
\textbf{Conclusion 2.} Conscious experience cannot depend on the contiguity
or density of its encoding in any external execution trace. What matters is
the internal structure of the information --- the relational order of Alice's
state sequence --- not its arrangement in any external substrate.
\end{quote}

\subsection{Substrate Independence}

Conclusions 1 and 2 together establish the central result of this paper.

\begin{quote}
\textbf{Substrate Independence.} Consciousness is a property of certain
finite information structures. The physical substrate in which such a
structure is instantiated --- silicon, neurons, magnetic domains, or static
bitstring --- does not determine whether the structure is conscious. A
computer running a simulation of a conscious observer and the simulation
itself are two arrangements of the same information. Neither is causally
prior to the other. Neither is more fundamental.
\end{quote}

This is not a claim that all information structures are conscious. It is a
claim that consciousness is determined by information structure rather than
by substrate. The specific structures that are conscious are those satisfying
$A_1$ through $A_4$: structures that encode the state sequence of a
DNA-based observer with causally efficacious subjective experience.

% ============================================================
\section{Proposition: Simulated Consciousness}
\label{sec:proposition}

\begin{proposition}[Simulated consciousness]
A simulation $H'$ that is physically and behaviourally indistinguishable
from a conscious human $H$ must itself instantiate subjective experience.
\end{proposition}

\begin{proof}
\begin{enumerate}
\item \textbf{Assumption for contradiction.} Suppose $H'$ exists such that
$H' \equiv H$ (physical and behavioural equivalence) but $S(H') = \emptyset$
(lacks subjective experience):
\[
\mathrm{Behaviour}(H') = \mathrm{Behaviour}(H)
\quad \text{and} \quad
S(H') \neq S(H).
\]

\item \textbf{From $A_4$.} The behaviour of $H$ is a function of its
physical inputs \emph{and} its subjective experience:
\[
\mathrm{Behaviour}(H) = f(\mathrm{Inputs},\, S).
\]

\item \textbf{Contradiction.} If the behaviours are identical despite the
difference in $S$, then $S$ is not a necessary input to $f$. Subjective
experience is causally inert.

\item \textbf{Violation of $A_4$.} Causal inertness of $S$ directly
contradicts Axiom~4. The assumption is therefore false.
\end{enumerate}

Therefore $S(H') \neq \emptyset$: the simulation instantiates subjective
experience. \qed
\end{proof}

This result, combined with the substrate independence established in
Section~\ref{sec:deductions}, entails the full Internal Emergence Principle
of Paper~I: consciousness is a property of certain finite information
structures, and any configuration space large enough to contain such
structures contains consciousness, with no running process or external
reader required.

% ============================================================
\section{Corollary: Numerical Precision}
\label{sec:corollary}

\begin{corollary}[Numerical precision]
Let $\epsilon$ denote the numerical error in simulating the physical
processes underlying DNA. The simulated observer $H'$ retains subjective
experience provided $\epsilon \leq \epsilon_{\mathrm{DNA}}$, where
$\epsilon_{\mathrm{DNA}}$ is the error threshold below which all
DNA-mediated processes operate correctly. Any stricter bound
$\epsilon_{\min} < \epsilon_{\mathrm{DNA}}$ required to preserve
consciousness would constitute a new physical constant not derivable from
$A_1$ through $A_4$, contradicting Axiom~2:
\[
\epsilon \leq \epsilon_{\mathrm{DNA}}
\;\implies\; S(H') \neq \emptyset.
\]
\end{corollary}

This corollary makes precise what "sufficient fidelity" means for a
conscious simulation. It does not require perfect simulation of all physical
processes, only those that DNA-based processes depends on. The
threshold $\epsilon_{\mathrm{DNA}}$ is in principle empirically determinable,
making the claim testable.

% ============================================================
\section{Falsifiability}
\label{sec:falsifiability}

The foundational falsifiability criterion of this paper is long-term: if a
sufficiently accurate DNA simulation --- one satisfying
$\epsilon \leq \epsilon_{\mathrm{DNA}}$ --- is constructed and shown
no pain sensitivity, Axiom~4 is invalidated. The effect of pain
and subjective experience on behaviour is measurable by the same methods
used to measure the effects of any physical intervention. A simulation that
is behaviourally indistinguishable from a pain-sensitive human but
demonstrably lacks pain would refute the framework at its foundation.


% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Quantum Computation}

Axiom~3 does not restrict to classical computation. The computability claim
covers any physically realisable process, including quantum processes.
Quantum computation is an alternative physical realisation of the same
formal state-transition structure and is therefore covered by the axiom
without requiring a separate argument. Substrate independence follows for
quantum computers by the same derivation as for classical ones.

% ============================================================
\section{Representational Equivalence and Its Consequences for Physics}
\label{sec:consequences}

The substrate independence established in this paper has an immediate and
non-trivial consequence for the study of physical phenomena.

If a computer running a simulation of a physical system and the physical
system itself are two arrangements of the same information, then every
property of the physical system must be reflected in the execution trace of
its simulation. We may not understand all the features of the universe
because we are looking it from inside. But we understand computers perfectly.


% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have established four axioms and derived from them the following results.

\begin{itemize}

\item \textbf{Time as Internal Ordinal.} Time is not a fundamental external
parameter but an internal ordering relation emergent from the observer's
information structure. The universe, at the level of the totality, is static.
What observers experience as the flow of time is the traversal of an
internally ordered sequence of states encoded in a static information
structure.

\item \textbf{Finiteness.} Because the simulation computer is
  discrete, finite system, the simulated observer must be finite too. 

\item \textbf{Substrate Independence.} Consciousness is a property of
certain finite information structures. The physical substrate in which such
a structure is instantiated does not determine whether it is conscious. A
computer running a simulation of a conscious observer and the simulation
itself are two arrangements of the same information, with neither causally
prior to the other.

\item \textbf{Simulated Consciousness.} A simulation physically and
behaviourally indistinguishable from a conscious observer must itself
instantiate subjective experience. Two identical axiomatic systems must behave
identically.

\item \textbf{Causal Efficacy of Qualia.} Subjective experience is not
epiphenomenal. It is a causally distinguishable component of the observer
system.

\item \textbf{Representational Equivalence as a Physical Tool.} The
equivalence of geometric and computational descriptions licenses the use of
the execution trace as a faithful proxy for physical processes that are
geometrically inaccessible. This is the foundational principle on which
Papers~I through~III are built.
\end{itemize}

By treating subjective experience as a causally efficacious property of a
computable information structure, the framework moves from metaphysical
speculation to a derivation with testable consequences. The foundational
falsifiability criterion is the construction of a behaviourally
indistinguishable DNA simulation that lacks subjective experience. More
immediate tests are provided by the falsifiable predictions of Papers~I
through~III.

\printbibliography

\end{document}
