\documentclass[12pt]{article}
\input{../../common.tex}
\addbibresource{../../references.bib} 


\title{Supplementary Material II: Perception, Memory, and the Informational Basis of Mortality}

\date{2024}

\begin{document}

\maketitle

\section*{Abstract}
We extend the theory of the Algorithmic Universe to the life-cycle of the intelligent observer. We demonstrate that perception is a relational alignment within a static ensemble rather than signal processing, and that mortality is a necessary consequence of algorithmic suppression (MDL) as internal complexity increases.

\section{Perception as Relational Alignment}
In a static configuration space $\Omega$, the classical concept of "signal transmission" (e.g., a photon traveling from source to detector) is reinterpreted as a relational property of the bitstring.
An observer does not "receive" information from an external source. Instead, an observer exists in a state $\mathbf{s} \in \Omega$ where the internal bit-patterns of the observer and the external environment are already correlated. Perception is the experience of this internal algorithmic consistency. 

\section{Memory and the Execution Trace}
Memory is defined as the accumulation of the \textbf{execution trace}—the internal record of the sequence of configurations traversed through the Superspace. 
\begin{itemize}
    \item \textbf{Subjective Past:} The set of bit-states already integrated into the observer-motif.
    \item \textbf{Subjective Future:} The set of potential configurations in $\Omega$ that are algorithmically consistent with the current trace and the Minimal Description Length (MDL) constraint.
\end{itemize}

\section{The Informational Theory of Mortality}
Within the Algorithmic Selection Principle (ASP), mortality arises not from biological decay, but from \textbf{vanishing probability measure}. 

Let $\sigma_t$ be the observer-motif at index $t$. As $t$ increases, $\sigma_t$ incorporates more information (memories, structural growth), increasing its Kolmogorov complexity $K(\sigma_t)$. Because the probability of a state's existence in $\Omega$ is governed by $P(\sigma) \approx 2^{-K(\sigma)}$, the number of consistent "continuations" for a highly complex observer eventually reaches zero. 



\begin{equation}
    \exists \, t_{max} \text{ such that } \{ \mathbf{s} \in \Omega : \sigma_{t_{max}} \subset \mathbf{s} \} \to \emptyset
\end{equation}

Death is thus the "eviction" of the observer from the static ensemble due to the inability of the $2^n$ configurations to support a motif of such high specificity.

\section{Conclusion}
This framework provides a non-biological derivation of life and death. Mortality is an inevitable byproduct of existence within a finite informational ensemble governed by algorithmic probability.

\end{document}
