\documentclass[11pt]{article}
\input{../../common.tex}
\input{../../glossary.tex}
\addbibresource{../../references.bib} 


\title{Base Pillars}

\date{2024}

\begin{document}
\maketitle


\section{The Ladder of Emergence}

\subsection*{Ladder 1: The Pre-Reality}
Reality begins as abstract information itself, with no privileged representation, alphabet, ordering, geometry, or probability measure.
Any concrete encoding (binary, ternary, continuous, or otherwise) is a matter of convenience rather than ontology.
Entropy, time, and physical law arise only relative to observer-defined representations and coarse-grainings.
There is no distinguished substrate or execution mechanism; existence is purely informational.

\subsection*{Ladder 2: The Observer Exists}
Since $\mathbf{P}$ excludes nothing, it necessarily contains finite informational subsets capable of self-reference.
The existence of structured, observer-containing worlds requires no external cause; such structures are unavoidable elements of infinite informational potential.
Observers are not observing the reality outside—they are sub-sets of it.

\subsection*{Ladder 3: The Static Multiverse}
A concrete world may be represented as a finite configuration space
\[
\mathcal{C} \cong \{0,1\}^n,
\]
where each configuration $\mathbf{s} \in \mathcal{C}$ is a static informational state.
There is no motion or dynamics at this level—only the coexistence of all configurations.
The binary representation is not fundamental; it is chosen purely for convenience.

Zero and One (yes, no) are not fundamental either. We might equally well write:
\[
\mathcal{C} \cong \{Head,Tail\}^n,
\]

All configurations are ontologically equal.

\subsection*{Ladder 4: The Emergence of Time}
Time is an induced ordering arising from traversals through configuration space.
An ordering of states is not given a priori; it is defined only relative to observer paths.
Timeless configuration space thus precedes experienced temporality.

\subsection*{Ladder 5: The Observer Filter}
An observer history is a finite path through the configuration space.
\[
\gamma = (s_1, s_2, \dots), \qquad s_i \in \mathcal{C}.
\]
Only a vanishing subset of all possible paths form the observer-compatible set
\[
\mathcal{T}_{\mathrm{obs}} \subset \mathcal{S},
\]
where $\mathcal{S}$ denotes the space of all paths through $\mathcal{C}$.
Only paths in $\mathcal{T}_{\mathrm{obs}}$ possess a subjective ``inside'' and are experienced as reality.

\subsection*{Ladder 6: The Spectral Selection Principle (SSP)}
Observer paths are not equally weighted.
Their measure is governed by compressibility in a spectral representation.
Paths admitting efficient spectral encodings dominate the ensemble of experienced histories.
Formally, preferred histories minimize spectral description length:
\[
H_{\mathrm{SSD}} = \arg\min_H \mathcal{L}(H),
\]
where $\mathcal{L}(H)$ denotes the minimal spectral encoding length of path $H$.
This statistical dominance yields an emergent, finite, and predictable physics without invoking fundamental dynamics.

\subsection*{Ladder 7: The Wavefunction as Optimal Encoding}
The quantum wavefunction $\Psi$ is not a physical field but an optimal spectral compression of the observer-path ensemble.
Superposition and interference reflect shared substructure within compressed descriptions.
The Born Rule emerges as a measure over paths:
\[
P(\gamma) \propto 2^{-\mathcal{L}(\gamma)} \sim |\Psi(\gamma)|^2.
\]
Thus probability is not fundamental; it is a consequence of informational weighting.


\subsection*{Ladder 8 --- Emergent Gravitation: Spectral Compression Geodesics}

Physical gravitation emerges as a macroscopic manifestation of the statistical weighting of observer-compatible paths in configuration space.

\paragraph{1. Observer paths and spectral measure:}  
Let $\mathcal{C}$ be the static configuration space, and $\mathcal{T}_{\mathrm{obs}} \subset \mathcal{S}$ the set of observer-compatible paths $\gamma = (s_1, s_2, \dots)$.
Each path is assigned a spectral encoding length $\mathcal{L}(\gamma)$ and a corresponding quantum amplitude:
\[
\Psi(\gamma) \propto 2^{-\mathcal{L}(\gamma)/2}, \qquad P(\gamma) = |\Psi(\gamma)|^2 \propto 2^{-\mathcal{L}(\gamma)}.
\]

The dominant paths are those minimizing their spectral encoding length:
\[
\delta \int_\gamma \mathcal{L}(\text{state})\, d\lambda = 0.
\]
These paths correspond to the classical trajectories experienced by observers.


Let $\rho(\mathbf{s})$ denote the density of reusable microstructure motifs within configuration space.  
High-density regions correspond to low spectral cost; an observer path naturally ``falls'' toward them.  

In the continuum limit, the effective metric $g_{\mu\nu}$ is defined by the gradient of spectral cost:
\[
\mathcal{L}(\gamma) \approx \int_\gamma \sqrt{g_{\mu\nu} \, \dot{x}^\mu \dot{x}^\nu} \, d\lambda,
\]
so that minimizing $\mathcal{L}$ coincides with minimizing proper length in spacetime — i.e., geodesics.

\paragraph{4. Connection to Quantum Theory:}  
Summation over all $\gamma \in \mathcal{T}_{\mathrm{obs}}$ weighted by $2^{-\mathcal{L}(\gamma)}$ reproduces a path-integral-like formalism:
\[
\langle \mathcal{O} \rangle = \sum_{\gamma \in \mathcal{T}_{\mathrm{obs}}} \mathcal{O}(\gamma) \, 2^{-\mathcal{L}(\gamma)},
\]
where $\mathcal{O}$ is any observable depending on path configurations.  

- \textbf{Everettian view:} All observer-compatible paths exist simultaneously, no branching; probability arises from spectral measure, not collapse.  
- \textbf{Wheeler–DeWitt view:} Configuration space is timeless; time emerges as an ordering along $\gamma$.  
- \textbf{Classical limit:} Low-$\mathcal{L}$ paths dominate, producing smooth geodesic motion as in GR.

Gravitation is thus not a fundamental force but a statistical-geometric phenomenon:  
\emph{classical geodesics are the trajectories of minimal spectral encoding length through configuration space, weighted by observer-compatible path measure.}  

\end{document}
