\documentclass[11pt]{article}
\input{../common.tex}
\input{../glossary.tex}
\addbibresource{../references.bib} 

\title{Pauli Exclusion as Emergent Antisymmetry via MDL}
\author{Juha Meskanen}
\date{2025}

\begin{document}

\maketitle

\begin{abstract}
TODO
\end{abstract}

\noindent \textbf{Keywords:} Spinor, Spectral Complexity, Minimal Description Length

\tableofcontents
\bigskip

\section{Pauli Exclusion as Emergent Antisymmetry via MDL}

\subsection{Motivation}

We began with the goal of constructing a minimal, observer-centric framework in which physical laws are not imposed but \emph{emerge} from information-theoretic principles. The initial attempt relied on hardcoding fermionic behavior, specifically Pauli's exclusion principle, through antisymmetric wavefunctions (e.g., Slater determinants).

While this approach reproduced desired phenomenology, it violated the guiding principle: \emph{all structure should emerge from a compression principle rather than be assumed}.


\subsection{Initial Assumptions and Limitations}

The initial model assumed:
\begin{itemize}
    \item Point-like particles with positional degrees of freedom.
    \item Gaussian wavepackets representing localized "blobs."
    \item Explicit antisymmetry to enforce fermionic behavior.
\end{itemize}

These assumptions were sufficient to produce stable, inertial-like dynamics, but they imposed:
\begin{itemize}
    \item Hard constraints (e.g., exclusion) rather than emergent behavior.
    \item A fixed particle ontology rather than a representation-based one.
\end{itemize}

\subsection{Shift in Perspective: Observer-Centric MDL}

We replaced the ontological perspective with an \emph{observer-centric Minimum Description Length (MDL)} framework:

\begin{itemize}
    \item The fundamental object is not a particle, but a \emph{description of observations}.
    \item The observer selects representations that minimize encoding cost.
\end{itemize}

We identified three key principles:

\begin{enumerate}
    \item \textbf{Observer existence:} Observers correspond to consistent sequences of encoded states.
    \item \textbf{Compression:} Descriptions are minimized in spectral complexity (low-frequency dominance).
    \item \textbf{Symmetry:} Equivalent configurations must not be redundantly distinguished.
\end{enumerate}

\subsection{Key Insight: Compression is a Collective Property}

A critical realization was that:

\begin{quote}
\emph{Compression is not a property of individual objects, but of the entire configuration.}
\end{quote}

This implies:
\begin{itemize}
    \item Physical statistics (bosonic vs fermionic) cannot be assigned locally.
    \item Instead, they emerge from global encoding constraints.
\end{itemize}

\subsection{The Trade-Off: Compression vs. Distinguishability}

We identified two competing encoding strategies:

\begin{enumerate}
    \item \textbf{Symmetric (bosonic) encoding:}
    \[
        A + B
    \]
    \begin{itemize}
        \item Minimizes spectral complexity.
        \item Merges identities, increasing ambiguity.
    \end{itemize}

    \item \textbf{Antisymmetric (fermionic) encoding:}
    \[
        A - B
    \]
    \begin{itemize}
        \item Preserves distinguishability.
        \item Introduces structural cost (higher frequencies).
    \end{itemize}
\end{enumerate}

\subsection{Ambiguity as a Cost Function}

We introduced a new term into the MDL framework:

\[
\text{Total Cost} = C_{\text{spectral}} + \lambda \cdot C_{\text{ambiguity}}
\]

where:
\begin{itemize}
    \item $C_{\text{spectral}}$ measures the frequency-weighted energy of the signal.
    \item $C_{\text{ambiguity}}$ penalizes indistinguishability between configurations.
\end{itemize}

This ambiguity cost becomes large when two states overlap and cannot be uniquely reconstructed.

\subsection{Emergence of Pauli Exclusion}

With this modified cost function, we observe:

\begin{itemize}
    \item When particles are well separated:
    \begin{itemize}
        \item Symmetric encoding is cheaper.
        \item Bosonic behavior emerges.
    \end{itemize}

    \item When particles overlap:
    \begin{itemize}
        \item Ambiguity cost diverges.
        \item Antisymmetric encoding becomes cheaper.
        \item Effective exclusion emerges.
    \end{itemize}
\end{itemize}

Thus:

\begin{quote}
\emph{Pauli exclusion is not a fundamental rule, but a consequence of minimizing description length under a distinguishability constraint.}
\end{quote}

\subsection{Role of Gaussian States}

Gaussian wavepackets naturally arise because they minimize joint uncertainty in position and frequency domains. They are therefore:

\begin{itemize}
    \item The optimal localized representation under spectral compression.
    \item The natural "atoms" of the observer’s encoding.
\end{itemize}

\subsection{Spinors as Minimal Representations}

Spinors were reinterpreted not as fundamental physical entities, but as:

\begin{quote}
\emph{minimal representations that encode rotational and exchange symmetries in a compressed form.}
\end{quote}

Their defining feature:
\begin{itemize}
    \item A $2\pi$ rotation introduces a sign change.
    \item This sign structure enables antisymmetric encoding.
\end{itemize}

Thus, spin-$\tfrac{1}{2}$ behavior is not assumed, but emerges as the minimal structure supporting antisymmetric distinguishability.

\subsection{Phase Transition Perspective}

The physical universe is hypothesized to exist at a critical point:

\begin{itemize}
    \item Pure bosonic regime: maximal compression, but loss of information.
    \item Pure fermionic regime: maximal distinguishability, but rigid structure.
\end{itemize}

The observed universe corresponds to a balance between these extremes:

\begin{quote}
\emph{A phase transition between compression and distinguishability.}
\end{quote}


\subsection{Conclusion}

The key result of this progression is the replacement of hardcoded physical laws with emergent constraints:

\begin{itemize}
    \item Pauli exclusion emerges from an ambiguity penalty.
    \item Bosons and fermions correspond to different optimal encoding strategies.
    \item Spinors arise as minimal representations of symmetry under compression.
\end{itemize}

This leads to a unified view:

\begin{quote}
\emph{Physics is the minimal description of observer-consistent data under competing pressures of compression and distinguishability.}
\end{quote}


\end{document}
