\documentclass[11pt]{article}
\input{../../common.tex}
\input{../../glossary.tex}
\addbibresource{../../references.bib}

\title{Planck Scale and Spectral Gravity}
\author{Juha Meskanen}

\begin{document}
\maketitle


\section{The Planck Scale as the Nyquist Frequency}  
In signal processing, the Nyquist frequency \(f_\mathrm{Nyquist} = f_s/2\) is the highest frequency that can be resolved given a sampling rate \(f_s\). Frequencies above \(f_\mathrm{Nyquist}\) are aliased and indistinguishable.  

In the Abstract Universe:

\begin{itemize}
    \item \textbf{Sampling Rate:} The rate of bit-flips along observer-compatible paths in configuration space \(\mathcal{C}\).  
    \item \textbf{Planck Limit:} The highest frequency resolvable in the spectral wavefunction \(\Psi(\gamma)\).  
    \item \textbf{Consequence:} Structures smaller than the Planck length or evolving faster than the Planck time cannot be resolved by \(\Psi\); they are mathematically unrepresentable and effectively non-existent for any observer.
\end{itemize}

\section{Discrete Time from Spectral Updates}  
Let \(\gamma = (s_1, s_2, \dots)\) be an observer path through \(\mathcal{C}\), with each \(s_i\) an \(n\)-bit configuration.  

Define a spectral encoding length \(\mathcal{L}(\gamma)\). Updates to \(\mathcal{L}\) occur only when significant motifs (observer-relevant patterns) change along the path.  

\[
t_{i+1} - t_i = 
\begin{cases}
1, & \text{if a motif update occurs} \\
0, & \text{otherwise}
\end{cases}
\]

Thus, \textbf{time emerges discretely}, with “ticks” corresponding to spectral transitions along observer-compatible paths.  

\section{Spectral Density and the Inverse-Square Law}  
Let a “mass” be a localized cluster of reusable microstructure motifs in configuration space. The number of configurations at distance \(r\) that can reference these motifs grows with the surface area of a 3D sphere:

\[
A(r) = 4 \pi r^2
\]

Consequently, the density of spectral-compatible configurations falls as:

\[
\rho(r) \propto \frac{1}{r^2} \, L(d)
\]

where \(L(d)\) is the log-normal distribution of motif distances \(d\).  

Observer paths minimize spectral cost \(\mathcal{L}(\gamma)\) and are biased toward regions of high \(\rho(r)\). In the continuum limit, the variational principle

\[
\delta \int_\gamma \mathcal{L}(\text{state}) \, ds = 0
\]

reproduces classical geodesics in 3D space, giving rise to an effective inverse-square law of interaction.  

\section{Connection to Quantum Gravity and Path Integrals}  
Summing over all observer-compatible paths weighted by their spectral amplitude:

\[
\langle \mathcal{O} \rangle = \sum_{\gamma \in \mathcal{T}_{\mathrm{obs}}} \mathcal{O}(\gamma) \, 2^{-\mathcal{L}(\gamma)}
\]

reproduces a path-integral formalism. In this framework:

\begin{itemize}
    \item \textbf{Everettian view:} All observer-compatible paths exist simultaneously; probabilities emerge from spectral weighting.  
    \item \textbf{Wheeler–DeWitt view:} Configuration space is timeless; time is induced along the observer path \(\gamma\).  
    \item \textbf{Classical limit:} Low-\(\mathcal{L}\) paths dominate, yielding smooth geodesics in agreement with General Relativity.
\end{itemize}

\section{Summary}  
The Planck scale, discrete time, and gravity emerge naturally from:

\begin{enumerate}
    \item The maximal resolvable spectral frequency (Nyquist-Planck analogy).  
    \item Discrete spectral updates along observer paths.  
    \item Statistical bias of observer paths toward high-density microstructure regions, producing effective inverse-square laws.  
\end{enumerate}

In this view, gravitation is not a fundamental force but a statistical-geometric phenomenon rooted in the spectral structure of observer-compatible paths in configuration space.

\printbibliography
\end{document}
