\section*{Changelog}

This page tracks significant changes across the various draft releases.

\begin{description}

\iftoggle{theology}{
\item[\textbf{2026-05-14}]
  Added the $E=mc^2$ example to Chapter~\nameref{chap:reality} to provide a
  minimal illustration of the problems that arise when attempting to model
  ``reality.''
  
  Added Chapter~\nameref{chap:wdw} regarding quantum cosmology.
  
  Improved typesetting, internal cross-referencing, and overall
  \LaTeX\ document structure.
  
\item[\textbf{2026-05-17}]
  Thought experiments in Chapter~\nameref{chap:the-problem-with-time} 
  were completed with exact definitions of Boltzmann entropy and page-footer references~\cite{page1983clock}.

\item[\textbf{2026-05-24}]
  Fixed an inconsistency bug in the equations within Chapter~\nameref{chap:framework-of-everything} and refactored
  the chapter. Added the ``guitar string'' analogy.

  The theory has finally name (or more like a placeholder for name): \iame, or \iameverything.
  Added a derivation section to Chapter~\nameref{chap:framework-of-everything}.
  Added a fourth requirement to \nameref{chap:definition-of-foundational-theories}: finding a name for the theory. How
  ridiculously difficult can it get these days, when all imaginable names have already been taken?

\item[\textbf{2026-05-26}]

  The Chapter~\nameref{chap:the-problem-with-time} expanded with known
  supporting the static timeless view.

\item[\textbf{2026-05-29}]
  Support for profiling to allow different versions of this book from a single master source.
}{
}

\item[\textbf{2026-05-31}]
  UML diagrams included with the Chapter~\nameref{chap:implementing-qm}
  and the Chapter~\nameref{chap:implementing-gr}

  Pandoc 3.1.3’s built-in HTML template injecting references to polyfill.io to serve malicious code fixed.

\item[\textbf{2026-06-02}]
  
  AI injected nonsese purged from the chapters \ref{chap:humans-as-axiomatic-systems} and \ref{chap:framework-of-everything}.
  

\item[\textbf{2026-06-06}]
  
  Added \href{https://meskanen.com/surveys/surveys.html}{Take the Survey: I am, am I?} link to the \nameref{chap:closure} chapter.


\item[\textbf{2026-06-07}]
  
  References to the three fundamental papers included.


\end{description}



\newpage
