\documentclass[11pt]{article}
\input{../../preamble.tex}
\input{../../paper_preamble.tex}
\addbibresource{../references.bib}

\title{Towards a Zero-Parameter Theory of Everything:\\
Existence, Compression, and the Emergence of Physical Law}

\author{Juha Meskanen}
\date{2026}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We sketch a programme in which both quantum mechanics and general
relativity emerge from a single principle: \emph{observers exist, and
their probability of existence is determined by the compressibility of
their description}.
The central object is a Boltzmann-like probability measure over
histories $\gamma$ compatible with an observer $O$:
\[
\mathbb{P}(\gamma \mid O)
= \frac{1}{Z_O}\exp\!\bigl(-\mathcal{C}_O[\gamma]\bigr),
\qquad \gamma \in \Gamma_O,
\]
where $\mathcal{C}_O[\gamma]$ is the total description cost of history
$\gamma$ given observer $O$.
Rather than deriving this from first principles --- a task left for
future work --- we collect here the evidence that it is the correct
organising equation.
Papers~VII and~VIII of this series establish exact, analytically
verified results: the off-diagonal density matrix of a purely fermionic
system yields a photon-like boson with amplitude
$\|B(\theta)\| = \sin(2\theta)/\sqrt{2}$; the identical decomposition
applied to metric configurations yields the Ricci/Weyl split,
gravitational wave polarisations, and a Newtonian potential
$V(r) = -Mm/(8\pi r)$ in Planck units.
These results share a single structure:
$\rho \mapsto \mathrm{diag}(\rho) + (\rho - \mathrm{diag}(\rho))$,
the local/non-local split of any compressed configuration.
We conjecture that this structure forces the
cost functional to decompose into three regimes --- discrete (\textbf{D}),
spectral ($\boldsymbol{\psi}$), and geometric (\textbf{G}) --- corresponding
to quantised spacetime, quantum mechanics, and general relativity
respectively.
The value $n \approx 184$ bits for the universe's information content,
derived independently from inflationary cosmology and black hole
thermodynamics in Papers~IV and~V, is expected to emerge as the
saddle point of the observer's own existence probability, though the
explicit computation remains open.
This paper is a progress report and a research agenda, not a
completed derivation.
\end{abstract}

\textbf{Keywords:} theory of everything, Solomonoff induction, spectral
complexity, quantum gravity, information theory, observer self-selection,
Kolmogorov complexity, Wheeler--DeWitt, De Sitter, Bekenstein--Hawking

% ============================================================
\section{Introduction}
\label{sec:intro}

General Relativity and quantum mechanics each require an input the other
cannot supply.
General Relativity takes the geometry of spacetime as given and derives
how matter curves it.
Quantum mechanics takes the Hilbert space structure of states as given
and derives how they evolve.
Neither theory explains why its own input exists.

The programme pursued across this series of papers begins from an
observation: both inputs may be consequences of a single integer $n$,
the total bit count of a finite, static, timeless universe, together
with the rule that structures dominate the measure in proportion to how
compressibly they can be described.
Papers~I--VIII \cite{meskanen2001, meskanen2002, meskanen2003,
meskanen2026iv, meskanen2026v, meskanen2026vi, meskanen2026vii,
meskanen2026viii} develop this programme in steps: establishing the
observer axiom, deriving the aspect ratio of spacetime, recovering
Bekenstein--Hawking entropy, matching the Friedmann equation, and
most recently proving exact results about the emergence of bosons and
gravitational structure from the density matrix of a compressed
fermionic system.

The present paper attempts to draw these threads together into a
unified picture.
It is written frankly as a sketch: the exact results of Papers~VII
and~VIII are solid, but the claim that they fit into a single
zero-parameter framework is a conjecture whose full justification
remains to be constructed.
We state what is proved, what is strongly suggested, and what is open.

% ============================================================
\section{The Organising Equation}
\label{sec:principle}

\subsection{The Observer as the First Principle}

We take as the starting point:

\begin{quote}
\textbf{Axiom (Observer Existence).} There exists at least one observer.
\end{quote}

An observer is defined operationally, following Paper~I
\cite{meskanen2001}: a finite information structure whose state sequence
encodes causally efficacious subjective experience.
Paper~I establishes from four modest axioms that such a structure must
be finite, substrate independent, and that the universe containing it
must be static and timeless.
We take those results as working hypotheses and build on them here.

\subsection{Probability of Existence via Solomonoff Induction}

Given the observer axiom, which universe does the observer find herself
in?
By substrate independence, the observer is a finite information
structure and the universe is the minimal information structure that
contains her.
The natural prior over universes is then determined by description
length:

\begin{equation}
\mathbb{P}(\gamma \mid O)
= \frac{1}{Z_O}\exp\!\bigl(-\mathcal{C}_O[\gamma]\bigr),
\qquad \gamma \in \Gamma_O,
\label{eq:master}
\end{equation}

where $\Gamma_O$ is the set of all histories compatible with the
existence of $O$, $\mathcal{C}_O[\gamma]$ is the total description cost
of $\gamma$ given $O$, and $Z_O = \sum_{\gamma \in \Gamma_O}
\exp(-\mathcal{C}_O[\gamma])$.
This is Solomonoff induction \cite{solomonoff1964} in path-integral
form.

The apparent free parameter $\lambda$ (an inverse temperature in the
Boltzmann analogy) has been set to unity: rescaling $\lambda$ is
equivalent to rescaling $\mathcal{C}_O$, and only the ordering of
histories by cost is physically meaningful.
$\lambda$ is a units choice, not a physical constant.

\subsection{Physical Law as Large-Deviation Minimiser}

The conjecture is that the observed regularities of physics are the
histories $\gamma^*$ that minimise $\mathcal{C}_O[\gamma]$ subject to
observer compatibility:
\[
\gamma^* = \arg\min_{\gamma \in \Gamma_O} \mathcal{C}_O[\gamma].
\]

A universe that is chaotic or structureless is not forbidden by this
equation; it simply has a higher description cost and contributes
negligibly to $Z_O$.
We do not assume finiteness; we expect to derive it, because finite
universes compress better than infinite ones.
We do not assume order; we expect to derive it, because ordered
histories are shorter to describe.
Whether equation~\eqref{eq:master} alone is sufficient to force these
properties in full generality, without additional structure, is an open
question.

% ============================================================
\section{Evidence from Papers VII and VIII}
\label{sec:evidence}

The strongest evidence for the programme comes from two recent exact
results, which we summarise here.
Both concern the same mathematical object: the off-diagonal density
matrix $B = \rho - \mathrm{diag}(\rho)$ of a compressed configuration.

\subsection{Bosons as Compression Residuals (Paper VII)}

Paper~VII \cite{meskanen2026vii} encodes two-frame fermion
configurations as complex wavefunctions $\psi = f_0 + if_1$ and
computes the density matrix $\rho = |\psi\rangle\langle\psi|$.
The off-diagonal part $B$ captures everything the codec records when a
fermion moves between frames.

Four results are established analytically and verified numerically:

\begin{enumerate}
  \item \textbf{Pauli exclusion.}
        For any stationary single fermion, $B = 0$ exactly.
        Double occupancy produces no compression residual; it is
        informationally invisible.
        Pauli exclusion is not a postulate but the statement that
        double occupancy has nothing for the codec to encode.

  \item \textbf{Universal boson.}
        Every single fermion hop produces $\|B\| = 1/\sqrt{2}$,
        independent of chain length, hop distance, and encoding.
        The residual carries a universal $-\pi/2$ phase, the defining
        signature of a propagator.

  \item \textbf{Born rule identity.}
        For a fermion in superposition
        $\psi(\theta) = \cos\theta\,e_i + i\sin\theta\,e_j$,
        \begin{equation}
          \|B(\theta)\| = \frac{\sin 2\theta}{\sqrt{2}}
                        = \sqrt{2\,P(1-P)},
          \label{eq:born}
        \end{equation}
        where $P = \sin^2\theta$ is the Born-rule hop probability.
        The boson amplitude is the interference term of the Born rule,
        expressed as a matrix norm.

  \item \textbf{Virtual propagator lifecycle.}
        $\|B\| = 0$ when the fermion is stationary, rises to
        $1/\sqrt{2}$ at equal superposition, and returns to $0$ when
        the hop is complete --- the lifecycle of a virtual particle,
        derived without field theory.
\end{enumerate}

These results follow from the single definition
$B = \rho - \mathrm{diag}(\rho)$ and the normalisation of $\psi$.
No interaction terms, coupling constants, or particle species were
introduced.

\subsection{Gravitational Structure as Compression Residual (Paper VIII)}

Paper~VIII \cite{meskanen2026viii} applies the identical decomposition
to metric configurations: a two-frame metric pair
$(g^{(0)}, g^{(1)})$ is encoded as $\psi(x) = g^{(0)}(x) + ig^{(1)}(x)$.
The same $\rho \mapsto \mathrm{diag}(\rho) + B$ split then yields:

\begin{enumerate}
  \item \textbf{Curvature decomposition.}
        $\rho$ decomposes exactly as
        $\rho_{\mathrm{Ricci}} + \rho_{\mathrm{tidal}} +
        \rho_{\mathrm{graviton}}$,
        with $\mathrm{Tr}(\rho_{\mathrm{Weyl}}) = 0$ in every case.
        The diagonal is identified with local Ricci curvature (the
        Einstein source term); the symmetric off-diagonal with tidal
        (Coulomb Weyl) forces; the antisymmetric off-diagonal with
        gravitational waves.

  \item \textbf{Two polarisations.}
        The two degenerate graviton polarisation modes emerge as
        eigenvectors of $\rho_{\mathrm{tidal}}$ with eigenvalue
        $-1/4$, protected by the reflection symmetry of the metric
        profiles.
        No spin-2 field is postulated.

  \item \textbf{Graviton amplitude theorem.}
        For orthogonal metric profiles at mixing angle $\epsilon$,
        \begin{equation}
          \|W(\epsilon)\| = \frac{\sin 2\epsilon}{\sqrt{2}},
          \label{eq:graviton}
        \end{equation}
        the exact gravitational analogue of equation~\eqref{eq:born}.
        The formula is independent of metric profiles, system size,
        and overall scale.

  \item \textbf{Newtonian potential.}
        Graviton flux conservation over spherical shells gives
        $A(r) \propto 1/r$ exactly, and the overlap integral yields
        \begin{equation}
          V(R) = -\frac{GMm}{R},
          \qquad G = \frac{1}{8\pi} \quad \text{(Planck units)}.
          \label{eq:newton}
        \end{equation}
        The factor $8\pi = 2 \times 4\pi$ arises from spherical
        geometry alone --- the same factor that recovers
        Bekenstein--Hawking entropy in Paper~IV
        \cite{meskanen2026iv}.

  \item \textbf{Minimum-complexity orbits.}
        The minimum spectral-complexity closed worldline is
        $\psi(t) = Ae^{i\omega t}$ --- a single Fourier mode, tracing
        an ellipse.
        Under Solomonoff induction, all other closed trajectories are
        exponentially suppressed.
        Keplerian orbits are selected by minimum description length,
        not by a force law.
\end{enumerate}

\subsection{The Shared Structure}

Equations~\eqref{eq:born} and~\eqref{eq:graviton} are identical in
form.
This is not a coincidence of presentation.
Both follow from the density matrix decomposition applied to a
normalised superposition of two orthogonal configurations.
The formula $\sin(2\alpha)/\sqrt{2}$ is the unique consequence of
this structure, regardless of whether $\alpha$ parametrises a fermion
in quantum superposition or a metric in gravitational superposition.

The conjecture is that this shared structure reflects a shared origin:
quantum field theory and general relativity are two projections of the
same compression principle onto different physical degrees of freedom.

% ============================================================
\section{The Conjectured D-$\psi$-G Trinity}
\label{sec:trinity}

\subsection{Three Regimes of the Cost Functional}

The exact results of Papers~VII and~VIII suggest, but do not yet
establish, that the cost functional $\mathcal{C}_O[\gamma]$ naturally
decomposes into three regimes:

\begin{equation}
\mathcal{C}_O[\gamma]
= \underbrace{C_D[n]}_{\text{bit budget}}
+ \underbrace{C_\psi[\psi_\gamma]}_{\text{spectral cost}}
+ \underbrace{C_G[g_\gamma]}_{\text{geometric cost}}.
\label{eq:decomp}
\end{equation}

The intuition is as follows.
Any observer requires well-defined boundaries.
The cost of specifying those boundaries depends on the representation:

\begin{itemize}
  \item \textbf{D} (Discrete): Boundaries are subsets of a bitstring.
        Cheap but structureless.
  \item \textbf{$\psi$} (Spectral): Boundaries are subspaces of a
        function space.
        The spectral cost $C_\psi$ suppresses high-frequency
        (rough) wavefunctions, producing the wave-like behaviour
        observed at quantum scales.
  \item \textbf{G} (Geometric): Boundaries are surfaces in a metric
        space.
        Closed surfaces in three dimensions admit compact,
        low-frequency descriptions.
        The geometric cost $C_G$ is minimised for smooth manifolds.
\end{itemize}

The claim that these three regimes are \emph{forced} by the boundary
structure of any self-describing observer --- rather than chosen by
hand --- is the key conjecture of this paper.
A derivation of the decomposition~\eqref{eq:decomp} from
equation~\eqref{eq:master} alone has not yet been carried out.

\subsection{Scale Dependence}

Under this conjecture, the three regimes dominate at different scales:

\begin{center}
\begin{tabular}{lll}
\hline
\textbf{Scale} & \textbf{Dominant cost} & \textbf{Emergent description} \\
\hline
Planck & $C_D[n]$ & Discrete, quantised spacetime \\
Quantum & $C_\psi[\psi_\gamma]$ & Quantum mechanics, Born rule \\
Observer & $C_G[g_\gamma]$ & General Relativity, Friedmann \\
\hline
\end{tabular}
\end{center}

The evidence that $C_\psi$ produces quantum mechanical structure is
solid (Paper~VII).
The evidence that $C_G$ produces gravitational structure is solid for
toy models (Paper~VIII).
The claim that the crossover between $C_\psi$ and $C_G$ corresponds
to the Planck scale, and that $C_D$ governs the discrete regime, is
physically motivated but not yet derived.

% ============================================================
\section{The Emergence of $n$}
\label{sec:n}

\subsection{$n$ as a Candidate Fixed Point}

The bit count $n$ appears in equation~\eqref{eq:decomp} as
$C_D[n] = \log_2 n$.
We conjecture that $n$ is not a free parameter but the saddle point
of the observer's own existence probability.

The probability that observer $O$ exists at all, marginalising over
all histories at bit count $n$, is:
\[
\mathbb{P}(O \mid n)
= \frac{Z_O(n)}{Z(n)}.
\]
The prior over $n$ is $\mathbb{P}(n) \propto 1/n$ (description length
of $n$ itself).
The joint probability is:
\[
\mathbb{P}(O, n)
\propto \frac{Z_O(n)}{Z(n) \cdot n}.
\]
The saddle-point condition --- the value of $n$ that maximises the
observer's probability of existence --- is formally:
\begin{equation}
\frac{\partial}{\partial n}
\Bigl[\log Z_O(n) - \log Z(n) - \log n\Bigr] = 0.
\label{eq:saddle}
\end{equation}

\subsection{Consistency with Independent Estimates}

Papers~IV and~V \cite{meskanen2026iv, meskanen2026v} derive a
preferred value $n \approx 184$ by two independent routes:
matching the inflationary aspect ratio and recovering
Bekenstein--Hawking entropy for a solar-mass black hole.
If equation~\eqref{eq:saddle} independently selects $n^* \approx 184$,
the three routes converge, which would constitute strong evidence
for the framework.

That convergence has not yet been verified.
The saddle-point equation requires an explicit model of $Z_O(n)$ ---
the partition function of observer-compatible histories as a function
of $n$ --- for which no tractable expression is currently available.
This is the central open problem of the programme.

% ============================================================
\section{Summary of Recovered Physics}
\label{sec:recovery}

We summarise what has been established, what is strongly suggested,
and what remains open.

\paragraph{Established (exact, verified).}
\begin{itemize}
  \item Pauli exclusion as zero compression residual (Paper~VII).
  \item Boson amplitude $\|B(\theta)\| = \sin(2\theta)/\sqrt{2}$,
        independent of system size (Paper~VII).
  \item Born rule as interference term of the density matrix (Paper~VII).
  \item Ricci/Weyl decomposition from the diagonal/off-diagonal split
        of metric density matrices (Paper~VIII).
  \item Two graviton polarisations as degenerate eigenmodes
        (Paper~VIII).
  \item Graviton amplitude $\|W(\epsilon)\| = \sin(2\epsilon)/\sqrt{2}$
        (Paper~VIII).
  \item Newtonian $1/r$ potential from graviton flux conservation,
        with $G = 1/(8\pi)$ in Planck units (Paper~VIII).
  \item Keplerian orbits as minimum-complexity closed worldlines
        (Paper~VIII).
\end{itemize}

\paragraph{Strongly suggested (structural agreement, not yet derived).}
\begin{itemize}
  \item The aspect ratio $\mathcal{A}(n) = 2^n/(n\ln n)$ matching
        inflationary expansion in Planck units (Papers~IV--V).
  \item The $\Lambda$CDM expansion profile recovered from the lognormal
        emergence law without a cosmological constant (Paper~V).
  \item The Friedmann equation under the identification of matter as
        bound bit structures (Paper~V).
  \item The proportionality $\|W\| \propto M$ needed to make
        Theorem~4 of Paper~VIII fully self-contained.
\end{itemize}

\paragraph{Open.}
\begin{itemize}
  \item Solving the saddle-point equation~\eqref{eq:saddle} and
        recovering $n^* \approx 184$.
  \item Deriving the decomposition~\eqref{eq:decomp} from
        equation~\eqref{eq:master} without additional assumptions.
  \item The continuum limit: recovering the full Einstein field
        equations from the discrete toy model of Paper~VIII.
  \item The emergence of energy: how $E \propto \hbar\omega$ arises
        from the spectral cost $C_\psi$ in physical units.
  \item Massive bosons ($W$, $Z$, Higgs) and the gauge structure of
        the Standard Model.
  \item The correct topology: why the boundary is a 3-sphere rather
        than some other manifold.
\end{itemize}

% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Why the Same Formula Governs Bosons and Gravitons}

The most striking result across Papers~VII and~VIII is that the formula
$\sin(2\alpha)/\sqrt{2}$ appears in both the boson amplitude and the
graviton amplitude, for entirely different physical degrees of freedom.
The explanation is structural: both follow from the off-diagonal norm
of a rank-one density matrix formed from a superposition of two
orthogonal unit vectors.
For a state $\psi = \cos\alpha\,u + i\sin\alpha\,v$ with $u \perp v$,
\[
  \|\rho - \mathrm{diag}(\rho)\|_F = \frac{|\sin 2\alpha|}{\sqrt{2}}
\]
is an algebraic identity, independent of what $u$ and $v$ represent.

This universality suggests that the density matrix decomposition is not
a feature of quantum mechanics or of gravity separately but of the
compression principle itself.
Whether this observation can be elevated to a theorem --- a proof that
any two-regime compression of normalised configurations yields this
formula --- is an open question.

\subsection{The Role of $4\pi$}

The factor $4\pi$ appears in three independent calculations in this
series: recovering Bekenstein--Hawking entropy (Paper~IV), recovering
the Newtonian potential (Paper~VIII Theorem~4), and the aspect ratio
between discrete and spherical geometry (Papers~IV--V).
In each case it is not a free parameter but the geometric conversion
between flat counting and spherical propagation.
Its consistent appearance is encouraging but not yet understood at the
level of the cost functional.

\subsection{What This Framework Is and Is Not}

The programme is not a completed theory.
It is a collection of exact results in discrete toy models, a set of
structural correspondences with known physics, and a conjectured
organising equation.
The toy models use 1D chains, conformal metrics, and two-frame
sequences.
Extensions to four continuous dimensions, the full tensorial Riemann
curvature, and the Fock space of many-fermion systems remain to be
carried out.

The framework does not yet contain a derivation of the Standard Model,
does not address the origin of energy in physical units, and does not
derive the masses of elementary particles.
These are long-term targets, not current achievements.

What the framework does suggest --- and what the results of Papers~VII
and~VIII make concrete --- is that the separation between quantum
mechanics and general relativity may be an artifact of working in
different effective descriptions of the same underlying compression
principle.
The density matrix decomposition $\rho \mapsto \mathrm{diag}(\rho) + B$
produces local/non-local structure whether applied to fermion
configurations or metric configurations.
The quantum/gravitational duality, in this view, is not a coincidence
to be explained but a consequence to be proved.

\subsection{Relation to Existing Approaches}

Jacobson \cite{jacobson1995} derived the Einstein equations from
thermodynamics of local Rindler horizons.
Verlinde \cite{verlinde2011} derived Newtonian gravity from entropic
forces.
The present approach derives geodesics, curvature decomposition,
graviton polarisations, and the Newtonian potential from spectral
complexity minimisation alone, without postulating thermodynamics,
entropy, or holography as inputs.
Whether entropy and thermodynamics emerge as consequences within the
framework is an interesting open question.

Reconstructions of quantum mechanics from information-theoretic axioms
\cite{hardy2001, chiribella2011} and the Bayesian/QBist programme
\cite{fuchs2014} seek to derive the formalism of quantum mechanics
without a physical model of the underlying degrees of freedom.
The present approach is complementary: it proposes a specific physical
model (fermions as pixels, bosons as compression residuals) and derives
the quantum formalism from it, but does not yet have a proof that the
model forces the full Hilbert space structure.

% ============================================================
\section{Open Problems}
\label{sec:open}

We list the primary open problems in order of foundational priority.

\begin{enumerate}
  \item \textbf{The saddle-point equation.}
        Solving equation~\eqref{eq:saddle} requires an explicit model
        of $Z_O(n)$.
        A tractable approximation to the observer partition function,
        even in a toy model, would be a significant step.

  \item \textbf{Emergence of energy.}
        The spectral cost $C_\psi$ is an information-theoretic
        quantity.
        Its identification with physical energy ($E \propto \hbar\omega$)
        requires a derivation of the unit conversion from Planck units
        to SI, analogous to the treatment of $G$ in Paper~VIII.

  \item \textbf{Derivation of the D-$\psi$-G decomposition.}
        Equation~\eqref{eq:decomp} is conjectured, not derived.
        A proof that the boundary structure of any self-describing
        observer forces this three-way split would convert the
        conjecture into a theorem.

  \item \textbf{Continuum limit.}
        The Einstein field equations should emerge as the
        large-deviation stationarity conditions of $C_G$ over metric
        configurations in the continuum limit of the discrete model
        of Paper~VIII.
        This requires a treatment of the limit that is beyond the
        scope of the current toy models.

  \item \textbf{The topology question.}
        Why is the observer's boundary a 3-sphere?
        The spherical topology emerges naturally from the minimal
        boundary extension of the 1D encoding, but a proof that no
        other topology can dominate the measure has not been given.

  \item \textbf{Massive bosons and gauge structure.}
        Paper~VII identifies a photon-like massless boson.
        Massive mediators and the $\mathrm{SU}(3) \times \mathrm{SU}(2)
        \times \mathrm{U}(1)$ gauge structure of the Standard Model
        are not yet addressed.
\end{enumerate}

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have sketched a programme in which a single equation ---
the Solomonoff-weighted measure over observer-compatible histories ---
is conjectured to be sufficient to derive the structure of spacetime,
the origin of quantum mechanics, and the unification of General
Relativity with quantum field theory.

The evidence for this conjecture, assembled across this series of
papers, includes exact analytical results: bosons emerge as
compression residuals of fermionic systems with amplitude
$\sin(2\theta)/\sqrt{2}$; gravitational wave structure, polarisations,
and the Newtonian potential emerge from the identical decomposition
applied to metric configurations; Keplerian orbits are selected by
minimum description length; Bekenstein--Hawking entropy and the
inflationary aspect ratio are recovered from a single bit count
$n \approx 184$.

None of these results required the postulation of interaction terms,
coupling constants, force laws, or Hilbert spaces.
They follow from one operation --- the diagonal/off-diagonal split of a
density matrix --- applied to compressed configurations of different
physical degrees of freedom.

Whether this operation is itself a theorem of the organising
equation~\eqref{eq:master}, or whether additional structure is needed,
is the central unresolved question.
The present paper is a progress report on the way to an answer.

\printbibliography

\end{document}
