\chapter{Closure}
\label{chap:closure}

\subsection*{Philosophical Satisfaction}

We have arrived at a zero-parameter framework that can be intuitively grasped.
In principle, it allows us to compute probabilities for any structure — including the one that supports our own existence.
More profoundly, it reveals what the universe is ultimately made of: pure abstract information, from which both subjective experience and logical reasoning (mathematics) naturally emerge.

In mathematics, no object holds ontological privilege over another.
``True'' is not more fundamental than ``false'' — they form a single binary structure, like two sides of a coin.
An empty set is no less legitimate than a non-empty one.
Infinity carries no greater weight than the finite.
If any rule favored nothingness over existence, we would immediately need to explain the origin of \emph{that} rule.
Non-empty structures exist for the same reason empty ones do: nothing forbids them.

\subsection*{The Nature of Static Abstraction}

What kind of substrate is ``the abstract''? We already create and inhabit such universes ourselves.
Modern video games demonstrate the principle vividly: rich worlds populated by convincing avatars, governed by coherent physics.
In the limit, these simulated worlds could become informationally indistinguishable from our own.

The bullets and explosions in these games are not ``real'' in the physical sense — they are patterns of data.
The same holds for our universe.
Whether the underlying implementation uses electric pulses in silicon or wooden spheres rolling through mechanical gates, what matters is \emph{structure}.
Structure has no mass, no charge, and cannot be directly measured with physical instruments.
The block universe simply \emph{is}.

What we experience as causality, interaction, and the flow of time are internal processes of self-organization within conscious, geometric observers.
Time is the internal ordering of configurations.
The spatially bounded geometric self is simply the most stable, highly compressible pattern available to the observer.

\subsection*{Neural Networks as Analogy}

Contemporary neural networks offer a compelling demonstration.
These systems consist of massively parallel, weighted connections — almost entirely static data.
When they generate responses, there is no internal physics engine ticking forward step by step.
The output emerges as a projection from a highly compressed informational structure through high-dimensional constraint satisfaction.

The distinction between ``running'' and ``static'' becomes largely irrelevant.
Yet these systems can convincingly simulate motion, causality, and physical intuition.
This suggests that rich, law-like experience does not require active sequential execution.

\subsection*{Free Will Reconsidered}

This brings us to the question of free will.
At first glance, deterministic code seems to preclude genuine choice:

\begin{verbatim}
if (probability_of_observer_walk() < 0.3) {
    A
} else {
    B
}
\end{verbatim}

The code is deterministic and sequential. Yet the constant \texttt{0.3} — indeed the entire snippet — is merely our narrow, localized interpretation of an underlying bitstring.
The space of \emph{consistent interpretations} of those bits is astronomically vast.

Alice is not a train locked onto fixed tracks.
She navigates an enormous interpretation space, experiencing deliberation, anticipation, pain, and choice.
Pain is the felt steepness of a compressibility gradient.
Joy is the internal signature of harmonious, highly compressible configurations.
These qualia form a powerful heuristic navigation system.

Only at the highest layer does Alice access the Turing machine — mathematics and formal reasoning — as a precision tool.

\subsection*{Conclusion}

From our four initial axioms, we concluded that we are finite informational objects: bitstrings.
There are not many fundamental attributes to play with.

We know there is pain. We know the universe obeys crisp mathematical laws.
There seems to be no room for qualia to influence physical outcomes.
Yet bitstrings admit \emph{astronomically many} consistent interpretations.
Only a microscopically small subset of that interpretation space corresponds to mathematical theorems.
Consciousness and qualia have vast freedom to operate within the rest.

Why does a system governed by physical laws \emph{feel} anything at all? Because feeling is an effective compression heuristic.
Systems that experience pain are more likely to survive. The induced measure favors them.

\vspace{1.5em}
\begin{center}
    \sffamily
    \textbf{Unknown} \\
    $\downarrow$ \\
    \textbf{Information} \\
    $\downarrow$ \\
    \textbf{Qualia} \\
    $\downarrow$ \\
    \textbf{Mathematics} \\
    $\downarrow$ \\
    \textbf{Logical Reasoning} \\
    $\downarrow$ \\
    \textbf{Alice}
\end{center}
\vspace{1.5em}

And more specifically:

\begin{center}
    \sffamily
    \textbf{Alice exists as finite information} \\
    $\downarrow$ \\
    \textbf{Alice can be simulated} \\
    $\downarrow$ \\
    \textbf{The simulation can be transformed into static data} \\
    $\downarrow$ \\
    \textbf{Alice is preserved under that transformation} \\
    $\downarrow$ \\
    \textbf{Execution is therefore irrelevant} \\
    $\downarrow$ \\
    \textbf{Time is internal to Alice} \\
    $\downarrow$ \\
    \textbf{Reality is structure, not active code} \\
    $\downarrow$ \\
    \textbf{Structure admits astronomically many interpretations} \\
    $\downarrow$ \\
    \textbf{Qualia operates in this space as a heuristic compression system} \\
    $\downarrow$ \\
    \textbf{Mathematics is the most compressible structure available to Alice} \\
    $\downarrow$ \\
    \textbf{The most compressible descriptions dominate the measure} \\
    $\downarrow$ \\
    \textbf{Alice therefore finds herself in a law-like, fine-tuned universe} \\
    $\downarrow$ \\
    \textbf{Because predictable information compresses best} \\
    $\downarrow$ \\
    \textbf{Predictability is what Alice calls the laws of physics.}
\end{center}
