\chapter{The Four Assumptions}

The theory is built from a single starting point: there is an observer,
and that observer notices a set of modest, empirically grounded facts about
the universe. The theory --- space, time, particles, forces --- is
what follows.

We begin with four rules. These are concrete, empirically grounded claims
about the world we actually inhabit. 

\section*{The Four Baseline Rules}

\begin{enumerate}
    \item \textbf{The Biological Blueprint.} Human DNA is a physical
    instruction set. Given the right environment --- nutrients, warmth,
    oxygen --- it reliably constructs a conscious organism capable of
    experiencing physical states like pain. 
    This is an empirical claim supported by every biological
    observation ever made.

    \item \textbf{Physical Closure.} That organism and its environment
    are made entirely of ordinary matter governed by ordinary physical
    laws. No supernatural ingredients. No unmeasurable forces slipping
    into the machinery. Everything that happens inside a human body is,
    in principle, physics.

    \item \textbf{Local Computability.} The physical processes that DNA
    relies on are computable. We do not need the entire universe to be
    simulatable, only need the specific, localised physics of biology
    to finite precision, on a sufficiently powerful
    machine. This is a strictly weaker claim than the full
    Church--Turing thesis.

    \item \textbf{Causal Efficacy.} Internal experiences matter. A
    human who feels pain behaves differently than one who does not.
    Experience is not a useless byproduct of neural activity. It
    changes physical trajectories. It is causally real.
\end{enumerate}

These four rules are the entire foundation. No additional assumptions
are made. 

\section{The Ultimate Optimisation}

Let us run a thought experiment.

We build a computer and use it to simulate a human --- call her Alice
--- based strictly on her digitised DNA profile and the laws of physics.
The computer simulates Alice step by step. As the program runs, virtual
Alice exists, experiencing her virtual universe in the same way
real-world Alice experiences ours.

Now we optimise the software. We replace common calculations with
precomputed lookup tables, so the computer no longer spends CPU cycles
working out how a protein folds or how a neuron fires under specific
conditions. It simply retrieves the answer from memory.

What happens?

\begin{itemize}
    \item \textbf{To an outside observer:} the simulation runs faster.
    The number of CPU cycles drops.
    \item \textbf{To Alice:} nothing changes. Her internal sequence of
    thoughts, perceptions, and experiences is identical.
\end{itemize}

We continue optimising. Every calculation, one by one, is replaced with
precomputed data. At each step, the output is the same. The only thing
changing is the software producing it.

Taken to its extreme limit, the software disappears entirely. What
remains is a static dataset --- a massive set of data on a hard drive ---
containing the complete record of every state Alice will ever pass
through. No processors are running. No electricity is flowing. No clock
is ticking. The external runtime is exactly zero.

Does Alice's experience still exist?

If the answer is no, then somewhere in that optimisation process
consciousness and pain switched off. There is a specific ratio of code to data
below which experience vanishes. But that threshold is a new physical
constant --- one that does not exist in any Turing Machine or computer. 
It is magick arguing that $a+b=c$ may or may not hold depending on
the code / data ration of the software computing it. 


\begin{quote}
\textbf{Conclusion 1: Time is internal.} The flow of time does not
require an active engine to push it forward. Time is an internal
ordering relation within a structure of information. What Alice
experiences as the passage of time is simply the ordered sequence of
states encoded in the dataset --- traversed from the inside, not driven
from the outside.
\end{quote}

\begin{quote}
\textbf{Conclusion 2: The observer is an equivalence class.} The
optimisation argument produces not one execution trace but a whole
family of them --- the original simulation, the partially optimised
versions, the fully precomputed dataset, and real-world Alice herself.
Each is a different description of the same observer. Alice is not any
particular trace. She is the invariant that all of them share: the
equivalence class $[A]$ of every execution trace encoding the same
internal experience sequence. Running the simulation again does not
create a new Alice. It produces another description of the same one.
Computers do not manufacture observers. They reveal information of a
structure that is already there.
\end{quote}

\section{The Shredder Experiment: Why Substrate Does Not Matter}

The optimisation argument established that the \emph{process} generating
Alice does not matter. A second thought experiment establishes that the
\emph{arrangement} of her data does not matter either.

Imagine our computer running Alice's simulation while simultaneously
running a billion other simulations. To handle the load, it interleaves
execution: one CPU cycle of Alice's code, then millions of steps from other
programs, then back to Alice. From the outside, the raw binary stream
containing Alice's data is completely fragmented. Her information is
separated by vast oceans of unrelated noise. It looks, to any external
observer, like randomness.

Does Alice notice? Does her experience fragment or degrade?

It cannot. Her internal experience depends entirely on the relational
order of her information --- which state follows which --- not on how
densely her bits are packed in a physical medium. You could print her
execution trace across a billion pieces of paper, scatter them across
the galaxy, and the internal coherence of her experience would be
unaffected. Based on massive empirical evidence we know multi-threading works.

The wooden computer makes this vivid. Build a computer from wooden
spheres and mechanical ports. Step it through the same DNA simulation.
The result is identical to the electronic version. Just as $2 + 2 = 4$
whether you are counting apples or bananas, the mathematics of an
observer is invariant across physical substrates.

\begin{quote}
\textbf{Conclusion 3: Substrate independence.} The medium is
irrelevant. Neurons, silicon, magnetic domains, interleaved bits
embedded in noise --- these are all different physical arrangements of
the same information. A conscious observer is not a property of any
particular substrate. It is a property of a particular information
structure, regardless of what carries it.
\end{quote}

\section{What the Observer Actually Is}

These three conclusions, taken together, force a precise answer to a
question that is usually left vague: what is an observer?

We don't pretend to know the answer. We know, however, what an observer is not.
She is not a body, a brain, a running process,  a
single specific arrangement of bits. Each of those is a \emph{description} --- one
representative from a much larger equivalence class. The observer is
the invariant that all valid descriptions share.

This matters because it dissolves a confusion that derails most
discussions of simulation and consciousness. People ask: if I copy
Alice's data and run it on three different computers, have I created
three Alices?

No. And physics already has a precise framework for understanding why.

When a physicist samples a quantum wavefunction $\psi$ multiple times,
each measurement produces a different outcome --- a different bitstring,
so to speak. But those multiple samples do not create multiple distinct
particles. They are localised glimpses of the same underlying invariant
$\psi$. The Born rule, $|\psi|^2$, gives the probability of each
glimpse. The particle is the equivalence class; the measurements are
its representatives.

Alice is the same structure. Running her simulation three times produces
three descriptions --- three samplings --- of the same observer. No new
information is created. No new Alice comes into existence. The
equivalence class was already there.

This analogy is not decorative. It will become load-bearing in later
chapters, where we derive the Born rule itself from the same compression
principle that produces observers. The structure here and the structure
there are the same structure.

\section{Causality between the Computer and the Simulated Alice}

It is often assumed that a computer must actively execute code to bring a simulated Alice into existence.
However, this simulated Alice is not created within our physical universe.
We cannot measure a spatial distance between the computer and Alice, nor can we measure any temporal delay between a bit-flip in the computer and its corresponding effect on her virtual state.
Therefore, one cannot argue that one created the other.
The computer executing the simulation and the simulated agent are not causally linked;
they are two distinct representations of the same underlying information.

\section{What This Means for Physics}

If a computer running a simulation of a physical system and the
physical system itself are two descriptions of the same information,
then every property of the physical system must appear in the execution
trace of its simulation. The geometric description and the
computational description are representationally equivalent.

This equivalence is the foundation on which everything that follows is
built. We may not fully understand the universe from the inside. But
we understand computers. And if the two are the same thing viewed
differently, then studying the computational description gives us
direct access to the physics.

The universe that emerges from this foundation has three properties
that we did not assume but were forced to accept:

\begin{enumerate}
    \item \textbf{Timeless.} At the level of the whole, the universe
    is a static structure. Time is an internal relation, not an
    external driver.

    \item \textbf{Finite.} The simulation computer is a discrete,
    finite system. The universe it describes is therefore characterised
    by a finite bit count $n$. This is not an assumption. It follows
    from Rule 3.

    \item \textbf{Substrate-independent.} The boundary between physics,
    geometry, and pure information dissolves. These are not different
    things. They are different descriptions of the same thing.
\end{enumerate}

